\chapter{\tlhe{b}}

\tlhw{bach}
\tlhd{n}{Shot.}
\tlhd{v}{Shoot.}
\tlhrf{The verb used for shoot when referring to disruptors is \tlhe{bach}. Technically speaking, one shoots the energy beam from the disruptor. The general word for any energy beam (``ray'')is \tlhe{tIH}, so a disruptor's beam is \tlhe{nISwI\' tIH}. Thus, the correct formation is \tlhe{nISwI\' tIH bach} (``shoot the disruptor beam''). As a practical matter, however, the \tlhe{tIH} is often left out, and \tlhe{nISwI\' bach} is the common way to say ``shoot a disruptor''. Similarly, \tlhe{pu\' bach} is ``shoot a phaser.''}{KGT}[56]
\tlhc
\tlhca{bach Do\', qaH}{A lucky shot sir...}[ST3]
\tlhca{logh veQDaq bachchugh, yoH \'e\' toblaHbe\' SuvwI\'}{Shooting space garbage is no test of a warrior's mettle.}[ST5 notes]
\tlhcf{baH}
\tlhf{TKD}[TKW][KGT]

\tlhw{bachHa\'}
\tlhd{v, slang}{Err.}[Make a mistake.]
\tlhco{Note that this is obviously not a noun. I'm convinced that he made this mistake on purpose, given the definition. It sets the tone for the degree of perfection we are to expect from this work. He knows there will be errors, so why wait. Put one in the first entry.}{charghwI\'}
\tlhc
\tlhca{jIbachHa\'pu\'}{I have made a mistake (slang: ``I have mis-shot'')}[KGT]
\tlhca{bIQ ngaS HIvje\'}{The cup contains water - ``This idiomatic expression means ``be quick mistaken, be totally wrong''.''}[KGT 20]
\tlhpun{botch}
\tlhcf{Qach}[muj][lughbe\'][qarbe\']
\tlhf{KGT}

\tlhw{bagh}
\tlhd{v}{(to) Tie.}
\tlhcf{muv}[rar][tay\']
\tlhf{TKDa}

\tlhw{baghneQ}
\tlhd{n}{Spoon.}
\tlhrf{Eating is done with hands only. There is no Klingon fork or spoon. If the cook has prepared the food properly, there should be no need to use a knife either, though, from time to time, one is quite useful.}{KGT}[99]
\tlhco{The Klingon word for ``fork'' is \tlhe{puq chonnaQ}. As is well known, Klingons prefer to get food into their mouths without the aid of implements (except for such things as the bowl containing soup or the goblet containing bloodwine). Nevertheless, they have become acquainted with the eating habits of other cultures and have become aware of such things as forks. On occasion, they'll even use the implements, most commonly when partaking of a non-Klingon meal (whether on a Klingon planet or elsewhere) but sometimes when eating Klingon food, as if to add an exotic touch to the meal experience. (Not all Klingons are skilled in using forks, however, and some simply refuse to deal with them. Those who do not use them seem to be not at all troubled by eating ``foreign'' food using Klingon means -- that is, hands.) ... The Klingon word for ``spoon'' is \tlhe{baghneQ}. Even though spoons were never typically used when eating, the word appears to have been in the language for a long time, suggesting that it may once have meant something else. One theory is that it comes from \tlhe{nagh beQ} ``flat stone, flat rock'' and that the initial sounds of the two words, n and b, were, for some reason, transposed. This is, however, just speculation.}{st.k 5/05/98}
\tlhpun{a spoonerism}
\tlhcf{bo\'Dagh}
\tlhf{st.k}

\tlhw{baH}
\tlhd{v}{Fire (torpedo, rocket, missle.)}
\tlhrf{Ships are also equipped with an explosive weapon called a {peng}, usually translated as ``torpedo'', though ``missile'' might work just as well... The verb used for ``launch'' or ``fire'' a weapon of this type is \tlhe{baH}, and there is even a special word, \tlhe{ghuS}, meaning prepare to launch.}{KGT}[56]
\tlhrf{The verb used for shoot when referring to disruptors is \tlhe{bach} ... the correct formation is \tlhe{nISwI\' tIH bach} (shoot the disruptor beam) ... however, the \tlhe{tIH} is often left out, and \tlhe{nISwI\' bach} is the common way to say ``shoot a disruptor''.}{KGT}[56]
\tlhco{This was one of the eleven original Klingon words created by James Doohan and spoken by Mark Lenard (through a set of prosthetic teeth) in the opening scene of STMP in which Marc Okrand used as the basis for \tlhe{tlhIngan Hol}. \tlhw{baH} for ``Fire!'' was clipped and, in Lenard's character's dialect, pronced ``\tlhe{maH}'' (as mentioned in TKD). ``(Having said that, the command meaning `Fire [a torpedo]!' - which I transcribed as ``\tlhw{baH}'' but which also sounds kind of like``maH'' - must have always had that meaning, since it's there a couple of times. [The H is pronced like the final ch in the name of the composer Bach.])''}{M.Ok. BBS 2/98}
\tlhco{Doohan, who knows some Russian MUST have gotten this from Russian. In Russian BAX pronced /baH/ means BANG! as in a gun shot. I can\'t help but believe that this was deliberate on the part of Doohan, despite protestations by the Philistines.}{(G. Proechel) STMP}
\tlhc
\tlhca{baH}{Fire!}[CK]
\tlhca{baHchu\'}{He fire (the torpedo) perfectly.}[TKD]
\tlhca{cha yIbaH qara\'DI\' \emph{or} qara\'DI\' cha yIbaH}{Fire the torpedoes at my command.}[TKD]
\tlhca{yIbaH!}{Fire (the torpedoes)!}[TKD]
\tlhca{yIbaH!}{Fire (the torpedo)!}[KGT]
\tlhca{Qo\'noS wa\'Daq baHta\' [`entepray']}{Enterprise fired on Kronos One...}[ST6]
\tlhca{nISwI\': cha\' chang\'engmey (telDaq lujomlu\', nItebHa\' lubaHlu\')}{Disruptor - 2 Pairs (Wing Mounted, Fire Linked)}[KBoP]
\tlhcf{bach}[baHjan][baHwI\'][chuH][laQ][vo\']
\tlhf{TKD}[KGT][KBoP]

\tlhw{baHjan}
\tlhd{n}{Launcher.}
\tlhc
\tlhca{peng baHjan}{Torpedo launcher.}[BoP]
\tlhca{\'otlh peng baHjan}{Photon torpedo launcher.}[BoP]
\tlhf{BoP}

\tlhw{baHwI\'}
\tlhd{n}{Gunner.}
\tlhrf{...which consists of the v \tlhtr{baH}{fire (a torpedo)} plus \tlhtr{-wI\'}{one who does}. Thus, \tlhe{baHwI\'} is literally ``one who fires [a torpedo]''.}{TKD}[20]
\tlhc
\tlhca{baHwI\'pu\'vam}{These gunners}[TKD]
\tlhca{baHwI\', DoS yUbuS. QuQ neH. yaj\'a\'}{Gunner: Target engine only. Understood?}[ST3]
\tlhca{po\'neS baHwI\'pu\'lI\'}{Gunner, stand ready.}[TDP]
\tlhca{baHwI\', \'ejyo\'waw\' DoS}{Target the starbase, gunner.}[TDP]
\tlhcf{matHa\'}
\tlhf{TKD}

\tlhw{baj}
\tlhd{v}{(Actively) work for.}[Garner.][Earn.]
\tlhrf{Though any perceived attack on one\'s honor may prompt one to issue a challenge, in one traditional form of duel, the goal is specifically for a man to ``win the favor of a women'' ({vuv be\' \'e\' baj} [literally, ``earn that a woman respect him'']) by competing with another man.}{KGT}[69][3]
\tlhc
\tlhca{yInlu\'taH \'e\' bajnISlu\'}{Survival must be earned.}[TKW]
\tlhca{yay chavlu\' \'e\' bajnISlu\'}{Victory must be earned.}[TKW]
\tlhf{TKW}[KGT]

\tlhw{bal}
\tlhd{n}{Jug.}[Jar.][Bottle.]
\tlhrf{A container from which drinks are consumed is a \tlhe{HIvje\'}. This word is used for any sort of drinking vessel: glass, cup, mug, stein, goblet, tumbler, and so on. [...] A \tlhe{bal} is any bottle or jug used to store liquid. Usually one will pour (\tlhe{qang}) the drink directly from the \tlhe{bal} into the \tlhe{HIvje\'}, but it is not considered uncivil to drink directly from the \tlhe{bal}. Ale [\tlhe{wornagh}] is often stored in a large vat (\tlhe{qegh}) rather than in a \tlhe{bal} Some drinks are served in special containers, such as the \tlhe{runpI\'} (teapot) used in the making of \tlhe{Dargh} (tea), and are then poured into the appropriate \tlhe{HIvje\'}.}{KGT}[98]
\tlhrf{a cook's ever-present supply of \tlhe{chanDoq} (marinade) is kept in a large \tlhe{bal}}{KGT}[97]
\tlhrf{A \tlhe{bIQ bal} (water jug) is ``traditionally kept filled and set at the front door of every warrior home. It is said, when a warrior returns from battle, he thirsts, and should not want for satisfaction.''}{KCD}
\tlhf{KGT}

\tlhw{banan naH}
\tlhd{n}{Banana.}
\tlhf{Eurotalk}

\tlhw{bang}
\tlhd{n}{Love, one who is loved.}
\tlhco{If you think ``Loved one'' when using this word, you can't go wrong.}{Voragh}
\tlhf{TKD}

\tlhw{bang bom}
\tlhd{n}{Love song.}
\tlhf{KGT}

\tlhw{bang pong}
\tlhd{n}{Pet name.}[Term of engearment.]
\tlhf{KGT}

\tlhw{baq}
\tlhd{v}{Terminate, discontinue.}
\tlhf{KGT}

\tlhw{baqghol}
\tlhd{n}{A type of liquor, usually fatal to non-Klingons.}
\tlhf{KGT}

\tlhw{baQ}
\tlhd{v}{Toss \tlhe{betleH} from hand to hand.}[Be fresh, just picked. (Said of a fruit or vegetable.)]
\tlhf{KGT}

\tlhw{baQa\'}
\tlhd{excl}{(invective) ~ ``stupid, idiot.''}
\tlhf{TKD}

\tlhw{bargh}
\tlhd{n}{Flat-bottomed pot for food preparation.}
\tlhcf{vutmeH \'un}[vut\'un]
\tlhf{KGT}

\tlhw{barat}
\tlhd{n}{India.}
\tlhf{Eurotalk}

\tlhw{barot}
\tlhd{name}{Barot.}
\tlhf{KGT 182}

\tlhw{baS}
\tlhd{n}{Metal.}
\tlhf{TKD}

\tlhw{baS \'In}
\tlhd{n}{Bell (percussion instrument made of metal).}
\tlhf{KGT 75}

\tlhw{batlh}
\tlhd{adj}{To be honored, with honor}[In an honored way.][Honorably.][Well.]
\tlhd{n}{Honor}[General, philosophical concept, associated with integrity.]
\tlhc
\tlhca{batlh potlh law\' yIn potlh puS}{Honor is more important than life.}[TKW 53]
\tlhca{batlh muQaHpu\'}{``He/she has helped me in an honored way'' or ``He/she has helped me with honor''.)}[KGT 39]
\tlhca{batlh choQaHpu\'}{``You have helped me in an honored way'' or ``You have helped me with honor''.}[KGT 39]
\tlhf{TKD}[TKW][KGT][HolQeD v12n3p9]

\tlhw{batlhHa\'}
\tlhd{adv}{Dishonorably.}
\tlhf{TKW}[KGT 55, p139]

\tlhw{bav}
\tlhd{v}{Orbit.}
\tlhf{TKD}

\tlhw{bay}
\tlhd{n}{The consonant ``b''.}
\tlhf{qep\'a\' wa\'maH javDIch 2009}

\tlhw{ba\'}
\tlhd{v}{Sit.}
\tlhf{TKD}

\tlhw{beb}
\tlhd{n}{Roof.}
\tlhf{st.k}

\tlhw{beb mutlhwI\'}
\tlhd{n}{roofer}
\tlhf{(email to Lieven, 9 january 2012)}

\tlhw{bech}
\tlhd{v}{Suffer.}
\tlhf{TKD}

\tlhw{begh}
\tlhd{n}{Deflectors.}
\tlhf{TKD}

\tlhw{beH}
\tlhd{n}{Rifle.}
\tlhf{TKD}

\tlhw{bej}
\tlhd{v}{Watch.}[\textit{(slang)} Be sure, be definite, be positive, be certain.]
\tlhf{TKD}

\tlhw{bel}
\tlhd{n}{Pleasure.}
\tlhd{v}{Be pleased.}
\tlhf{TKD}

\tlhw{belHa\'}
\tlhd{v}{Be displeased.}
\tlhf{TKD}

\tlhw{bem}
\tlhd{n}{Sole of the foot.}
\tlhf{KGT}

\tlhw{ben}
\tlhd{n}{Years ago.}
\tlhf{TKD}

\tlhw{bep}
\tlhd{n}{Agony.}
\tlhd{v}{Complain, object, gripe.}
\tlhf{TKD}

\tlhw{beq}
\tlhd{n}{Crew, crewman.}
\tlhf{TKD}

\tlhw{beqquj}
\tlhd{n}{\emph{bekpuj}, a type of mineral (orange in color).}
\tlhf{KGT}

\tlhw{bergh}
\tlhd{v}{Be irritable.}
\tlhf{TKD}

\tlhw{bertlham}
\tlhd{n}{End (of an opera, play, story, speech)}
\tlhco{This refers to the final section of such a performance (last act, scene, sentence, etc.)}{HolQeQ}
\tlhf{HolQeQ v12n2p8}

\tlhw{betleH}
\tlhd{n}{Type of hand weapon.}
\tlhf{TKD}

\tlhw{betleH bey\'}
\tlhd{n}{\emph{Bat\'telh} display.}
\tlhf{KCD}

\tlhw{betleH \'obe\'}
\tlhd{n}{Order of the \emph{Bat\'telh}.}
\tlhf{KGT}

\tlhw{bey}
\tlhd{n}{Howl, wail.}
\tlhf{KGT}

\tlhw{bey\'}
\tlhd{n}{Ceremonial display.}
\tlhf{KGT}

\tlhw{be\'}
\tlhd{n}{Female, woman.}
\tlhf{TKD}

\tlhw{be\'etor}
\tlhd{name}{B’Etor - one of the DuraS sisters.}
\tlhf{KGT 197}[HolQeD v5n3p15][S26.]

\tlhw{be\'Hom}
\tlhd{n}{Girl.}
\tlhf{TKD}

\tlhw{be\'joy\'}
\tlhd{n}{Ritualized torture by women.}
\tlhf{Sarek}

\tlhw{be\'nal}
\tlhd{n}{Wife.}
\tlhf{TKD}

\tlhw{be\'nI\'}
\tlhd{n}{Sister.}
\tlhf{TKD}

\tlhw{bID}
\tlhd{n}{Half.}
\tlhf{TKD}

\tlhw{bIghHa\'}
\tlhd{n}{Prison.}[Jail.]
\tlhf{TKD}

\tlhw{bIH}
\tlhd{pro}{They, them (incapable of language.)}
\tlhf{TKD}

\tlhw{bIj}
\tlhd{n}{Punishment.}
\tlhd{v}{Punish.}
\tlhf{TKD}

\tlhw{bIm}
\tlhd{n}{Second tone of nonatonic scale.}
\tlhf{KGT}

\tlhw{bIng}
\tlhd{n}{Area below, area under.}
\tlhf{TKD}

\tlhw{bIp}
\tlhd{num.}{Hundred thousand.}
\tlhf{TKD}

\tlhw{bIQ}
\tlhd{n}{Water.}
\tlhco{Water is used as the symbol of weakness and disdain.}{TKW}
\tlhf{TKD}[TKW]

\tlhw{bIQ bal}
\tlhd{n}{Water jug.}
\tlhf{KGT}

\tlhw{bIQDep}
\tlhd{n}{Fish.}
\tlhf{KGT}

\tlhw{bIQSIp}
\tlhd{n}{Hydrogen.}
\tlhf{BoP Poster}

\tlhw{bIQSIp \'ugh}
\tlhd{n}{Deuterium isotope.}
\tlhf{BOP}

\tlhw{bIQtIq}
\tlhd{n}{River.}
\tlhc
\tlhca{Doq bIQtIq bIQ}{The river water is red.}[KGT 123]
\tlhca{Doq bIQtIq}{The river is red.}[KGT 123.]
\tlhf{TKD}[KGT]

\tlhw{bIQ\'a\'}
\tlhd{n}{Ocean.}
\tlhf{TKD}

\tlhw{bIr}
\tlhd{v}{Be cold.}
\tlhf{TKD}

\tlhw{bIraqlul}
\tlhd{n}{\emph{brak’lul}, redundancy in body parts.}
\tlhf{KGT}

\tlhw{bIreQtagh}
\tlhd{n}{\emph{Bregit} lung.}
\tlhco{Bregit is a Klingon animal that resembles a Terran dog. Bregit lung is a warrior's food often prepared on Klingon warships.}{KCD}
\tlhf{TKDa}[KCD]

\tlhw{bIS\'ub}
\tlhd{n}{Bottom (interior).}
\tlhf{HolQeD v8n3p2}

\tlhw{bIt}
\tlhd{v}{Be nervous, uneasy.}
\tlhf{TKD}[KGT]

\tlhw{bIv}
\tlhd{v}{Break (rules).}
\tlhf{TKD}

\tlhw{bI\'rel tlharghDuj}
\tlhd{n}{B\'rel-class Scout.}
\tlhf{BoP Poster}

\tlhw{bI\'reqtal}
\tlhd{n}{\emph{Brek’tal} ritual.}
\tlhf{KGT}

\tlhw{bI\'reS}
\tlhd{n}{Beginning (of an opera, play, story, speech.)}
\tlhrf{This refers to the opening section of such a performance (last act, scene, sentence, etc.)}{HolQeD}[v12n2p8]
\tlhf{HolQeD}

\tlhw{bI\'}
\tlhd{v}{Sweep (away).}
\tlhf{KGT}

\tlhw{bo}
\tlhd{n}{Feather.}
\tlhf{HolQeD v10n4p5}

\tlhw{bobcho\'}
\tlhd{n}{Module.}
\tlhf{TKD}

\tlhw{boch}
\tlhd{v}{Shine, be shiny.}
\tlhf{TKD}

\tlhw{bochmoHwI\'}
\tlhd{n}{Sychophant, flatterer. (slang)}
\tlhf{KGT}

\tlhw{boD}
\tlhd{n}{Forehead. (regional)}
\tlhf{KGT}

\tlhw{bogh}
\tlhd{v}{Be born.}
\tlhf{TKD}

\tlhw{boH}
\tlhd{v}{Be impatient.}
\tlhf{TKD}

\tlhw{boj}
\tlhd{v}{Nag.}
\tlhf{TKD}

\tlhw{bol}
\tlhd{v}{Drool.}
\tlhf{KGT}

\tlhw{bolwI\'}
\tlhd{n}{Traitor. (slang)}
\tlhf{KGT}

\tlhw{bom}
\tlhd{n}{Song, chant.}
\tlhd{v}{Sing, chant.}
\tlhf{TKW}[KGT]

\tlhw{bomwI\'}
\tlhd{n}{Singer.}
\tlhf{KGT}

\tlhw{bom mu\'}
\tlhd{n}{Lyric, lyrics.}
\tlhf{KGT}

\tlhw{bong}
\tlhd{adv}{Accidentally, by accident.}
\tlhf{TKD}

\tlhw{bop}
\tlhd{v}{Be (all) about (something).}
\tlhc
\tlhca{\'e\' bop}{That's what it's all about}[qep\'a\' wa\'maH cha\'DIch 2005]
\tlhf{qep\'a\'}[HolQeD v14n2-3p21]

\tlhw{boq}
\tlhd{n}{Alliance.}[Bloc, coalition.][Fusion.]
\tlhd{v}{Ally with, form an alliance with.}[Used in math for addition.]
\tlhcf{chen}
\tlhf{TKD}[KGT][BoP Poster][HolQeD v9n3p10]

\tlhw{boqHa\'}
\tlhd{v}{Used in math for subtraction. }
\tlhcf{chen}
\tlhf{HolQeD v9n3p10}

\tlhw{boqHa\'\'egh}
\tlhd{v}{Used in math for division: \tlhe{X-logh boqHa\'\'egh Y}.}
\tlhcf{chen}
\tlhf{HolQeD v9n3p10}

\tlhw{boqrat}
\tlhd{n}{\emph{bokrat}, a type of animal.}
\tlhf{KGT}

\tlhw{boq\'egh}
\tlhd{v}{Used in math for multiplication: \tlhe{X-logh boq\'egh Y}.}
\tlhcf{chen}
\tlhf{HolQeD v9n3p10}

\tlhw{boq ru}
\tlhd{n}{Coalition.}
\tlhf{KGT}

\tlhw{boQ}
\tlhd{n}{Aid, assistance.}
\tlhd{v}{Assist.}
\tlhf{TKD}

\tlhw{boQDu\'}
\tlhd{n}{Aide-de-camp.}
\tlhf{TKD}

\tlhw{bor}
\tlhd{v}{Gurgle.}
\tlhco{This specifically refers to the sound that a stomach makes.}{HolQeD}[v12n4p8]
\tlhf{HolQeD}

\tlhw{borghel}
\tlhd{n}{A very small bird whose eggs are considered quite tasty.}
\tlhf{HolQeD v10n4p4}

\tlhw{bortaS}
\tlhd{n}{Revenge.}
\tlhf{TKD}

\tlhw{bortaS DIb}
\tlhd{n}{Right of Vengeance, a legal right.}
\tlhf{TKW}[KGT]

\tlhw{boS}
\tlhd{v}{Collect.}
\tlhf{TKD}

\tlhw{bot}
\tlhd{v}{Prevent, block, prohibit.}
\tlhf{TKD}[TKW]

\tlhw{botjan}
\tlhd{n}{Shields (force field on ship).}
\tlhf{KGT}

\tlhw{botlh}
\tlhd{n}{Center, middle.}
\tlhf{TKD}

\tlhw{bov}
\tlhd{n}{Era.}
\tlhf{TKD}

\tlhw{bo\'Dagh}
\tlhd{n}{Scoop.}[Scooping implement (to serve food).]
\tlhf{KGT}

\tlhw{bo\'Degh}
\tlhd{n}{Bird.}[The most general word for a bird-like creature.]
\tlhf{HolQeD v10n4p4}

\tlhw{bo\'DIj}
\tlhd{n}{Court.}
\tlhf{TKD}

\tlhw{buD}
\tlhd{v}{Be lazy.}
\tlhf{TKD}

\tlhw{bup}
\tlhd{v}{Quit.}
\tlhf{TKD}

\tlhw{buq}
\tlhd{n}{bag, sack, pouch, pocket}
\tlhco{When clarity is needed, one could say, for example, \tlhtr{yopwaH buq}{pants pouch} or \tlhtr{wep buq}{coat pouch} for "pants pocket" or "coat pocket," but when the context is clear, \tlhe{buq} alone would suffice for "pocket."}
\tlhf{(qepHom 2010, email to Lieven, 15 November 2010)}

\tlhw{buQ}
\tlhd{v}{Threaten.}
\tlhf{TKD}

\tlhw{bur}
\tlhd{n}{Hiccup.}
\tlhf{HolQeD v12n4p8}

\tlhw{burgh}
\tlhd{n}{Stomach.}
\tlhf{TKD}

\tlhw{burgh quD}
\tlhd{n}{Naturally produced \emph{qud}.}
\tlhf{TKD}

\tlhw{buS}
\tlhd{v}{Concentrate on, focus on, think only about.}[Pay heed to, pay attention to.]
\tlhf{TKD}[TKW]

\tlhw{buSHa\'}
\tlhd{v}{Ignore.}
\tlhf{TKW}[KGT]

\tlhw{butlh}
\tlhd{n}{Dirt under fingernails.}
\tlhco{Symbolic equivalent of gall, an admirable trait to have.}{TKW}
\tlhf{TKD}[TKW p142]

\tlhw{buv}
\tlhd{n}{Classification.}
\tlhd{v}{Classify.}
\tlhf{TKD}

\tlhw{buy\'}
\tlhd{v}{Be full, be filled up.}
\tlhf{KGT}

\tlhw{buy\' ngop}
\tlhd{excl}{Good news.}[It’s good to hear that. (slang)]
\tlhf{KGT}

\tlhw{bu\'}
\tlhd{n}{Sergeant.}
\tlhf{TKD}
