\chapter{\tlhe{ch}}

\tlhw{cha}
\tlhd{n}{Torpedoes.}
\tlhf{TKD}

\tlhw{chab}
\tlhd{n}{Pie, tart, dumpling.}[Invention, innovation. (slang)]
\tlhf{TKD}[KGT]

\tlhw{chach}
\tlhd{n}{Emergency.}
\tlhf{TKD}

\tlhw{chaDvay\'}
\tlhd{n}{Hertz (frequency).}
\tlhf{KGT}[veS QonoS][HolQeD v1n3p9]

\tlhw{chagh}
\tlhd{v}{Drop.}
\tlhf{TKD}

\tlhw{chaH}
\tlhd{pro}{They, them (capable of using language.)}
\tlhf{TKD}

\tlhw{chaj}
\tlhd{n}{Close female friend of a female.}
\tlhf{KGT}

\tlhw{chal}
\tlhd{n}{Sky.}
\tlhf{TKD}

\tlhw{cham}
\tlhd{n}{Technology.}
\tlhf{KGT}

\tlhw{chamwI\'}
\tlhd{n}{Technician.}
\tlhf{TKD}

\tlhw{chan}
\tlhd{n}{Eastward, area toward the east.}
\tlhrf{90 degrees on Terran 360 degree compass with north as 0 and degrees count clockwise.}{HolQeD}
\tlhf{HolQeD v8n4p6}

\tlhw{chan}
\tlhd{n}{Marinade.}
\tlhf{KGT}

\tlhw{chang\'eng}
\tlhd{n}{Pair.}
\tlhf{BoP}

\tlhw{chap}
\tlhd{n}{Back (of hand.)}
\tlhf{TKD}

\tlhw{chaq}
\tlhd{adv}{Perhaps, maybe.}
\tlhf{TKD}[TKW]

\tlhw{chaQ}
\tlhd{v}{Thrust upward with end of bat'leth.}
\tlhf{KGT}

\tlhw{char}
\tlhd{v}{Be slimy.}
\tlhf{KGT}

\tlhw{chargh}
\tlhd{v}{Conquer.}
\tlhf{TKD}[TKW]

\tlhw{charghwI\'}
\tlhd{n}{1. Conqueror. 2. Victor.}
\tlhf{TKW}[KGT]

\tlhw{chaS}
\tlhd{n}{Top of walking cane or ceremonial cane.}
\tlhf{HolQeD v12n2p7}

\tlhw{chatlh}
\tlhd{n}{1. Soup. 2. Nonsense, balderdash.}
\tlhf{KGT}

\tlhw{chav}
\tlhdd{n}{Achievement.}
\tlhd{v}{Achieve.}
\tlhf{TKD}[TKW]

\tlhw{chaw\'}
\tlhdd{n}{Permit.}
\tlhd{v}{Allow, permit.}
\tlhco{This entry was missing from the Klingon-English side of TKD, though it is used as a noun in PK.}{Voragh}
\tlhc \tlhca{cha\'puj vIngevmeH chaw\' HInobneS}{Give me a permit to sell dilithium, your honor.}{PK}
\tlhf{TKD, PK, TKW p37}

\tlhw{chay}
\tlhd{n}{The consonant "ch".}
\tlhf{qep\'a\' wa\'maH javDIch 2009}

\tlhw{chay\'}
\tlhd{ques}{How?}
\tlhf{TKD}

\tlhw{cha\'}
\tlhdd{num}{Two.}
\tlhd{v}{Show, display (picture.)}
\tlhf{TKD}

\tlhw{cha\'bIp}
\tlhd{n}{A bird noted for its speed.}
\tlhf{HolQeD v10n4p4}

\tlhw{cha\'DIch n, num. –n. 1. Second. 2. Aide. –num. 3. Second.}
\tlhf{TKD, TKW p21}

\tlhw{cha\'Do\'}
\tlhd{n}{A species of Klingon bird about which little is known.}
\tlhf{HolQeD v10n4p4}

\tlhw{cha\'\'etlh pe\'wI\'}
\tlhd{n}{Scissors.}
\tlhf{Eurotalk}

\tlhw{cha\'Hu\'}
\tlhd{n}{Day before yesterday.}
\tlhf{TKD}

\tlhw{cha\'leS}
\tlhd{n}{Day after tomorrow.}
\tlhf{TKD}

\tlhw{cha\'logh}
\tlhd{adv}{Twice.}
\tlhf{TKD}

\tlhw{cha\'naS}
\tlhd{n}{A small bird which digs up bugs to eat.}
\tlhf{HolQeD v10n4p4}

\tlhw{cha\'nob}
\tlhd{n}{Ritual gifts.}
\tlhf{KGT}

\tlhw{cha\'par}
\tlhd{n}{A bird noted for its song.}
\tlhf{HolQeD v10n4p4}

\tlhw{cha\'puj}
\tlhd{n}{Dilithium.}
\tlhf{TKD}

\tlhw{cha\'pujqut}
\tlhd{n}{Dilithium crystal.}
\tlhf{TKD}

\tlhw{cha\'qu\'}
\tlhd{n}{A bird with a noisy, repetitive cry.}
\tlhf{HolQeD v10n4p4}

\tlhw{cheb}
\tlhd{n}{Unit weight approx. 5 pounds (2.25kg)}
\tlhf{MSN}

\tlhw{cheb\'a\'}
\tlhd{n}{Unit weight approx. 45-46.4 pounds (20.4-21.05 kg) (9-9.28 cheb)}

\tlhw{Commentary: "The numbers don't quite mesh, but this is the best I can figure from different equivalences we've been given." KLI.org New Words List}
\tlhf{BoP}

\tlhw{chech}
\tlhd{v}{Be drunk, intoxicated.}
\tlhf{TKD}

\tlhw{chechtlhutlh}
\tlhd{n}{/chech-tluth/, a type of liquor.}
\tlhf{KGT}

\tlhw{chegh}
\tlhd{v}{Return, return (to).}

\tlhw{Commentary: Direct object is place returned to.}
\tlhf{TKD, st.k}

\tlhw{cheH}
\tlhd{v}{Defect.}
\tlhf{TKD}

\tlhw{chej}
\tlhd{n}{Liver.}
\tlhf{KGT}

\tlhw{chel}
\tlhd{v}{Add.}
\tlhf{TKD}

\tlhw{chemvaH}
\tlhd{n}{/chemvah/, a type of animal.}
\tlhf{KGT}

\tlhw{chen}
\tlhd{v}{Build up, take form.}

\tlhw{Commentary: "Used in math where the subject is the result of the formula statement (typically using some form of the verb {boq}). The {chen} statement follows the {boq} statement with a semicolon dividing them in Okrand's notation." HolQeD}
\tlhf{TKD, HolQeD v9n3p9}

\tlhw{chenmoH}
\tlhd{v}{Form, make, create.}
\tlhf{KGT}

\tlhw{chenmoHwI\'}
\tlhd{n}{Creator, maker.}
\tlhf{KGT}

\tlhw{cheng name. Chang.}
\tlhf{KGT}

\tlhw{chep}
\tlhd{v}{Prosper, be prosperous.}
\tlhf{TKD}

\tlhw{cher}
\tlhd{v}{Establish, set up.}
\tlhf{TKD}

\tlhw{chergh}
\tlhd{v}{Tolerate.}
\tlhf{TKD}

\tlhw{cheSvel}
\tlhd{n}{Coat, jacket. (general term, regional)}
\tlhf{KGT}

\tlhw{chetvI\'}
\tlhd{n}{1. /chevti/, a spear-throwing device. 2. Torpedo tube.}

\tlhw{Cf. 1. {wob}}
\tlhf{TKD}[KGT]

\tlhw{chev}
\tlhd{v}{Separate.}
\tlhf{TKD}

\tlhw{chevwI\' tlhoy\'}
\tlhd{n}{Territorial wall (e.g. Berlin Wall).}
\tlhf{st.k}

\tlhw{che\'}
\tlhd{v}{1. Rule, reign, run. 2. Preside.}
\tlhf{TKD}

\tlhw{che\'ron}
\tlhd{n}{Battlefield.}
\tlhf{TKD}

\tlhw{chIch}
\tlhd{adv}{Purposely, on purpose, intentionally.}
\tlhf{TKD}

\tlhw{chID}
\tlhd{v}{Admit.}
\tlhf{TKD, TKW p212}

\tlhw{chIj}
\tlhd{v}{Navigate.}
\tlhf{TKD, TKW p63}

\tlhw{chIjwI\'}
\tlhd{n}{Navigator.}
\tlhf{TKD}

\tlhw{chIl}
\tlhd{n}{Lose, misplace.}
\tlhf{TKD}

\tlhw{chIm}
\tlhd{v}{Be empty, deserted, uninhabited.}
\tlhf{TKD}

\tlhw{chIp}
\tlhd{v}{Cut, trim (hair).}
\tlhf{TKD}

\tlhw{chIrgh}
\tlhd{n}{Temple (structure).}
\tlhf{TKD}

\tlhw{chIS}
\tlhd{v}{Be white.}
\tlhf{TKD}

\tlhw{chob}
\tlhd{n}{Corridor.}
\tlhf{BoP}

\tlhw{chob\'a\'}
\tlhd{n}{Main corridor.}
\tlhf{BoP}

\tlhw{choghvat}
\tlhd{n}{1. Stairway leading to door of a ship. 2. Gangplank.}
\tlhf{KGT}

\tlhw{choH n,}
\tlhd{v}{-n. 1. Change. -v. 2. Alter, change.}
\tlhf{TKD}

\tlhw{chol}
\tlhd{v}{1. Close in. 2. Get closer. 3. Come nearer.}
\tlhf{TKD}

\tlhw{choljaH}
\tlhd{n}{Ponytail holder.}
\tlhf{TKD}

\tlhw{chom}
\tlhd{n}{Bartender.}
\tlhf{TKD}

\tlhw{chon}
\tlhd{n}{Hunt.}
\tlhf{KGT}

\tlhw{chon bom}
\tlhd{n}{Hunting song.}
\tlhf{KGT}

\tlhw{chonnaQ}
\tlhd{n}{Hunting spear.}
\tlhf{KGT}

\tlhw{chontay}
\tlhd{n}{Ritual hunt.}
\tlhf{KGT}

\tlhw{chong}
\tlhd{v}{1. Be vertical. 2. Be good, excellent, great, wonderful, splendid. (slang) 3. Be profound, thorough, careful. (slang)}

\tlhw{Commentary: Entries differ in source. They are combined here.}
\tlhf{KGT}

\tlhw{chop}
\tlhd{v}{Bite.}
\tlhf{TKD}

\tlhw{choptaH}
\tlhd{v}{Gnaw.}
\tlhf{KGT}

\tlhw{choq v.t. 1. Preserve. 2. Save.}
\tlhf{TKD, TKW p211}

\tlhw{choQ}
\tlhd{n}{Deck (on a ship).}
\tlhf{BoP}

\tlhw{chor}
\tlhd{n}{1. Belly. 2. Ceramic flat-bottomed pot. (Regional)}

\tlhw{Cf. 2. chor bargh}
\tlhf{TKD}[KGT]

\tlhw{chor bargh}
\tlhd{n}{Ceramic flat-bottomed pot. (Regional)}

\tlhw{Cf. chor}
\tlhf{KGT}

\tlhw{chorgh n, num. -n. 1. Eighth tone of nonatonic musical scale. -num. 2. Eight.}
\tlhf{TKD}[KGT]

\tlhw{chorghDIch num. Eighth.}
\tlhf{TKD}

\tlhw{choS n,}
\tlhd{v}{-n. 1. Twilight. -v. 2. Desert.}
\tlhf{TKD}

\tlhw{chot}
\tlhd{v}{Murder.}
\tlhf{TKD}

\tlhw{chotwI\'}
\tlhd{n}{Murderer.}
\tlhf{KGT 25}

\tlhw{chov}
\tlhd{v}{Assess, evaluate.}
\tlhf{TKD}

\tlhw{chovnatlh}
\tlhd{n}{Specimen.}
\tlhf{TKD}

\tlhw{cho\' n,}
\tlhd{v}{-n. 1. Succession. -v. 2. Succeed (to authority).}
\tlhf{TKD}

\tlhw{chuch}
\tlhd{n}{Ice.}
\tlhf{TKD, TKW p190}

\tlhw{chuH}
\tlhd{v}{1. Throw (a spear) at, hurl (a spear) at. 2. Explain clearly to, clarify for, specify for. (slang)}

\tlhw{Cf. 2. {chuHchu\'}}
\tlhf{KGT}

\tlhw{chuHchu\'}
\tlhd{v}{Hit with a thrown spear.}
\tlhf{KGT 64}

\tlhw{chun}
\tlhd{v}{Be innocent.}
\tlhf{TKD}

\tlhw{chunDab}
\tlhd{n}{Meteor.}
\tlhf{TKD}

\tlhw{chung}
\tlhd{v}{Accelerate.}
\tlhf{TKD}

\tlhw{chup}
\tlhd{v}{Recommend, suggest.}
\tlhf{TKD}

\tlhw{chuq}
\tlhd{n}{1. Range. 2. Distance.}
\tlhf{TKD}

\tlhw{chuq\'a\'}
\tlhd{n}{Long range.}
\tlhf{BoP}

\tlhw{chuQun}
\tlhd{n}{Nobility.}
\tlhf{TKD}

\tlhw{chuQ\'a\'}
\tlhd{n}{Long range.}
\tlhf{BoP}

\tlhw{chuS}
\tlhd{v}{Be noisy.}
\tlhf{TKD}

\tlhw{chuS\'ugh}
\tlhd{n}{Type of musical instrument.}
\tlhf{TKD}

\tlhw{chut}
\tlhd{n}{Law.}
\tlhf{TKD}

\tlhw{chuv}
\tlhd{v}{Be left over.}
\tlhf{TKD}

\tlhw{chuvmey pl.n. Leftovers (grammatical term).}
\tlhf{TKD}

\tlhw{chuy}
\tlhd{v}{Sneeze.}
\tlhf{HolQeD v12n4p8}

\tlhw{chuyDaH}
\tlhd{n}{Thrusters.}
\tlhf{TKD}

\tlhw{chu\'}
\tlhd{v}{1. Be new. 2. Engage. 3. Activate (a device). 4. Play (a musical instrument).}

\tlhw{Cf. 4. {chu\'wI\'}}
\tlhf{TKD}[KGT]

\tlhw{chu\'wI\'}
\tlhd{n}{1. Trigger. 2. Player (of an instrument).}

\tlhw{Cf. 2. {chu\'}}
\tlhf{TKD}[KGT]
