\chapter{\tlhe{D}}

\tlhw{Da}
\tlhd{v}{Behave as.}[Act in the manner of.]
\tlhf{TKD}

\tlhw{Dab}
\tlhd{v}{Dwell in/at, reside in/at.}
\tlhf{KGT}

\tlhw{Dach}
\tlhd{v}{Be absent.}[Be not attentive, be distracted, lack focus, to not pay attention. (slang)]
\tlhco{Entries differs in source. They are combined here.}{TKD}
\tlhf{TKD}[KGT]

\tlhw{Daghtuj}
\tlhd{n}{Mixture of animal parts, animal parts mixed together.}
\tlhf{KGT}

\tlhw{DaH}
\tlhd{adv}{Now.}
\tlhd{n}{Array, bank (of weapons).}
\tlhf{TKD, BoP}[KGT 56]

\tlhw{DaHjaj}
\tlhd{n}{Today.}
\tlhf{TKD}[TKW p203]

\tlhw{DaHjaj gheD}
\tlhd{n}{Dish in a restaurant, "catch of the day".}
\tlhf{KGT}

\tlhw{Daj}
\tlhd{v}{Be interesting.}[Test inconclusively.]
\tlhf{TKD}[TKW p104]

\tlhw{Dal}
\tlhd{v}{Be boring.}
\tlhf{TKD}[TKW p201]

\tlhw{Dan}
\tlhd{v}{Occupy (military term).}
\tlhf{TKD}

\tlhw{Dap}
\tlhd{n}{Nonsense.}
\tlhf{TKD}

\tlhw{Dap bom}
\tlhd{n}{Nonsense song.}
\tlhf{KGT}

\tlhw{Daq}
\tlhd{n}{Site, location.}
\tlhd{v}{Eavesdrop.}
\tlhf{TKD}

\tlhw{Daqtagh}
\tlhd{n}{/d'k tahg/, warrior's knife.}
\tlhf{KGT}

\tlhw{DaQ}
\tlhd{n}{Ponytail.}
\tlhf{TKD}

\tlhw{Dargh}
\tlhd{n}{Tea.}
\tlhf{TKD}

\tlhw{Dargh HIvje\'}
\tlhd{n}{Teacup.}
\tlhf{KGT}

\tlhw{Dargh HIvje\' jengva\' je}
\tlhd{n}{Cup and 'saucer.}
\tlhf{Eurotalk}

\tlhw{DarSeq}
\tlhd{n}{/darsek/, unit of currency.}
\tlhf{KGT}

\tlhw{DaS}
\tlhd{n}{Boot.}
\tlhf{TKD}

\tlhw{DaSpu\'}
\tlhd{n}{Boot spike.}
\tlhf{TKD}

\tlhw{Dat}
\tlhd{n}{Everywhere.}
\tlhf{TKD}

\tlhw{Dav}
\tlhd{v}{Sway (aircraft moves to the side without yawning).}
\tlhco{It is unknown if this has meaning outside of the context of aircraft attitude descriptions.}{HolQeD}
\tlhf{HolQeD v11n2p9}

\tlhw{DavHam}
\tlhd{n}{False honor.}
\tlhf{HolQeD v12n3p9}

\tlhw{Daw\'}
\tlhd{n}{Revolt, revolution.}
\tlhd{v}{Revolt.}
\tlhf{TKD}

\tlhw{Day}
\tlhd{n}{The consonant "D".}
\tlhf{qep\'a\' wa\'maH javDIch 2009}

\tlhw{DaypuS}
\tlhd{n}{/dikus/, type of plant.}
\tlhf{KGT}

\tlhw{Da\'}
\tlhd{n}{Corporal (rank).}
\tlhf{TKD}

\tlhw{Da\'nal}
\tlhd{n}{A bird characterized by erratic, unpredictable behavior.}
\tlhcf{Da'vI'}
\tlhf{HolQeD v10n4p4}

\tlhw{Da\'vI\'}
\tlhd{n}{A bird characterized by erratic, unpredictable behavior.}
\tlhcf{Da'nal}
\tlhf{HolQeD v10n4p4}

\tlhw{Deb}
\tlhd{n}{Desert.}
\tlhf{TKD}

\tlhw{Dech}
\tlhd{v}{Surround.}
\tlhf{TKD}

\tlhw{Degh}
\tlhd{n}{Helm.}[Medal, emblem, symbol, insignia.]
\tlhd{v}{Act without a plan, improvise. (slang)}
\tlhf{TKD}

\tlhw{DeghwI\'}
\tlhd{n}{Helmsman.}
\tlhf{TKD}

\tlhw{DeH}
\tlhd{v}{Be ripe, be overripe.}

\tlhw{Fruit, vegetable.}
\tlhf{KGT}

\tlhw{Dej}
\tlhd{v}{Collapse.}
\tlhf{TKD}

\tlhw{Del}
\tlhd{v}{Describe.}
\tlhf{TKD}

\tlhw{DenIb}
\tlhd{n}{Denebia.}
\tlhcf{DenIbya'}
\tlhco{In the E--K TKD, spelled \tlhe{DenIbya\'}.}{TKD}
\tlhf{TKD}

\tlhw{DenIb Qatlh}
\tlhd{n}{Denebian slime devil.}
\tlhcf{DenIbya' Qatlh}
\tlhco{In the E--K TKD, spelled \tlhe{DenIbya\' Qatlh}.}{TKD}
\tlhf{TKD}

\tlhw{DenIbngan}
\tlhd{n}{Denebian.}
\tlhcf{DenIbya'ngan}
\tlhco{In the E--K TKD, spelled \tlhe{DenIb}.}{TKD}
\tlhf{TKD}

\tlhw{DenIbya\'}
\tlhd{n}{Denebia.}
\tlhcf{DenIb}
\tlhco{In the K--E TKD, spelled \tlhe{DenIb}.}{TKD}
\tlhf{TKD}

\tlhw{DenIbya\'ngan}
\tlhd{n}{Denebian.}
\tlhcf{DenIbngan}
\tlhco{In the K--E TKD, spelled \tlhe{DenIb}.}{TKD}
\tlhf{TKD}

\tlhw{DenIbya\' Qatlh}
\tlhd{n}{Denebian slime devil.}
\tlhcf{DenIb Qatlh}
\tlhco{In the K--E TKD, spelled \tlhe{DenIb Qatlh}.}{TKD}
\tlhf{TKD}

\tlhw{Denmargh}
\tlhd{n}{Denmark.}
\tlhf{Eurotalk}

\tlhw{Dep}
\tlhd{n}{Being (nonhumanoid).}
\tlhf{TKD}

\tlhw{DeQ}
\tlhd{n}{Credit (monetary unit).}
\tlhf{TKD}

\tlhw{Der}
\tlhd{v}{Yaw (aircraft nose points left or right).}
\tlhco{It is unknown if this has meaning outside of the context of aircraft attitude descriptions.}{HolQeD}
\tlhf{HolQeD v11n2p9}

\tlhw{DeS}
\tlhd{n}{Arm (body part).}[Handle, ax handle.]
\tlhf{TKD}[KGT]

\tlhw{DeSqIv}
\tlhd{n}{Elbow.}
\tlhf{KGT}

\tlhw{Dev}
\tlhd{v}{Lead, guide.}
\tlhf{TKD}[TKW piii]

\tlhw{DevwI\'}
\tlhd{n}{Leader.}
\tlhf{TKW}[KGT 63,66]

\tlhw{De\'}
\tlhd{n}{Data, information.}
\tlhf{TKD}

\tlhw{De\'wI\'}
\tlhd{n}{Computer.}
\tlhf{TKD}

\tlhw{De\' chu\' ghItlh}
\tlhd{n}{Newspaper.}
\tlhf{Eurotalk}

\tlhw{De\' jengva\'}
\tlhd{n}{Compact disc, CD.}
\tlhf{Eurotalk}

\tlhw{DI}
\tlhd{n}{Litter, rubble, debris, garbage.}
\tlhf{TKD}[TKW p104]

\tlhw{DIb}
\tlhd{n}{Right, privilege.}
\tlhf{TKD}[KGT]

\tlhw{DIch}
\tlhd{n}{Certainty.}
\tlhf{TKD}

\tlhw{DIghna\'}
\tlhd{n}{Type of plant.}
\tlhf{KGT}

\tlhw{DIj}
\tlhd{v}{Slide sword blade along opponent's blade.}[Use a pigment stick, paint with pigment stick.]
\tlhf{KGT}

\tlhw{DIl}
\tlhd{v}{Pay for.}
\tlhf{TKD}

\tlhw{DIlyum}
\tlhd{n}{Trillium.}
\tlhco{An unspecified commodity. In the TOS episode "Errand of Mercy", Spock poses as a Vulcan trader in kevas and trillium. Thus, both apear in TKD. Star Trek Enchyclopedia lists them as a gemstone, established in DS9 "Trials and Tribble-ations".}{TKD}
\tlhf{TKD}

\tlhw{DIn}
\tlhd{n}{Open entryway (to corridor, tunnel, conduit, Jeffries tube, branch of sewer).}
\tlhcf{DIn][lojmIt}
\tlhco{This is the open entryway of any enclosed space longer than wide in which people might find themselves. If there is a door that closes, this is not a \tlhe{DIn}. It is merely a \tlhe{lojmIt}.}{HolQeD}
\tlhf{HolQeD v12n2p7}

\tlhw{DIng}
\tlhd{v}{Spin.}
\tlhf{TKD}

\tlhw{DIp}
\tlhd{n}{Noun.}
\tlhf{TKD}

\tlhw{DIr}
\tlhd{n}{Skin.}
\tlhf{TKD}

\tlhw{DIron}
\tlhd{n}{Type of plant.}
\tlhf{KGT}

\tlhw{DIr qanwI\' taS}
\tlhd{n}{Suntan lotion.}
\tlhf{Eurotalk}

\tlhw{DIr \'In}
\tlhd{n}{Drum.}
\tlhco{Percussion instrument with a stretched animal skin.}{KGT}
\tlhf{KGT 75}

\tlhw{DIS}
\tlhd{n}{Cave.}[Year (Klingon).]
\tlhd{v}{Confess.}
\tlhf{TKD}

\tlhw{DIv}
\tlhd{v}{Be guilty.}
\tlhf{TKD}

\tlhw{DIvI\'}
\tlhd{n}{Federation, organization, association, league, union.}
\tlhf{TKD}[HolQeD v4n4p11]

\tlhw{DIvI\'may\'Duj}
\tlhd{n}{Federation battle cruiser.}
\tlhf{TKD}

\tlhw{Do}
\tlhd{n}{Velocity.}
\tlhf{TKD}

\tlhw{Doch}
\tlhd{n}{Thing.}
\tlhd{v}{Be rude.}
\tlhf{TKD}

\tlhw{DoD}
\tlhd{n}{Mark (in coordinates).}
\tlhf{TKD}

\tlhw{Dogh}
\tlhd{v}{Be foolish, silly.}
\tlhf{TKD}

\tlhw{Doghjey}
\tlhd{n}{Unconditional surrender.}
\tlhf{TKD}

\tlhw{DoH}
\tlhd{v}{Back away from, back off, get away from.}
\tlhf{TKD}[TKW p211]

\tlhw{Doj}
\tlhd{v}{Be impressive.}
\tlhf{TKD}

\tlhw{Dol}
\tlhd{n}{Entity, whole.}
\tlhf{TKD}[TKW p209]

\tlhw{Dom}
\tlhd{n}{Radan (crude dilithium crystal).}
\tlhf{TKD}

\tlhw{Don}
\tlhd{v}{Be parallel, go parallel to.}
\tlhf{TKD}

\tlhw{Dop}
\tlhd{n}{Side.}
\tlhf{TKD}

\tlhw{Doq}
\tlhd{v}{Be orange, red.}
\tlhf{TKD}

\tlhw{Doqqu\' \'ej wov}
\tlhd{v}{Be pink.}
\tlhf{Eurotalk}

\tlhw{Doq \'ej SuD}
\tlhd{v}{Be purple.}
\tlhf{Eurotalk}

\tlhw{Doq \'ej wovbe\'}
\tlhd{v}{Be brown.}
\tlhf{HolQeD v8n1p7}

\tlhw{DoQ}
\tlhd{v}{Claim (territory).}
\tlhf{TKD}

\tlhw{Dor}
\tlhd{v}{Escort.}[End.]
\tlhco{The subject of this verb is a period or unit of time. This verb is not adjectival and takes no direct object.}{HolQeD}
\tlhf{TKD}[HolQeD v12n2p8]

\tlhw{DoS}
\tlhd{n}{Target.}
\tlhf{TKD}[TKW p81,191]

\tlhw{DoS chIl}
\tlhd{v}{To digress.}
\tlhco{"I digress" would be \tlhw{DoS vIchIl}.}
\tlhf{(Marc Okrand to Lawrence M. Schoen on 28 July 2008)}

\tlhw{Dotlh}
\tlhd{n}{Status.}
\tlhf{TKD}

\tlhw{Dov\'agh}
\tlhd{n}{Flute, fife.}
\tlhf{KGT}

\tlhw{DoyIchlan}
\tlhd{n}{Germany.}
\tlhf{Eurotalk}

\tlhw{Doy\'}
\tlhd{v}{Be tired.}
\tlhf{TKD}

\tlhw{Doy\'yuS}
\tlhd{n}{Troyius.}
\tlhf{TKD}

\tlhw{Do\'}
\tlhd{adv}{Luckily, with luck.}
\tlhd{v}{Be fortunate, lucky.}
\tlhf{TKD}

\tlhw{Do\'Ha\'}
\tlhd{adv}{Unfortunately.}
\tlhd{v}{Be unfortunate.}
\tlhf{TKD}[KGT]

\tlhw{Dub}
\tlhd{n}{Back (of body).}
\tlhd{v}{Improve.}
\tlhf{TKD}

\tlhw{DuD}
\tlhd{v}{Mix.}
\tlhf{TKD}

\tlhw{DuDwI\'}
\tlhd{n}{Stirring stick, mixing stick.}
\tlhf{KGT}

\tlhw{Dugh}
\tlhd{v}{Be vigilant.}
\tlhf{TKD}

\tlhw{DuH}
\tlhd{n}{Possibility, option.}
\tlhd{v}{Be possible.}
\tlhf{TKD}

\tlhw{Duj}
\tlhd{n}{Ship, vessel.}[Instincts.]
\tlhf{TKD}[TKW p27,63,74,182]

\tlhw{Duj ngaDHa\'}
\tlhd{n}{Irresponsible person, undisciplined person. (slang)}
\tlhf{KGT}

\tlhw{Dum}
\tlhd{v}{Nap.}
\tlhf{TKD}

\tlhw{Dun}
\tlhd{v}{Be wonderful, great.}
\tlhf{TKD}[TKW p15]

\tlhw{Dung}
\tlhd{n}{Area above, area overhead.}
\tlhf{TKD}

\tlhw{DungluQ}
\tlhd{n}{Noon.}
\tlhf{TKD}

\tlhw{Dup}
\tlhd{n}{Strategy.}
\tlhf{TKD}

\tlhw{Duq}
\tlhd{n}{Bowl (small).}
\tlhf{KGT}

\tlhw{DuQ}
\tlhd{v}{Stab, prick.}[Touch (emotionally).]
\tlhf{TKD}[TKW p131]

\tlhw{DuQwI\'}
\tlhd{n}{Spike.}
\tlhf{KGT}

\tlhw{DuQwI\'Hommey}
\tlhd{pl n}{Small spikes on the /d'k tahg/ pommel.}
\tlhf{KGT 61}

\tlhw{Duran lung DIr}
\tlhd{n}{Durani lizard skins.}
\tlhco{A food item which never uses a plural suffix.}{KCD}
\tlhf{KCD}

\tlhw{DuraS}
\tlhd{name}{Duras.}
\tlhf{HolQeD v5n3p15}[S26]

\tlhw{DuS}
\tlhd{n}{Torpedo tube.}
\tlhf{TKD}

\tlhw{DuSaQ}
\tlhd{n}{School.}
\tlhf{TKD}

\tlhw{Duv}
\tlhd{v}{Advance.}
\tlhf{TKD}

\tlhw{Duy}
\tlhd{n}{Agent, emissary.}
\tlhf{TKD}

\tlhw{Duy\'}
\tlhd{n}{Defect.}
\tlhd{v}{Be defective.}
\tlhf{TKD}

\tlhw{Duy\'a\'}
\tlhd{n}{Ambassador.}
\tlhf{TKD}

\tlhw{Du\'}
\tlhd{n}{Farm.}
\tlhf{TKD}

\tlhw{Du\' naH}
\tlhd{n}{Produce.}
\tlhf{KGT}