\chapter{\tlhe{ng}}

\tlhw{ngab}
\tlhd{v}{Disappear.}[Vanish.]
\tlhf{TKD}

\tlhw{ngach}
\tlhd{v}{Debate.}
\tlhf{TKD}

\tlhw{ngan}
\tlhd{n}{Inhabitant.}
\tlhf{TKD}

\tlhw{ngaq}
\tlhd{n}{Support (military term).}
\tlhf{TKD}

\tlhw{ngaS}
\tlhd{v}{Contain (have inside).}
\tlhf{TKD}

\tlhw{ngat}
\tlhd{n}{Gunpowder.}
\tlhf{TKD}

\tlhw{ngav}
\tlhd{n}{Writer's cramp.}
\tlhf{TKD}

\tlhw{ngay}
\tlhd{n}{The consonant "ng".}
\tlhf{qep\'a\' wa\'maH javDIch 2009}

\tlhw{nga\'chuq}
\tlhd{v}{Sex (i.e, perform sex; always subject) }

\tlhw{It is unclear about what part of speech this may be, though it is probably a verb and "always subject" probably refers to the concept that all involved parties collectively make the subject of this verb.}
\tlhf{veS QonoS}[HolQeD v1n3p9]

\tlhw{ngeb}
\tlhd{v}{Be counterfeit.}[Be false.][Be fake.]
\tlhf{TKD}

\tlhw{ngech}
\tlhd{n}{Valley.}[Cleavage. (slang)]
\tlhf{TKD}

\tlhw{ngeD}
\tlhd{v}{Be easy.}
\tlhf{TKD}

\tlhw{ngeH}
\tlhd{v}{Send.}
\tlhf{TKD}

\tlhw{ngeHbej}
\tlhd{n}{Cosmos.}
\tlhf{TKD}

\tlhw{ngej}
\tlhd{v}{Infect.}
\tlhf{TKD}

\tlhw{ngem}
\tlhd{n}{Forest, woods.}
\tlhf{TKD}

\tlhw{ngeng}
\tlhd{n}{Lake.}
\tlhf{TKD}

\tlhw{ngep}
\tlhd{v}{Override.}
\tlhf{TKD}

\tlhw{nger}
\tlhd{n}{Theory.}
\tlhf{TKD}

\tlhw{ngev}
\tlhd{v}{Sell.}
\tlhf{TKD}

\tlhw{nge\'}
\tlhd{v}{Take away.}
\tlhf{TKD}

\tlhw{ngIl}
\tlhd{v}{Dare.}
\tlhf{TKD}

\tlhw{ngIm}
\tlhd{v}{Be putrid.}
\tlhf{TKD}

\tlhw{ngIp}
\tlhd{v}{Borrow.}
\tlhf{TKD}

\tlhw{ngIq}
\tlhd{n}{Singular one.}[Individual one.]
\tlhco{There were two distinct examples of this word's use seen in the opera. One of them was obviously detailing how a group of warriors was killed one at a time: \tlhtr{ngIq nuv luHoH}{they killed the warriors one by one.} The other was a thrice-repeated \tlhe{ngIq tonSaw lo\' \'ej...}{in one single move, he...}, making it clear that each of several results was the result of the same action. I surmise that \tlhe{ngIq} by itself could mean "the one and only". Again, I didn\'t copy the lyrics, and what I have written here is probably not exactly what was on the pages we inspected.}{ghunchu\'wI\' \'utlh}
\tlhf{qep\'a\' wa\'maH chorghDIch 2011}

\tlhw{ngIv}
\tlhd{v}{Patrol.}
\tlhf{TKD}

\tlhw{ngoD}
\tlhd{n}{Fact.}
\tlhf{TKD}

\tlhw{ngoch}
\tlhd{n}{Policy.}
\tlhf{TKD}

\tlhw{ngoH}
\tlhd{v}{Smear.}
\tlhf{TKD}

\tlhw{ngoj}
\tlhd{v}{Be restless.}
\tlhf{TKD}

\tlhw{ngong}
\tlhd{n}{Experiment.}
\tlhd{v}{Experiment.}
\tlhf{TKD}

\tlhw{ngoq}
\tlhd{n}{Code.}
\tlhf{TKD}

\tlhw{ngoQ}
\tlhd{n}{Goal.}
\tlhf{TKD}

\tlhw{ngor}
\tlhd{v}{Cheat.}
\tlhf{TKD}

\tlhw{ngoS}
\tlhd{v}{Dissolve.}
\tlhf{TKD}

\tlhw{ngotlh}
\tlhd{v}{Be fanatical.}
\tlhf{TKD}

\tlhw{ngoy\'}
\tlhd{v}{Be responsible.}
\tlhf{TKD}

\tlhw{ngo\'}
\tlhd{v}{Be old (not new).}
\tlhf{TKD}

\tlhw{ngup}
\tlhd{n}{Cape (clothing).}
\tlhf{TKD}

\tlhw{ngu\'}
\tlhd{v}{Identify.}
\tlhf{TKD}