\chapter{\tlhe{gh}}

\tlhw{ghab}
\tlhd{n}{Meat from midsection of animal.}
\tlhf{KGT}

\tlhw{ghab tun}
\tlhd{n}{Meat from midsection of animal,n.bones. (regional)}
\tlhf{KGT}

\tlhw{ghagh}
\tlhd{v}{Gargle.}
\tlhf{TKD}

\tlhw{ghaH}
\tlhd{pro}{He, she, him, her.}
\tlhf{TKD}[TKW p61,139,190]

\tlhw{ghaj}
\tlhd{v}{Have, possess.}
\tlhf{TKD}[TKW p142]

\tlhw{ghal}
\tlhd{n}{Be jealous (of), envy.}
\tlhf{HolQeD v8n3p4}

\tlhw{gham}
\tlhd{n}{Limb (of an animal).}
\tlhf{KGT}

\tlhw{ghanjaq}
\tlhd{n}{Mace, stick, club.}
\tlhf{KGT}

\tlhw{ghang}
\tlhd{v}{End prematurely (an event, voyage, battle, play, opera, story, song, etc.)}
\tlhco{The subject of this verb is the person causing an event to prematurely end. The direct object is the event that is ended.}{HolQeD}
\tlhf{HolQeD v11n2p8}

\tlhw{ghantoH}
\tlhd{n}{model, example, pattern.}
\tlhco{Maltz is assuming this refers to something one copies or refers to in making things -- or something along those lines.}{M.Ok.}
\tlhf{(email to Agnieszka, 15 january 2012)}

\tlhw{ghangwI\'}
\tlhd{n}{Horizon.}
\tlhf{HolQeD v11n2p8}

\tlhw{ghap}
\tlhd{conj}{Or, either/or (joining nouns).}
\tlhf{TKD}[TKW p119]

\tlhw{ghaq}
\tlhd{v}{Contribute.}
\tlhf{TKD}

\tlhw{ghar}
\tlhd{n}{Diplomacy.}
\tlhd{v}{Conduct diplomacy.}
\tlhf{TKD}

\tlhw{gharlIq \'oQqar}
\tlhd{n}{Garlic.}
\tlhf{Eurotalk}

\tlhw{ghargh}
\tlhd{n}{Serpent, worm.}
\tlhf{TKD}

\tlhw{gharwI\'}
\tlhd{n}{diplomat.}
\tlhf{TKD}

\tlhw{ghatlh}
\tlhd{v}{Dominate.}
\tlhf{TKD}

\tlhw{ghawran}
\tlhd{name}{Gowron.}
\tlhf{KGT, HolQeD v5n3p15, SB25}[SB26]

\tlhw{ghaw\'}
\tlhd{n}{/igvah/ liver soup.}[Insecure on, one full of selt-doubt, self-doubter. (regional, slang)]
\tlhco{Entries differ in source. They are combined here.}{\emph{ed}}
\tlhf{KGT}

\tlhw{ghay}
\tlhd{n}{The consonant "gh".}
\tlhf{qep\'a\' wa\'maH javDIch 2009}

\tlhw{ghay}
\tlhd{v}{Spray, [bombard?]}
\tlhco{Can take \tlhw{jorwI'} as its object}{\emph{Ed}}
\tlhf{(Klingon Monopoly, published 1 November 2011)}

\tlhw{ghaytan}
\tlhd{adv}{Likely.}
\tlhf{TKD}

\tlhw{ghaytanHa\'}
\tlhd{adv}{Unlikely.}
\tlhf{British TV Magazine}["Radio Times"]

\tlhw{ghay\'cha\'}
\tlhd{excl}{General invective.}
\tlhf{TKD}

\tlhw{gha\'tlhIq}
\tlhd{n}{Ode of respect.}
\tlhf{KGT}

\tlhw{gheb}
\tlhd{n}{Horn (musical instrument).}
\tlhf{HolQeD v10n2p9}

\tlhw{gheD}
\tlhd{n}{Prey.}
\tlhf{TKW}[KGT 161]

\tlhw{ghegh}
\tlhd{v}{Be rough.}
\tlhf{TKD}

\tlhw{ghel}
\tlhd{v}{Ask (a question).}
\tlhf{KGT}

\tlhw{ghem}
\tlhd{n}{Midnight snack.}
\tlhf{TKD}

\tlhw{gher v.Formulate, compile, pull together.}
\tlhf{BBS}

\tlhw{gheS}
\tlhd{v}{Assume duties of, take on responsibilities of.}
\tlhf{KGT}

\tlhw{ghet}
\tlhd{v}{Pretend.}
\tlhco{No deception is implied; simply role-playing.}{HolQeD}
\tlhf{HolQeD v12n3p8}

\tlhw{ghetwI\'}
\tlhd{v}{Pretender.}
\tlhco{No deception is implied; simply role-playing.}{HolQeD}
\tlhf{HolQeD v12n3p8}

\tlhw{ghevchoq Sep}
\tlhd{n}{Gevchok region.}
\tlhf{KGT}

\tlhw{ghevI\'}
\tlhd{n}{Sauce for /gagh/.}
\tlhf{KGT}

\tlhw{ghew}
\tlhd{n}{Bug, cootie.}
\tlhf{TKD}[TKW p207]

\tlhw{ghe\'naQ}
\tlhd{n}{Opera.}
\tlhf{KGT}

\tlhw{ghe\'tor}
\tlhd{n}{Gre'thor (where spirits of the dishonored go).}
\tlhf{KGT}

\tlhw{ghe\'\'or}
\tlhd{n}{Netherworld (where dishonored go).}
\tlhf{TKD}

\tlhw{ghIb}
\tlhd{v}{Consent.}
\tlhf{TKD}

\tlhw{ghIch}
\tlhd{n}{Nose.}
\tlhf{TKD}

\tlhw{ghIgh}
\tlhd{n}{Necklace.}[Assignment, task, duty. (slang)]
\tlhf{TKD}[KGT]

\tlhw{ghIH}
\tlhd{v}{Be messy, sloppy.}
\tlhf{TKD}

\tlhw{ghIj}
\tlhd{v}{Scare.}
\tlhf{TKD}[TKW]

\tlhw{ghIlab ghew}
\tlhd{n}{Glob fly.}
\tlhf{TKW}[KGT]

\tlhw{ghIlaSnoS}
\tlhd{n}{Glasnost, an ancient political movement.}
\tlhf{TKW}

\tlhw{ghIm}
\tlhd{v}{Exile.}
\tlhf{TKD}

\tlhw{ghIn}
\tlhd{n}{/gin\'tak/, battle spear.}[Monastery.]
\tlhco{This is a pretty general term for a religious community (and the term ``religious'' could be interpreted in various ways as well) , so it can be modified. A \tlhe{ghIn\'a\'} would be a pretty important monastery, for example.}{M.Ok.}[][2]
(qepHom 2010, email to Lieven, 15 November 2010)
\tlhf{TKD}

\tlhw{ghIpDIj}
\tlhd{v}{Court-martial.}
\tlhf{TKD}

\tlhw{ghIq}
\tlhd{adv}{Then, subsequently.}
\tlhf{HolQeD v8n3p4}

\tlhw{ghIqtal}
\tlhd{excl}{To the death. (archaic)}
\tlhf{TKD}

\tlhw{ghIQ}
\tlhd{v}{Vacation, take a vacation.}
\tlhf{TKD}

\tlhw{ghIr}
\tlhd{v}{Descend.}
\tlhf{TKD}

\tlhw{ghIrep naHmey}
\tlhd{pl n}{Grapes.}
\tlhf{Eurotalk}

\tlhw{ghIrIlqa\'}
\tlhd{name}{Grilka.}
\tlhf{KGT}

\tlhw{ghISDen}
\tlhd{n}{Scales (animal covering).}
\tlhf{KGT}

\tlhw{ghIt}
\tlhd{n}{Ax blade, blade of ax.}[Open, flat hand.]
\tlhco{Entries differ in source. They are combined here.}{\emph{ed}}
\tlhf{KGT}

\tlhw{ghItlh}
\tlhd{n}{Manuscript.}
\tlhd{v}{Write.}[Engrave, incise.][Mark (upon).]
\tlhco{Entries differ in souce. The K->E entry is used here.}
\tlhco{'To mark (upon)' something is \tlhe{ghItlh}. This isn\'t just writing; it\'s any kind of marking. (Note that \tlhe{ghItlh}, \'to write,\' refers to the physical act of writing. It doesn\'t mean \'to create a composition.\')}{Okrand\'s notes, HolQeD v2n4p18}
\tlhf{TKD}[KGT][HolQeD v2n4p18]

\tlhw{ghItlhwI\'}
\tlhd{n}{Stylus, writing implement, pen.}[Person who writes.]
\tlhf{KGT}

\tlhw{gho}
\tlhd{n}{Circle.}[Hoop.]
\tlhf{TKD}

\tlhw{ghob}
\tlhd{n}{Ethics.}[Virtue.]
\tlhd{v}{Wage war.}[Battle.][Fight.]
\tlhf{TKW}[KGT]

\tlhw{ghobchuq loDnI\'pu\'}
\tlhd{name}{The Brothers Fight One Another - famous statue.}
\tlhf{KCD}

\tlhw{ghobe\'}
\tlhd{excl}{No (answer to a question).}
\tlhf{TKD}

\tlhw{ghoch}
\tlhd{n}{Destination.}
\tlhd{v}{Track, track down.}
\tlhf{TKD}

\tlhw{ghochwI\'}
\tlhd{n}{The constellation "Tracker".}
\tlhf{KCD}

\tlhw{ghoD}
\tlhd{v}{Stuff.}
\tlhf{TKD}

\tlhw{ghogh}
\tlhd{n}{Voice.}
\tlhf{TKD}[TKW]

\tlhw{ghogh HablI\'}
\tlhd{n}{Telephone.}
\tlhf{HolQeD v5n2p20}

\tlhw{ghoH}
\tlhd{v}{Argue, dispute.}
\tlhf{TKD}

\tlhw{ghoj}
\tlhd{v}{Learn.}
\tlhf{TKD}

\tlhw{ghojmeH taj}
\tlhd{n}{Boy's knife.}
\tlhf{KCD}

\tlhw{ghojmoH}
\tlhd{v}{Teach, instruct.}
\tlhf{TKD}

\tlhw{ghojmoH ghItlhwI\'}
\tlhd{n}{Pencil.}
\tlhf{Eurotalk}

\tlhw{ghojmoq}
\tlhd{n}{Nurse, nanny, governess.}
\tlhf{TKD}

\tlhw{ghojwI\'}
\tlhd{n}{Student.}
\tlhf{TKD}[TKW]

\tlhw{ghol}
\tlhd{n}{Opponent, adversary.}
\tlhf{TKD}

\tlhw{ghom}
\tlhd{n}{Group, party, band.}
\tlhd{v}{Meet, encounter, assemble, rendezvous.}
\tlhf{TKD, TKW}[KGT]

\tlhw{ghomHa\'}
\tlhd{v}{Scatter, disperse.}
\tlhf{TKD}

\tlhw{ghom\'a\'}
\tlhd{n}{Crowd.}
\tlhf{TKD}

\tlhw{ghonDoq}
\tlhd{n}{Type of knife (slender blade).}
\tlhf{KGT}

\tlhw{ghong}
\tlhd{n}{Abuse.}
\tlhd{v}{Abuse.}
\tlhf{TKD}

\tlhw{ghop}
\tlhd{n}{Hand.}
\tlhf{TKD}

\tlhw{ghopDap}
\tlhd{n}{Asteroid.}
\tlhf{TKD}

\tlhw{ghoq}
\tlhd{v}{Spy, spy on.}
\tlhf{TKD}[KGT]

\tlhw{ghoqwI\'}
\tlhd{n}{Spy.}
\tlhf{TKD}

\tlhw{ghoQ}
\tlhd{v}{Be fresh, just be killed (meat).}
\tlhf{KGT}

\tlhw{ghor}
\tlhd{n}{Surface (of a planet).}
\tlhd{v}{Break.}
\tlhf{TKD}[TKW]

\tlhw{ghorgh}
\tlhd{ques}{When?.}
\tlhf{TKD}

\tlhw{ghoS}
\tlhd{v}{Approach, go away from, proceed, follow (a course).}[Thrust.]
\tlhf{TKD}

\tlhw{ghot}
\tlhd{n}{Person (humanoid).}
\tlhf{TKD}[TKW]

\tlhw{ghotI\'}
\tlhd{n}{Fish, most general word for fish-like creature.}

\tlhw{PUN: George Bernard-Shaw\'s wordplay joke about spelling "fish" with the "gh" from enough, the "o" from women, and the "ti" from nation.}
\tlhf{HolQeD v10n4p5}

\tlhw{ghov}
\tlhd{v}{Recognize.}
\tlhf{TKD}

\tlhw{gho\'}
\tlhd{v}{Step on.}
\tlhf{TKD}

\tlhw{gho\'Do}
\tlhd{n}{Sublight speed.}
\tlhf{TKD}

\tlhw{ghu}
\tlhd{n}{Baby.}
\tlhf{TKD}

\tlhw{ghubDaQ}
\tlhd{n}{First born child.}
\tlhf{HolQeD v9n3p9}

\tlhw{ghuH}
\tlhd{n}{Alert.}
\tlhd{v}{Prepare for, be alerted to.}
\tlhf{TKD}[TKW]

\tlhw{ghuHmoH}
\tlhd{v}{Alert, warn.}
\tlhf{TKD}

\tlhw{ghum}
\tlhd{n}{Alarm.}
\tlhd{v}{Alarm, sound an alarm.}
\tlhf{TKD}

\tlhw{ghun}
\tlhd{v}{Program (a computer).}
\tlhf{TKD}

\tlhw{ghung}
\tlhd{v}{Be hungry.}
\tlhf{TKD}

\tlhw{ghup}
\tlhd{v}{Swallow.}
\tlhf{TKD}

\tlhw{ghur}
\tlhd{v}{Increase.}
\tlhf{TKD}

\tlhw{ghuS}
\tlhd{v}{Be prepared, ready (to launch).}[Lower (spear) to horizontal to attack.]
\tlhf{TKD}[KGT]

\tlhw{ghuv}
\tlhd{n}{Recruit.}
\tlhf{TKD}

\tlhw{ghuy\'}
\tlhd{excl}{Klingon curse, damn.}
\tlhf{ST5}[HolQeD v8n4p13]

\tlhw{ghuy\'cha\'}
\tlhd{excl}{Klingon curse, damn.}
\tlhca{This is the curse one would use after receiving an unsettling communiqué. It is far worse than "Darn it!"}{CK}
\tlhco{Probably akin to "goddamn it!"}{vo'an}
\tlhf{CK}

\tlhw{ghu\'}
\tlhd{n}{Situation.}
\tlhf{TKD}