\chapter{\tlhe{H}}

\tlhw{Hab}
\tlhd{v}{Be smooth.}
\tlhf{TKD}[TKW]

\tlhw{HablI\'}
\tlhd{n}{Data transceiving device.}
\tlhf{TKD}

\tlhw{Hach}
\tlhd{v}{Be developed (e.g, civilization).}
\tlhf{TKD}

\tlhw{HaD}
\tlhd{v}{Study.}
\tlhc
\tlhe{yIHaDqu\'}{Study him well!}
\tlhe{tlhIngan Hol DaHaDrup\'a\'}{Are you ready to study Klingon?}
\tlhe{yIHaDqa\'}{Start studying again}
\tlhca{jIHaD}{I study.}[CK]
\tlhf{TKD}[CK]

\tlhw{Hagh}
\tlhd{v}{Laugh.}
\tlhf{TKD}[TKW]

\tlhw{HaH}
\tlhd{v}{Marinade.}[Soak, drench.]
\tlhf{KGT}

\tlhw{Haj}
\tlhd{v}{Dread.}
\tlhf{TKD}

\tlhw{HajDob}
\tlhd{n}{Leg (served as food).}
\tlhf{KGT}

\tlhw{Hal}
\tlhd{n}{Source.}
\tlhf{TKD}

\tlhw{HanDogh}
\tlhd{n}{Nacelle.}
\tlhf{TKD}

\tlhw{Hap}
\tlhd{n}{Matter.}
\tlhf{TKD}

\tlhw{Haq}
\tlhd{n}{Surgery, perform surgery (on).}
\tlhd{v}{Intervene in (a situation). (slang)}
\tlhf{TKD}[KGT]

\tlhw{Haqtaj}
\tlhd{n}{Scalpel.}
\tlhf{TKD}

\tlhw{HaqwI\'}
\tlhd{n}{Surgeon.}
\tlhf{TKD}

\tlhw{HaQchor}
\tlhd{n}{Saccharin.}
\tlhf{TKD}

\tlhw{Har}
\tlhd{n}{Believe.}
\tlhf{TKD}

\tlhw{Hargh}
\tlhd{v}{Fight, battle. (major confrontation)}
\tlhcf{lul}[Qor][tlhaS][vay]
\tlhf{KGT}

\tlhw{HaSta}
\tlhd{n}{Visual display.}
\tlhf{TKD}

\tlhw{HaSta jIH}
\tlhd{n}{Television.}
\tlhf{Eurotalk}

\tlhw{Hat}
\tlhd{v}{Be illegal.}
\tlhf{TKD}

\tlhw{Hat}
\tlhd{n}{Temperature.}
\tlhf{TKD}

\tlhw{Hatlh}
\tlhd{n}{Country, countryside.}
\tlhf{TKD}

\tlhw{Haw\'}
\tlhd{v}{Flee, get out, run away.}
\tlhf{TKD}[TKW]

\tlhw{Hay}
\tlhd{n}{The consonant "H".}
\tlhf{qep\'a\' wa\'maH javDIch 2009}

\tlhw{Hay\'}
\tlhd{v}{Duel.}
\tlhf{TKD}

\tlhw{Hay\'chu}
\tlhd{v}{Duel to the death.}
\tlhf{KGT 49}

\tlhw{Ha\'}
\tlhd{excl}{Let's go, come on.}
\tlhf{TKD}

\tlhw{Ha\'DIbaH}
\tlhd{n}{Meat, animal.}[Dog, cur, inferior person. (slang)]
\tlhf{TKD}[TKW][KGT]

\tlhw{Ha\'DIbaH ghIH tIr ngogh je}
\tlhd{n}{Burger}
\tlhf{Eurotalk}

\tlhw{Ha\'quj}
\tlhd{n}{Baldric, sash.}
\tlhf{KGT}

\tlhw{He}
\tlhd{n}{Course, route.}
\tlhf{TKD}

\tlhw{Hech}
\tlhd{v}{Intend, mean to.}
\tlhf{TKD}[TKW]

\tlhw{HeD}
\tlhd{v}{Retreat.}
\tlhf{TKD}

\tlhw{HeDon}
\tlhd{n}{Parallel course.}
\tlhf{TKD}

\tlhw{Hegh}
\tlhd{n}{Death.}
\tlhd{v}{Die.}
\tlhf{TKD}[TKW]

\tlhw{Heghbat}
\tlhd{n}{Ritual suicide.}
\tlhf{TKW}[KGT]

\tlhw{Heghba\'}
\tlhd{n}{Ritual suicide.}
\tlhf{Sarek}

\tlhw{HeghmoH}
\tlhd{v}{be fatal.}
\tlhf{TKD}

\tlhw{Heghtay}
\tlhd{n}{Death ritual.}
\tlhf{KGT}

\tlhw{HeH}
\tlhd{n}{Edge.}
\tlhf{TKD}

\tlhw{Hej}
\tlhd{v}{Rob.}
\tlhf{TKD}

\tlhw{HejwI\'}
\tlhd{n}{Robber.}
\tlhf{KGT}

\tlhw{Hem}
\tlhd{v}{Be proud.}
\tlhf{TKD}[TKW]

\tlhw{Heng}
\tlhd{v}{Finger (holes, strings of instruments) to vary sounds.}
\tlhf{KGT}

\tlhw{HeQ}
\tlhd{v}{Comply.}
\tlhf{TKD}

\tlhw{Hergh}
\tlhd{n}{Medicine.}
\tlhf{TKD}

\tlhw{HerghwI\'}
\tlhd{n}{Hypo.}
\tlhf{TKD}

\tlhw{Hergh QaywI\'}
\tlhd{n}{Pneumatic hypo.}
\tlhf{TKD}

\tlhw{HeS}
\tlhd{v}{Commit a crime.}
\tlhf{TKD}

\tlhw{HeS}
\tlhd{n}{Crime.}
\tlhf{TKD}

\tlhw{HeSwI\'}
\tlhd{n}{Criminal.}
\tlhf{TKD}

\tlhw{Hev}
\tlhd{v}{Receive.}
\tlhf{TKD}

\tlhw{Hew}
\tlhd{n}{Statue.}
\tlhf{KGT}

\tlhw{Hew chenmoHwI\'}
\tlhd{v}{Sculptor.}
\tlhf{KGT}

\tlhw{He\'}
\tlhd{v}{Smell, emit odor.}
\tlhf{TKD}

\tlhw{He\'So\'}
\tlhd{v}{Stink.}
\tlhf{TKD}

\tlhw{HIch}
\tlhd{n}{Handgun.}
\tlhf{TKD}

\tlhw{HIchDal}
\tlhd{n}{airlock.}
\tlhf{TKD}

\tlhw{HIDjolev}
\tlhd{n}{menu.}
\tlhf{TKD}

\tlhw{HIgh}
\tlhd{v}{Fight dirty.}
\tlhf{TKD}

\tlhw{HIj}
\tlhd{v}{Deliver, transport goods.}
\tlhf{TKD}

\tlhw{HIja\'}
\tlhd{excl}{Yes, true (answer to yes/no question).}
\tlhf{TKD}

\tlhw{HIjmeH chaw\'}
\tlhd{n}{Stamp.}
\tlhf{Eurotalk}

\tlhw{HIjwI\'}
\tlhd{n}{Delivery person.}
\tlhf{TKW}[KGT]

\tlhw{HIp}
\tlhd{n}{Uniform.}
\tlhf{TKD}

\tlhw{HIq}
\tlhd{n}{Ale, beer, liquor, wine.}
\tlhf{KGT}

\tlhw{HISlaH}
\tlhd{excl}{Yes, true (answer to yes/no question).}
\tlhf{TKD}

\tlhw{HIv}
\tlhd{v}{Attack.}
\tlhf{TKD}[TKW]

\tlhw{HIvDuj}
\tlhd{n}{Attack fighter (vessel).}
\tlhf{KGT}

\tlhw{HIvje\'}
\tlhd{n}{Glass (tumbler), mug.}
\tlhf{TKD}

\tlhw{HIvje\' bom}
\tlhd{n}{Drinking song.}
\tlhf{KGT}

\tlhw{HIvneS}
\tlhd{v}{Honor attack (suicide mission.)}
\tlhf{KGT}

\tlhw{HI\'}
\tlhd{n}{Dictator.}
\tlhf{TKD}

\tlhw{HI\'tuy}
\tlhd{n}{Dictatorship.}
\tlhf{TKD}

\tlhw{Hob}
\tlhd{v}{Yawn.}
\tlhf{TKD}

\tlhw{Hoch}
\tlhd{n}{Everyone, all, everything, each, every.}
\tlhco{If noun following Hoch is explicitly plural, it means "all". If it is singular, it means "each".}{KGT}
\tlhco{First in noun-noun pair in TKW.}{KGT}
\tlhf{TKD}[TKW][HolQeD v5n2p11]

\tlhw{HochDIch}
\tlhd{num}{Allth, final, last.}
\tlhf{KGT 177}

\tlhw{HochHom}
\tlhd{n}{Most, greater part.}
\tlhf{KGT}

\tlhw{Hochlogh}
\tlhd{adv}{All times, always (emphasis).}
\tlhf{KGT 178}

\tlhw{HoD}
\tlhd{n}{Captain.}
\tlhf{TKD}

\tlhw{Hogh}
\tlhd{n}{Week (Klingon).}
\tlhf{TKD}

\tlhw{HoH}
\tlhd{n}{Killing.}
\tlhd{v}{Kill.}
\tlhf{TKD}[TKW]

\tlhw{HoH\'egh}
\tlhd{v}{commit suicide.}
\tlhf{TKD}

\tlhw{Hoj}
\tlhd{v}{Be cautious.}
\tlhf{TKD}

\tlhw{Hol}
\tlhd{n}{Language.}
\tlhf{TKD}

\tlhw{Hol ghoQ}
\tlhd{n}{Fresh language (slang).}

\tlhco{Mostly used by younger Klingons.}{KGT 35}
\tlhf{KGT}

\tlhw{HolQeD}
\tlhd{n}{Linguistics.}[The Klingon Language Institute's journal.]
\tlhf{KGT}[HolQeD]

\tlhw{Hom}
\tlhd{n}{Bone.}[Weakling, runt, scrawny one, skinny one. (slang)]
\tlhd{v}{Use the second toe.}
\tlhf{TKD, KGT}[HolQeD v10n2p10]

\tlhw{HomwI\'}
\tlhd{n}{Second toe.}
\tlhf{HolQeD v10n2p10}

\tlhw{Hon}
\tlhd{v}{Doubt.}
\tlhf{TKD}

\tlhw{Hong}
\tlhd{n}{Impulse power.}
\tlhf{TKD}

\tlhw{Hong boq chuyDaH}
\tlhd{n}{Impulse fusion thrusters.}
\tlhf{BoP}

\tlhw{Hongghor}
\tlhd{n}{Impulse drive.}
\tlhf{BoP}

\tlhw{Hop}
\tlhd{v}{Be remote, far.}
\tlhf{TKD}

\tlhw{Hoq}
\tlhd{n}{Expedition.}
\tlhf{TKD}

\tlhw{Hoqra\'}
\tlhd{n}{Tricorder.}
\tlhf{TKD}

\tlhw{HoQ}
\tlhd{v}{Be honored falsely, be falsely honorable}
\tlhco{Deception is implied.}{KLI.org New Words list}
\tlhf{HolQeD v12n3p8}

\tlhw{HoS}
\tlhd{n}{Strength, energy, power.}
\tlhd{v}{Be strong.}
\tlhf{TKD}[TKW]

\tlhw{HoS choHwI\'}
\tlhd{n}{Transtator.}
\tlhf{HolQeD v1n3p9}

\tlhw{HoSchem}
\tlhd{n}{Energy field.}
\tlhf{TKD}

\tlhw{HoSDo\'}
\tlhd{n}{Energy beings.}
\tlhf{TKD}

\tlhw{HoSghaj}
\tlhd{v}{Be powerful.}
\tlhf{TKD}

\tlhw{Hot}
\tlhd{v}{Touch, feel.}
\tlhf{TKD}

\tlhw{Hotlh}
\tlhd{v}{Project, put on (screen).}[Scan.]
\tlhf{TKD}

\tlhw{HotlhwI\'}
\tlhd{n}{Scanner.}
\tlhf{TKD}

\tlhw{Hov}
\tlhd{n}{Star.}
\tlhf{TKD}

\tlhw{Hovpov}
\tlhd{n}{Stardate.}
\tlhf{KGT}

\tlhw{Hovtay\'}
\tlhd{n}{Star system.}
\tlhf{TKD}

\tlhw{Hoy\'}
\tlhd{v}{Congratulate.}
\tlhf{TKD}

\tlhw{Ho\'}
\tlhd{n}{Tooth.}[Idol. (slang)]
\tlhd{v}{Admire.}
\tlhf{TKD}[KGT]

\tlhw{Ho\' teywI\'}
\tlhd{n}{Toothbrush.}
\tlhf{Eurotalk}

\tlhw{Ho\' \'etlh}
\tlhd{n}{Serrated blade (e.g, /kut'luch/).}
\tlhf{KGT 61}

\tlhw{Ho\'Du\' Say\'moHwI\' tlhagh}
\tlhd{n}{Toothpaste.}
\tlhf{Eurotalk}

\tlhw{Ho\'\'oy\'}
\tlhd{n}{Toothache.}
\tlhf{TKD}

\tlhw{Hu}
\tlhd{n}{Zoo.}
\tlhf{TKD}

\tlhw{Hub}
\tlhd{n}{Defense.}
\tlhd{v}{Defend.}
\tlhf{TKD}

\tlhw{HubneS}
\tlhd{n}{Honor defend (suicide mission).}
\tlhf{KGT 49}

\tlhw{Huch}
\tlhd{n}{Money.}
\tlhf{TKD}

\tlhw{HuchQeD}
\tlhd{n}{Economics.}
\tlhco{This is slang with no non-slang counterpart.}{KGT 148}
\tlhf{KGT}

\tlhw{HuD}
\tlhd{n}{Mountain, hill.}
\tlhf{TKD}

\tlhw{HuD beQ yuS}
\tlhd{n}{Flat Mountain district (in the Sakrej region).}
\tlhf{KGT 17}

\tlhw{Hugh}
\tlhd{n}{Throat.}
\tlhf{TKD}

\tlhw{HuH}
\tlhd{n}{Gall, bile, slime.}
\tlhf{TKW, KGT}[PK]

\tlhw{Huj}
\tlhd{v}{Be strange.}
\tlhf{TKD}

\tlhw{Huj}
\tlhd{v}{Charge (up).}
\tlhf{TKD}

\tlhw{Hum}
\tlhd{v}{Be sticky.}
\tlhf{TKD}

\tlhw{Human}
\tlhd{n}{Human.}
\tlhf{TKD}

\tlhw{Hun}
\tlhd{n}{/khrun/, a type of animal.}
\tlhco{I didn't read that section of the opera myself, but I think I heard someone say it was a fast animal. The /khrun/ is described in other resources as a Klingon riding animal. Note: this isn't the way we saw it in the draft of the opera. After being reminded that the name of the animal had been given a tlhIngan Hol rendering in Keith DeCandido's book "A Burning House", Marc Okrand decided to use the already published spelling.}{ghunchu'wI' 'utlh}
\tlhf{qep\'a\' wa\'maH chorghDIch 2011}

\tlhw{Hung}
\tlhd{n}{Security.}
\tlhf{TKD}

\tlhw{Hup}
\tlhd{v}{Punish.}
\tlhf{TKD}

\tlhw{Huq}
\tlhd{v}{Transact.}
\tlhf{TKD}

\tlhw{Hur}
\tlhd{n}{Outside.}
\tlhf{TKD}

\tlhw{HurDagh}
\tlhd{n}{Stringed instrument (general term).}
\tlhf{KGT}

\tlhw{Hurgh}
\tlhd{v}{Be dark.}
\tlhf{TKD}

\tlhw{Hurgh}
\tlhd{n}{Pickle (cucumber).}
\tlhf{TKD}

\tlhw{Hur\'Iq}
\tlhd{n}{Outsider, foreigner. (slang)}
\tlhf{KGT}

\tlhw{Hur\'Iqngan}
\tlhd{n}{Hur'q (person).}
\tlhf{KGT}

\tlhw{HuS name,}
\tlhd{v}{name. 1. Huss.}
\tlhd{v}{Hang.}
\tlhf{TKD}[KGT 197]

\tlhw{Hut}
\tlhd{num}{Nine.}
\tlhf{TKD}

\tlhw{HutDIch}
\tlhd{num}{Ninth.}
\tlhf{TKD}

\tlhw{Hutvagh}
\tlhd{n}{Too many people or things in a place at once.}

\tlhw{Disclosed during a congested drive on Interstate 95.}
\tlhf{HolQeD v11n3p13}

\tlhw{Hutlh}
\tlhd{v}{To lack, to be without, to not have.}
\tlhf{TKW}[KGT]

\tlhw{Huv}
\tlhd{v}{Be clear, not obstructed.}
\tlhf{TKD}

\tlhw{Huy\'}
\tlhd{n}{Eyebrow.}
\tlhf{TKD}

\tlhw{Huy\'Dung}
\tlhd{n}{Forehead. (Regional)}
\tlhf{KGT}

\tlhw{Hu\'}
\tlhd{n}{Days ago.}
\tlhd{v}{Get up.}
\tlhf{TKD}

\tlhw{Hu\'tegh}
\tlhd{excl}{General invective.}
\tlhf{TKD p178}