\chapter{\tlhe{j}}

\tlhw{jab}
\tlhd{v}{Serve (food).}
\tlhf{TKD}

\tlhw{jabbI\'ID}
\tlhd{n}{data transmission.}
\tlhf{TKD}

\tlhw{jabwI\'}
\tlhd{n}{Food server, waiter, waitress.}
\tlhf{KGT}

\tlhw{jach}
\tlhd{v}{Scream, cry out, shout, yell.}
\tlhf{TKD}

\tlhw{jaD}
\tlhd{v}{Throw around, hurl about.}
\tlhf{KGT}

\tlhw{jagh}
\tlhd{n}{Enemy.}
\tlhf{TKD}

\tlhw{jaghla\'}
\tlhd{n}{Enemy commander.}
\tlhf{TKD p64,65}

\tlhw{jaH}
\tlhd{v}{Go, travel.}
\tlhf{TKD}[TKW]

\tlhw{jaj}
\tlhd{n}{Day (from dawn to dawn).}
\tlhf{TKD}

\tlhw{jajlo\'}
\tlhd{n}{Dawn.}
\tlhf{TKD}

\tlhw{jajlo\' Qa\'}
\tlhd{n}{A noisy animal (not a bird) known for making a ruckus at dawn, like a rooster.}
\tlhf{HolQeD v10n4p5}

\tlhw{jajvam}
\tlhd{n}{Today.}
\tlhf{TKW}

\tlhw{jan}
\tlhd{n}{Device.}[Musical instrument.]
\tlhf{TKD}[KGT]

\tlhw{janluq pIqarD HoD}
\tlhd{name}{Captain Jean-Luc Picard.}
\tlhf{HolQeD v5n3p15}[S25]

\tlhw{jang}
\tlhd{v}{Answer, reply.}
\tlhf{TKD}

\tlhw{jaq}
\tlhd{v}{Be bold.}
\tlhf{TKD}

\tlhw{jaqmoH}
\tlhd{v}{Embolden.}
\tlhf{KGT 71}

\tlhw{jar}
\tlhd{n}{Month (Klingon).}
\tlhf{TKD}

\tlhw{jargh}
\tlhd{n}{Forehead. (Regional)}
\tlhf{KGT}

\tlhw{jaS}
\tlhd{adv}{Differently.}
\tlhf{TKD}

\tlhw{jat}
\tlhd{n}{Tongue.}
\tlhd{v}{Speak incoherently, mumble. (slang)}
\tlhf{TKD}[KGT]

\tlhw{jatyIn}
\tlhd{n}{Spiritual possesion.}
\tlhf{KGT}

\tlhw{jatlh}
\tlhd{v}{Say, tell, speak.}
\tlhf{TKD}[TKW]

\tlhw{jav n, num.}
\tlhd{n}{Prisoner. (slang)}[Sixth tone on the nonatonic scale.]
\tlhd{num}{Six.}
\tlhf{TKD}[KGT]

\tlhw{javDIch}
\tlhd{num}{sixth.}
\tlhf{TKD}

\tlhw{jaw}
\tlhd{n}{Lord.}
\tlhd{v}{Chat.}
\tlhf{TKD}

\tlhw{jay}
\tlhd{n}{The consonant "j".}
\tlhf{qep\'a\' wa\'maH javDIch 2009}

\tlhw{jay\'}
\tlhd{adv}{(Invective) Intensely.}
\tlhf{TKD}

\tlhw{ja\'}
\tlhd{v}{Tell, report.}
\tlhf{TKD}

\tlhw{ja\'chuq}
\tlhd{v}{discuss, confer.}
\tlhf{TKD}

\tlhw{ja\'chuq}
\tlhd{n}{succession ritual (ancient).}
\tlhf{TKD}

\tlhw{je}
\tlhd{adv}{Also.}
\tlhd{conj}{Also, and.}
\tlhco{Described in grammar section, it follows the verb it modifies. It joins nouns.}{TKD}
\tlhf{TKD}

\tlhw{jech}
\tlhd{n}{Disguise, costume.}
\tlhd{v}{Disguise.}
\tlhf{TKD}

\tlhw{jeD}
\tlhd{v}{Be dense, be thick, be viscous.}
\tlhf{KGT}

\tlhw{jegh}
\tlhd{v}{Surrender, give up (something).}
\tlhf{TKD}[TKW]

\tlhw{jeH}
\tlhd{v}{Be absentminded.}
\tlhf{TKD}

\tlhw{jej}
\tlhd{v}{Be sharp.}
\tlhf{TKD}

\tlhw{jejHa\'}
\tlhd{v}{Be dull, be blunt (of blades).}
\tlhf{TKW}[KGT]

\tlhw{jen}
\tlhd{v}{Be high.}
\tlhf{TKD}

\tlhw{jengva\'}
\tlhd{n}{Plate (for eating).}
\tlhf{KGT}

\tlhw{jeqqIj}
\tlhd{n}{Club, bludgeon.}
\tlhf{KGT}

\tlhw{jeQ}
\tlhd{v}{Be self-confident.}
\tlhf{TKD}

\tlhw{jer}
\tlhd{v}{Surge (aircraft suddenly moves forward or backward).}
\tlhco{It is unknown if this has meaning outside of the context of aircraft attitude descriptions.}{HolQeD}
\tlhf{HolQeD v11n2p9}

\tlhw{jeS}
\tlhd{v}{Participate.}
\tlhf{TKD}

\tlhw{jev}
\tlhd{v}{Storm.}[Wheeze, breathe noisily.]
\tlhrf{A well-known Klingon myth tells of a man in the ancient city of Quin'lat who dies during a storm because he remained outside the walls of the city in order to show that he was not afraid of the storm and to make the storm respect him. Kahless, who was in the city at the time, remarked \tlhe{qoH vuvbe\' SuS} ("The wind does not respect a fool"), which has become a frequently repeated proverb."}{KGT}[122]
\tlhrf{To breathe noisily, or 'wheeze', is \tlhe{jev}. (This is also the verb \'storm\', though Maltz couldn\'t decide whether this was meaningful or just a coincidence.)}{HolQeD v12n4p8}
\tlhco{At the Farpoint Convention (10/2000), MO told Krankor that \tlhe{jev} "is a weather term" meaning "be rainy, windy, thundering" and is used similarly to {SIS}.}{(q.v.)}
\tlhcf{SuS'a'}[weH][HIv]
\tlhf{TKD}[HolQeD v12n4p8]

\tlhw{jey}
\tlhd{n}{Itinerary.}
\tlhd{v}{Defeat.}
\tlhf{TKD}

\tlhw{jey\'naS}
\tlhd{n}{Double-headed ax, ax with a double-head.}
\tlhco{Entries differs in source. They are combined here.}{\emph{ed}}
\tlhco{The general term for ax is \tlhe{\'obmaQ}. Its handle is its \tlhe{DeS} and its blade is its \tlhe{ghIt}. These words, \tlhe{DeS} and \tlhe{ghIt}, when referring to humanoid anatomy, mean "arm" and "open, flat hand" (as opposed to a fist), respectively, suggesting that at one time the ax was considered an extension of the warrior himself. A double-headed ax is a \tlhe{jey\'naS}, while an ax with an added spike at the end is a \tlhe{\'alngegh}. To wield or swing an ax is \tlhe{Qach}.}{KGT 63}
\tlhc
\tlhe{yuD; jey\'naS rur}{dishonest as a double-headed ax}
\tlhpun{Janus" (the double-faced Roman god).}
\tlhf{KGT}

\tlhw{je\'}
\tlhd{v}{Buy, purchase.}[Feed (someone else).][Enrich (sprit).]
\tlhf{TKD}[TKW]

\tlhw{jIb}
\tlhd{n}{Hair (on head).}
\tlhf{TKD}

\tlhw{jIb Ho\'Du\'}
\tlhd{n}{Comb.}
\tlhf{Eurotalk}

\tlhw{jIb yachwI\'}
\tlhd{n}{Hairbrush.}
\tlhf{Eurotalk}

\tlhw{jIH n,pro. v.}
\tlhd{n}{Viewing screen.}
\tlhd{pro}{I, me.}
\tlhd{v}{Monitor.}
\tlhf{TKD}[TKDa]

\tlhw{jIj}
\tlhd{v}{Cooperate, be cooperative.}

\tlhw{Listed only in the Klingon-English side of TKD while \tlhe{yeq} is listed only in the English-Klingon side. KGT lists it as "be cooperative".}
\tlhf{TKD}[KGT]

\tlhw{jIl}
\tlhd{n}{Neighbor.}
\tlhf{TKD}

\tlhw{jIm}
\tlhd{v}{Heave (aircraft rising or falling without pitching).}
\tlhco{It is unknown if this has meaning outside of the context of aircraft attitude descriptions.}{HolQeD}
\tlhf{HolQeD v11n2p9}

\tlhw{jInaq}
\tlhd{n}{Amulet symbolizing a girl's coming of age.}
\tlhf{KGT}

\tlhw{jInmol}
\tlhd{n}{Project.}
\tlhf{TKD}

\tlhw{jIp}
\tlhd{n}{Penalty.}
\tlhf{TKD}

\tlhw{jIr}
\tlhd{v}{Rotate, twirl.}
\tlhf{KGT}

\tlhw{jIrmoH}
\tlhd{v}{Twirl bat'leth, cause bat'leth to rotate.}
\tlhf{KGT}

\tlhw{jIv}
\tlhd{v}{Be ignorant.}
\tlhf{TKD}

\tlhw{jo}
\tlhd{n}{Resources.}
\tlhf{TKD}

\tlhw{joch}
\tlhd{v}{Be harmful.}
\tlhf{TKD}

\tlhw{joD}
\tlhd{v}{Stoop.}
\tlhf{TKD}

\tlhw{jogh}
\tlhd{n}{Quadrant.}
\tlhf{TKD}

\tlhw{joH}
\tlhd{n}{Lady (female head of house), Lord (male head of house).}

\tlhw{Entries differs in source. They are combined here.}
\tlhf{TKD}[KGT]

\tlhw{joj}
\tlhd{n}{Area between.}
\tlhf{TKD}

\tlhw{jojlu\'}
\tlhd{n}{Consul.}
\tlhf{TKD}

\tlhw{jol}
\tlhd{n}{Transport beam.}
\tlhd{v}{Beam (aboard.)}
\tlhf{TKD}

\tlhw{jolpat}
\tlhd{n}{Transport system.}
\tlhf{BoP}

\tlhw{jolpa\'}
\tlhd{n}{Transport room.}
\tlhf{TKD}

\tlhw{jolvoy\'}
\tlhd{n}{Transporter ionizer unit.}
\tlhf{TKD}

\tlhw{jom}
\tlhd{v}{Install (a device).}
\tlhf{KGT}

\tlhw{jon}
\tlhd{v}{Capture.}
\tlhf{TKD}

\tlhw{jonpIn}
\tlhd{n}{Egineering officer.}
\tlhf{KGT}

\tlhw{jonta\'}
\tlhd{n}{Engine.}
\tlhf{TKD}

\tlhw{jonwI\'}
\tlhd{n}{Engineer.}
\tlhf{TKD}

\tlhw{jop}
\tlhd{v}{Lunge, thrust.}
\tlhf{KGT}

\tlhw{joq}
\tlhd{conj}{Or, and/or (joining nouns).}
\tlhd{v}{Flap, flutter, wave.}
\tlhf{TKD}

\tlhw{joqwI\'}
\tlhd{n}{Flag.}
\tlhf{TKD}

\tlhw{joQ}
\tlhd{n}{Rib.}[Strip of material in a /susdek/ instrument.]
\tlhf{KGT}

\tlhw{jor}
\tlhd{v}{Explode.}
\tlhf{TKD}

\tlhw{jorneb}
\tlhd{n}{Warhead (of a torpedo).}

\tlhw{{jorneb} in KGT, TKD E-K, \tlhe{jornub} in TDK K-E.}
\tlhf{TKD}[KGT]

\tlhw{jorwI\'}
\tlhd{n}{Explosive.}
\tlhf{TKD}

\tlhw{joS}
\tlhd{v}{Gossip.}
\tlhf{TKD}

\tlhw{joS}
\tlhd{n}{Rumor, gossip.}
\tlhf{TKD}

\tlhw{jot}
\tlhd{v}{Be calm.}
\tlhf{TKD}

\tlhw{jotHa\'}
\tlhd{v}{Be uneasy.}
\tlhf{TKD}

\tlhw{jotlh}
\tlhd{v}{Take down.}
\tlhf{TKD}

\tlhw{joy\'}
\tlhd{v}{Torture.}
\tlhf{TKD}

\tlhw{jo\'}
\tlhd{n}{Machinery.}
\tlhf{TKD}

\tlhw{jub}
\tlhd{v}{Be immortal.}
\tlhf{TKD}

\tlhw{jubbe\'}
\tlhd{v}{Be mortal.}
\tlhf{TKD}

\tlhw{juch}
\tlhd{v}{Have a width of.}
\tlhf{MSN}

\tlhw{juH}
\tlhd{n}{Home.}
\tlhf{TKD}

\tlhw{juHolQeDo\'}
\tlhd{n}{Home world.}
\tlhf{TKD}

\tlhw{jul}
\tlhd{n}{Sun.}
\tlhco{This is a general term for a planet's sun. It is not the name of a specific sun.}{ghunchu'wI' 'utlh}
\tlhf{qep\'a\' wa\'maH chorghDIch 2011}

\tlhw{jum}
\tlhd{v}{Be odd.}
\tlhf{TKD}

\tlhw{jun}
\tlhd{v}{Evade, take evasive action.}
\tlhf{TKD}

\tlhw{jungwoq}
\tlhd{n}{China.}
\tlhf{Eurotalk}

\tlhw{jup}
\tlhd{n}{Friend.}
\tlhf{TKD}

\tlhw{juS}
\tlhd{v}{Overtake, pass.}
\tlhf{TKD}

\tlhw{juv}
\tlhd{v}{Measure.}
\tlhf{TKD}