\chapter{\tlhe{Q}}

\tlhw{Qab}
\tlhd{n}{Theragen.}
\tlhf{TKD}

\tlhw{QaD}
\tlhd{v}{Be dry.}
\tlhf{TKD}

\tlhw{QaDmoHwI\' DIr}
\tlhd{n}{Towel.}
\tlhf{Eurotalk}

\tlhw{Qagh}
\tlhd{n}{Error, mistake.}
\tlhf{TKD}

\tlhw{Qagh}
\tlhd{v}{Err, be mistaken, make a mistake.}
\tlhf{TKD}

\tlhw{QaH}
\tlhd{n}{Help.}
\tlhf{TKD}

\tlhw{QaH}
\tlhd{v}{Help, aid.}
\tlhf{TKD}

\tlhw{Qam}
\tlhd{v}{Stand.}
\tlhf{TKD}

\tlhw{Qan}
\tlhd{v}{Protect.}
\tlhf{TKD}

\tlhw{Qang}
\tlhd{n}{Chancellor.}
\tlhf{TKD}

\tlhw{Qap}
\tlhd{v}{Work, function, succeed.}
\tlhf{TKD}

\tlhw{Qapla\'}
\tlhd{n}{Success.}
\tlhf{TKD}

\tlhw{QaQ}
\tlhd{v}{Be good.}
\tlhf{TKD}

\tlhw{Qargh}
\tlhd{n}{Fissure.}
\tlhcf{Seq}[Qom][ngech]['och][DIS]
\tlhf{TKD}

\tlhw{QaS}
\tlhd{n}{Troops.}
\tlhf{TKD}

\tlhw{Qat}
\tlhd{v}{Be popular.}
\tlhf{TKD}

\tlhw{Qatlh}
\tlhd{v}{Be difficult.}
\tlhf{TKD}

\tlhw{Qav}
\tlhd{v}{Be final, last.}
\tlhf{TKD}

\tlhw{Qaw\'}
\tlhd{n}{Destroy.}
\tlhf{TKD}

\tlhw{Qay}
\tlhd{n}{The consonant "Q".}
\tlhf{qep\'a\' wa\'maH javDIch 2009}

\tlhw{Qay}
\tlhd{v}{Transfer.}
\tlhf{TKD}

\tlhw{Qay\'}
\tlhd{v}{Blow one's top.}
\tlhf{TKD}

\tlhw{Qa\'}
\tlhd{n}{Type of animal.}
\tlhf{TKD}

\tlhw{Qeb}
\tlhd{n}{Ring (for finger.)}[Ponytail holder. (slang)]
\tlhco{The usual word for the popular hairstyle called a "ponytail" in Federation Standard is \tlhe{DaQ}, but the word \tlhe{naQ} (literally, "cane, staff") is frequently heard instead. This may have originated with the Krotmag dialect\'s pronunciation of \tlhe{DaQ}, which sounds like \tlhe{naQ} to most Klingons... Incidentally, a slang expression for "ponytail holder", usually \tlhe{choljaH}, is \tlhe{Qeb}, the regular word for "ring" (for the finger).}{KGT}{154-55}
\tlhcf{choljaH}[jIb]
\tlhf{TKD}[KTG]

\tlhw{QeD}
\tlhd{n}{Science.}
\tlhf{TKD}

\tlhw{QeDpIn}
\tlhd{n}{Science officer.}
\tlhf{TKD}

\tlhw{QeH}
\tlhd{n}{Anger.}
\tlhf{TKD}

\tlhw{QeH}
\tlhd{v}{Be angry, mad.}
\tlhf{TKD}

\tlhw{Qel}
\tlhd{n}{Doctor, physician.}
\tlhf{TKD}

\tlhw{Qey}
\tlhd{v}{Be tight.}
\tlhf{TKD}

\tlhw{QeyHa\'}
\tlhd{v}{Be loose.}
\tlhf{TKD}

\tlhw{QeyHa\'moH}
\tlhd{v}{Loosen.}
\tlhf{TKD}

\tlhw{QeymoH}
\tlhd{v}{Tighten.}
\tlhf{TKD}

\tlhw{Qe\'}
\tlhd{n}{Restaurant.}
\tlhf{TKD}

\tlhw{QIghpej}
\tlhd{n}{Klingon agonizer.}
\tlhf{TKD}

\tlhw{QIb}
\tlhd{n}{Shadow.}
\tlhf{TKD}

\tlhw{QIch}
\tlhd{n}{Speech (vocal sounds).}
\tlhf{TKD}

\tlhw{QID}
\tlhd{v}{Wound.}
\tlhf{TKD}

\tlhw{QIH}
\tlhd{v}{Damage, cause damage.}
\tlhf{TKD}

\tlhw{QIH}
\tlhd{n}{Damage, destruction.}
\tlhf{TKD}

\tlhw{QIj}
\tlhd{v}{Explain.}
\tlhf{TKD}

\tlhw{QImmey}
\tlhd{pl n}{Eggs.}
\tlhf{Eurotalk}

\tlhw{Qin \'echletHom}
\tlhd{n}{Postcard.}
\tlhf{Eurotalk}

\tlhw{QIncha\'}
\tlhd{n}{/krincha/, a type of animal.}
\tlhco{The English accompanying the opera said "Krincha, short and fierce". I found references to a "Krencha" in several places, including the KLI's Extended Corpus Project and the Memory Beta wiki. It is described as a large, quick lizard native to Qo'noS, variously said to have either four or six legs. I suspect that the opera's author intended to invoke that animal but misspelled its name.}{ghunchu'wI' 'utlh}
\tlhf{qep\'a\' wa\'maH chorghDIch 2011}

\tlhw{QIp}
\tlhd{v}{Be stupid.}
\tlhf{TKD}

\tlhw{QIt}
\tlhd{adv}{Slowly.}
\tlhf{TKD}

\tlhw{QIv}
\tlhd{v}{Be inferior.}
\tlhf{TKD}

\tlhw{QI\'}
\tlhd{n}{Military.}
\tlhf{TKD}

\tlhw{QI\'tomer}
\tlhd{n}{Khitomer.}
\tlhf{TKD}

\tlhw{QI\'tu\'}
\tlhd{n}{Paradise.}
\tlhf{TKD}

\tlhw{Qob}
\tlhd{v}{Be dangerous.}
\tlhf{TKD}

\tlhw{Qob}
\tlhd{n}{Danger.}
\tlhf{TKD}

\tlhw{Qoch}
\tlhd{v}{Disagree.}
\tlhf{TKD}

\tlhw{Qochbe\'}
\tlhd{v}{Agree.}
\tlhf{TKD}

\tlhw{QoD}
\tlhd{v}{Maneuver (engines).}
\tlhf{TKD}

\tlhw{Qogh}
\tlhd{n}{Type of animal.}
\tlhf{TKD}

\tlhw{Qoj}
\tlhd{v}{Make war.}
\tlhf{TKD}

\tlhw{Qol}
\tlhd{v}{Beam away.}
\tlhf{TKD}

\tlhw{Qom}
\tlhd{v}{Experience an earthquake/tremor.}
\tlhf{TKD}

\tlhw{QonoS}
\tlhd{n}{Journal, log.}
\tlhf{TKD}

\tlhw{Qong}
\tlhd{v}{Sleep.}
\tlhf{TKD}

\tlhw{Qop}
\tlhd{v}{Be worn out.}
\tlhf{TKD}

\tlhw{QopmoH}
\tlhd{v}{Wear out.}
\tlhf{TKD}

\tlhw{Qor}
\tlhd{v}{Very minor fight.}
\tlhcf{lul}[Hargh][tlhaS][vay]
\tlhf{KGT}

\tlhw{Qorgh}
\tlhd{v}{Take care of, care for.}
\tlhf{TKD}

\tlhw{QoS}
\tlhd{v}{Be sorry.}
\tlhf{TKD}

\tlhw{Qot}
\tlhd{v}{Lie, recline.}
\tlhf{TKD}

\tlhw{Qotlh}
\tlhd{v}{Disable.}
\tlhf{TKD}

\tlhw{Qoy}
\tlhd{v}{Hear.}
\tlhf{TKD}

\tlhw{Qo\'}
\tlhd{excl}{No, I won't, I refuse, I disagree.}
\tlhf{TKD}

\tlhw{Qo\'noS}
\tlhd{n}{Kronos.}
\tlhf{TKD}

\tlhw{Qub}
\tlhd{v}{Think.}
\tlhf{TKD}

\tlhw{Quch}
\tlhd{v}{Be happy.}
\tlhf{TKD}

\tlhw{Quch .n Forehead.}
\tlhf{TKD}

\tlhw{QuchHa\'}
\tlhd{v}{Be unhappy.}
\tlhf{TKD}

\tlhw{QuD}
\tlhd{n}{Insurrection.}
\tlhf{TKD}

\tlhw{Qugh}
\tlhd{n}{Disaster.}
\tlhf{TKD}

\tlhw{Quj}
\tlhd{n}{Game.}
\tlhd{v}{Play a game.}
\tlhf{TKD}

\tlhw{QujmeH moQ}
\tlhd{n}{Ball.}
\tlhf{Eurotalk}

\tlhw{Qul}
\tlhd{v}{Research.}
\tlhf{TKD}

\tlhw{Qum}
\tlhd{v}{Communicate.}
\tlhf{TKD}

\tlhw{QumpIn}
\tlhd{n}{Communications officer.}
\tlhf{TKD}

\tlhw{QumwI\'}
\tlhd{n}{Communicator, communications device.}
\tlhf{TKD}

\tlhw{Qup}
\tlhd{v}{Be young.}
\tlhf{TKD}

\tlhw{QuQ}
\tlhd{n}{Engine.}
\tlhf{TKD}

\tlhw{QuS}
\tlhd{n}{Conspiracy.}
\tlhf{TKD}

\tlhw{QuS}
\tlhd{v}{Conspire.}
\tlhf{TKD}

\tlhw{Qut}
\tlhd{v}{Be vulgar.}
\tlhf{TKD}

\tlhw{Quv}
\tlhd{n}{Coordinates.}
\tlhf{TKD}

\tlhw{Qu\'}
\tlhd{n}{Duty, quest, mission, task, chore.}
\tlhf{TKD}