\chapter{\tlhe{l}}

\tlhw{lab}
\tlhd{v}{Transmit data (away from a place).}
\tlhf{TKD}

\tlhw{lach}
\tlhd{v}{Exaggerate.}
\tlhf{TKD}

\tlhw{laD}
\tlhd{v}{Read.}
\tlhf{TKD}

\tlhw{laH}
\tlhd{n}{Ability.}
\tlhf{TKD}

\tlhw{lagh}
\tlhd{n}{Ensign.}
\tlhd{v}{Take apart, disassemble.}
\tlhf{TKD}

\tlhw{laj}
\tlhd{n}{Acceptance.}
\tlhd{v}{Accept.}
\tlhf{TKD}

\tlhw{lajQo\'}
\tlhd{v}{Reject.}
\tlhf{KGT}

\tlhw{lalDan}
\tlhd{n}{Religion.}
\tlhf{TKD}

\tlhw{lam}
\tlhd{n}{Dirt.}
\tlhd{v}{Be dirty.}
\tlhf{TKD}

\tlhw{lan}
\tlhd{v}{Place.}
\tlhf{TKD}

\tlhw{lang}
\tlhd{v}{Be thin.}
\tlhf{TKD}

\tlhw{laq}
\tlhd{v}{Flap (move like a bird's wings, not like a flag).}
\tlhc
\tlhca{laq tel. tel laqmoH bo\'Degh.}{\emph{(No translation given. -ed)}}
\tlhf{HolQeD v10n4p5}

\tlhw{laQ}
\tlhd{v}{Fire, energize (e.g, thrusters).}
\tlhf{TKD}

\tlhw{largh}
\tlhd{v}{Smell, sense odors.}
\tlhf{TKD}

\tlhw{laSvargh}
\tlhd{n}{Factory.}
\tlhf{TKD}

\tlhw{lat}
\tlhd{n}{Shrine.}
\tlhf{KGT}

\tlhw{latlh}
\tlhd{n}{Additional one, other one.}
\tlhf{TKD}

\tlhw{lav}
\tlhd{n}{Shrub.}
\tlhf{KGT}

\tlhw{law\'}
\tlhd{v}{Many, be many, numerous, be numerous.}
\tlhf{TKD}

\tlhw{lay}
\tlhd{n}{The consonant "l".}
\tlhf{qep\'a\' wa\'maH javDIch 2009}

\tlhw{lay\'}
\tlhd{v}{Promise.}
\tlhf{TKD}

\tlhw{lay\'Ha\'}
\tlhd{v}{Break one's word.}
\tlhf{KGT}

\tlhw{la\'}
\tlhd{n}{Commander.}
\tlhf{TKD}

\tlhw{la\'quv}
\tlhd{n}{Supreme Commander.}
\tlhf{TKD}

\tlhw{la\'\'a\'}
\tlhd{n}{Commandant.}
\tlhf{TKD}

\tlhw{legh}
\tlhd{v}{See.}
\tlhf{TKD}

\tlhw{leH}
\tlhd{n}{Maintenance.}
\tlhd{v}{Maintain.}
\tlhf{TKD}

\tlhw{lel}
\tlhd{v}{Get out, take out.}
\tlhf{TKD}

\tlhw{lem}
\tlhd{n}{Hoof.}
\tlhf{KGT}

\tlhw{len}
\tlhd{n}{Recess, break.}
\tlhf{TKD}

\tlhw{leng}
\tlhd{n}{Trip, voyage.}
\tlhd{v}{Roam, travel, rove.}
\tlhf{TKD}

\tlhw{lengwI\'}
\tlhd{n}{Rover (grammatical term).}
\tlhf{TKD}

\tlhw{leQ}
\tlhd{n}{Switch.}
\tlhf{TKD}

\tlhw{leS}
\tlhd{n}{Days from now.}
\tlhd{v}{Rest, relax.}
\tlhf{TKD}

\tlhw{leSpoH}
\tlhd{n}{Shore leave.}
\tlhf{TKD}

\tlhw{leSpal}
\tlhd{n}{A type of stringed instrument.}
\tlhf{KGT}

\tlhw{leSSov}
\tlhd{n}{Foresight.}
\tlhf{TKD}

\tlhw{let}
\tlhd{v}{Be hard (like a rock).}
\tlhf{TKD}

\tlhw{letlh}
\tlhd{n}{Stairs, stairway.}[Stairway leading to door of a ship. (Regional)]
\tlhf{KGT}

\tlhw{lev}
\tlhd{n}{Roof of the mouth, or hard palate.}
\tlhd{v}{Move /bat'leth/ from vertical to horizontal orientation.}
\tlhf{KGT}[HolQeD v14n2-3p21.]

\tlhw{ley\'}
\tlhd{n}{Nose. (Regional)}
\tlhf{KGT}

\tlhw{le\'}
\tlhd{v}{Be special, exceptional.}
\tlhf{TKD}

\tlhw{le\'be\'}
\tlhd{v}{Be unexceptional, nonspecific.}
\tlhf{KGT}

\tlhw{le\'mIS}
\tlhd{n}{Blockade.}
\tlhf{Klingon Monopoly}

\tlhw{le\'yo\'}
\tlhd{n}{Pride.}
\tlhco{Just to be clear, this describes an emotion, as opposed to a tribe of lions.}{ghunchu'wI' 'utlh}
\tlhf{qep\'a\' wa\'maH chorghDIch 2011}

\tlhw{lIch}
\tlhd{v}{Pour (into/onto anything).}
\tlhf{KGT}

\tlhw{lIgh}
\tlhd{v}{Ride.}
\tlhf{TKD}

\tlhw{lIghon}
\tlhd{n}{Ligon.}
\tlhf{KGT}

\tlhw{lIghon DuQwI\' pogh}
\tlhd{n}{Glavin, Ligonian spike glove.}
\tlhf{KGT}

\tlhw{lIH}
\tlhd{v}{Introduce.}[Begin a song.]
\tlhf{HolQeD v12n2p9}

\tlhw{lIj}
\tlhd{v}{Forget.}
\tlhf{TKD}

\tlhw{lIm}
\tlhd{v}{Panic .}
\tlhf{TKD}

\tlhw{lInDab}
\tlhd{n}{Espionage.}
\tlhf{TKD}

\tlhw{lIngwI\'}
\tlhd{n}{Generator.}
\tlhf{TKD}

\tlhw{lIng}
\tlhd{v}{Produce.}
\tlhf{TKD}

\tlhw{lIngta\'}
\tlhd{n}{/lingta/, a type of animal.}
\tlhf{KGT}

\tlhw{lIq}
\tlhd{v}{Round up.}
\tlhf{TKD}

\tlhw{lIr}
\tlhd{n}{A nocturnal bird.}
\tlhf{HolQeD v10n4p4}

\tlhw{lIS}
\tlhd{v}{Adjust.}
\tlhf{TKD}

\tlhw{lIt}
\tlhd{v}{Get on (to).}
\tlhf{KGT}

\tlhw{lItHa\'}
\tlhd{v}{Get off (of).}
\tlhf{KGT}

\tlhw{lIw}
\tlhd{n}{Substitute, stand-in, temporary surrogate.}
\tlhf{KGT}

\tlhw{lIy}
\tlhd{n}{Comet.}
\tlhf{TKD}

\tlhw{lI\'}
\tlhd{v}{Be useful.}[Transmit data (to a place).]
\tlhf{TKD}

\tlhw{lob}
\tlhd{v}{Obey.}
\tlhf{TKD}

\tlhw{lobHa\'}
\tlhd{v}{Disobey.}
\tlhf{TKD}

\tlhw{loch}
\tlhd{n}{Mustache.}
\tlhf{TKD}

\tlhw{loD}
\tlhd{n}{Male, man.}
\tlhf{TKD}

\tlhw{loDHom}
\tlhd{n}{Boy.}
\tlhf{TKD}

\tlhw{loDnal}
\tlhd{n}{Husband.}
\tlhf{TKD}

\tlhw{loDnI\'}
\tlhd{n}{Brother.}
\tlhf{TKD}

\tlhw{logh}
\tlhd{n}{Space, outer space.}
\tlhf{TKD}[TKW]

\tlhw{loghqam}
\tlhd{n}{Unit of measure, 1.25 light years.}
\tlhf{BoP}

\tlhw{loH}
\tlhd{n}{Administration.}
\tlhd{v}{Administer.}
\tlhf{TKD}

\tlhw{loj}
\tlhd{v}{Be all gone.}
\tlhf{TKD}

\tlhw{lojmIt}
\tlhd{n}{Door, gate.}
\tlhf{TKD}

\tlhw{lol}
\tlhd{n}{A specific position in the martial arts form Mok'bara.}
\tlhd{v}{Be in an attitude (aircraft).}[Be in a stance, be in a pose (people animals or martial arts).]
\tlhco{It is unknown if this is used outside the context of aircraft attitude.}{HolQeD}
\tlhf{HolQeD v11n2p8}

\tlhw{lolchu\'}
\tlhd{v}{Be in a correct attitude (aircraft).}
\tlhco{It is unknown if this is used outside the context of aircraft attitude.}{HolQeD}
\tlhf{HolQeD v11n2p8}

\tlhw{lolchu\'taH}
\tlhd{v}{Maintain a correct attitude (aircraft).}
\tlhco{It is unknown if this is used outside the context of aircraft attitude.}{HolQeD}
\tlhf{HolQeD v11n2p8}

\tlhw{lolmoH}
\tlhd{v}{Maneuver (an aircraft) to be in an attitude.}
\tlhco{It is unknown if this is used outside the context of aircraft attitude.}{HolQeD}
\tlhc
\tlhca{qughmeH Duj vIlolmoH}{\emph{(No translation given. -ed)}}
\tlhf{HolQeD v11n2p8}

\tlhw{loltaH}
\tlhd{v}{Maintain an attitude (aircraft).}
\tlhco{It is unknown if this is used outside the context of aircraft attitude.}{HolQeD}
\tlhf{HolQeD v11n2p8}

\tlhw{lolSeHcha}
\tlhd{n}{Attitude-control thrusters.}
\tlhf{TKD}

\tlhw{lom}
\tlhd{n}{Corpse.}
\tlhf{TKD}

\tlhw{lon}
\tlhd{v}{Abandon.}
\tlhf{TKD}

\tlhw{lop}
\tlhd{n}{Celebration.}
\tlhd{v}{Celebrate.}[Observe, celebrate (a ritual).]
\tlhf{TKD}[KGT]

\tlhw{lopno\'}
\tlhd{n}{Party celebration.}
\tlhf{KGT}

\tlhw{loQ}
\tlhd{adv}{Slightly, a little bit.}
\tlhf{TKD}

\tlhw{lor}
\tlhd{n}{Cousin (mother's brother's child or father's sister's child), niece or nephew (man's sister's child or woman's brother's child).}
\tlhf{HolQeD v9n3p8}

\tlhw{lorbe\'}
\tlhd{n}{Female cousin (mother's brother's daughter or father's sister's daughter), niece (man's sister's daughter or woman's brother's daughter).}
\tlhf{HolQeD v9n3p8}

\tlhw{lorloD}
\tlhd{n}{Male cousin (mother's brother's son or father's sister's son), nephew (man's sister's son or woman's brother's son).}
\tlhf{HolQeD v9n3p8}

\tlhw{loS n, num, v.}
\tlhd{n}{Fourth tone of nonatonic musical scale.}
\tlhd{num}{}[Four.]
\tlhd{v}{-3. Wait (for).}
\tlhco{Older Klingon music was based on a nonatonic scale--that is, one made up of nine tones. Each tone has a specific name, comparable to the "do, re, mi" system used in describing music on Earth. The nine tone names are (the first and ninth, as with Earth's "do," being the same): \tlhe{yu} \tlhe{bIm}, \tlhe{\'egh}, \tlhe{loS}, \tlhe{vagh}, \tlhe{jav}, \tlhe{Soch}, \tlhe{chorgh}, \tlhe{yu}. While the first three (and ninth) of these words apparently are used only for singing the scale, the remaining five are also numerals: \tlhe{loS}, "four"; \tlhe{vagh}, "five"; \tlhe{jav}, "six"; \tlhe{Soch}, "seven"; \tlhe{chorgh}, "eight." It is possible that, at some time in the past, the numerals were "borrowed" into the lexicon of music in order to sing the scale but, for some reason, the first three (presumably \tlhe{wa\'}, \tlhe{cha\'}, \tlhe{wej} [``one, two, three'']) were either changed or never used. It is far more likely, however, that the borrowing went in the other direction. As is well documented, the Klingon counting system was originally a ternary system (one based on three, with numbers higher than three formed from the words for "one," "two," and "three"). Later, owing to outside influences, it changed to a decimal system (based on ten). The independent words for the numbers three through nine were not originally a part of the Klingon counting system, but they had to come from somewhere. The musical scale is the likely source. The word for the fourth musical tone, \tlhe{loS}, began to be used for the number four, and so on through the eighth tone, \tlhe{chorgh}. (The origins of the words \tlhe{Hut} ``nine'' and the suffix \tlhe{-maH}, used in the words for "ten," "twenty," "thirty," and so on, are obscure.)}{KGT 72}
\tlhf{TKD}

\tlhw{loSDIch}
\tlhd{num}{Fourth.}
\tlhf{TKD}

\tlhw{loSpev}
\tlhd{n}{Quadrotriticale.}
\tlhf{TKD}

\tlhw{lot}
\tlhd{n}{Catastrophe.}
\tlhf{TKD}

\tlhw{lotlh}
\tlhd{v}{Rebel.}
\tlhf{TKD}

\tlhw{lotlhmoq}
\tlhd{n}{A bird that swoops into the water in order to catch food, but cannot swim.}
\tlhf{HolQeD v10n4p4}

\tlhw{lotlhwI\'}
\tlhd{n}{Rebel.}
\tlhf{TKD}

\tlhw{loy}
\tlhd{v}{Guess.}
\tlhf{TKD}

\tlhw{lo\'}
\tlhd{n}{Use.}
\tlhd{v}{Use.}
\tlhf{TKD}

\tlhw{lo\' law\'}
\tlhd{n}{Utility (precedes noun it modifies).}
\tlhf{BoP}

\tlhw{lo\'laH}
\tlhd{v}{Be valuable.}
\tlhf{TKD}

\tlhw{lo\'laHbe\'}
\tlhd{v}{Be worthless.}
\tlhf{TKD}

\tlhw{lo\'laHbe\'ghach}
\tlhd{v}{Worthlessness.}
\tlhf{TKD}

\tlhw{lo\'laHghach}
\tlhd{v}{Value.}
\tlhf{TKD}

\tlhw{lu}
\tlhd{v}{Fall (suffer loss of status).}
\tlhf{KGT}

\tlhw{luch}
\tlhd{v}{Equipment, gear.}
\tlhf{TKD}

\tlhw{lugh}
\tlhd{v}{Be right, correct.}
\tlhf{TKD}

\tlhw{luH}
\tlhd{n}{Intestine.}
\tlhd{v}{Yank.}[Cause (someone) to confess or reveal a secret, exact a confession. (slang)]
\tlhco{Entries differs in source. They are combined here.}{TKD}
\tlhf{TKD}[KGT]

\tlhw{luHwI' tIH}
\tlhd{n}{Tractor beam.}
\tlhco{An linear graviton beam used by starships and space stations to control the movement of external objects.}{M.Ok.}
\tlhf{qepHom 2010, email to Lieven, 15 November 2010}

\tlhw{luj}
\tlhd{v}{Fail, lose (a competition), lose (not win).}
\tlhf{TKD, TKW}[KGT]

\tlhw{lul}
\tlhd{v}{Fight, battle. (Relatively major fight.)}
\tlhcf{Hargh}[Qor][tlhaS][vay]
\tlhf{KGT}

\tlhw{lulIgh}
\tlhd{n}{Refuge.}
\tlhf{TKD}

\tlhw{lum}
\tlhd{v}{Postpone, procrastinate.}
\tlhf{TKD}

\tlhw{lung}
\tlhd{v}{/long/, type of animal (lizardlike).}
\tlhf{KGT}

\tlhw{lup}
\tlhd{n}{Second (of time).}
\tlhd{v}{Transport.}
\tlhf{TKD}

\tlhw{lupwI\'}
\tlhd{n}{Jitney, bus.}
\tlhf{KGT}

\tlhw{lupDujHom}
\tlhd{n}{Shuttlecraft.}
\tlhf{TKD}

\tlhw{luq}
\tlhd{excl}{Yes, okay, I will.}
\tlhf{TKD}

\tlhw{lur}
\tlhd{n}{Pupil (of eye).}
\tlhf{TKD}

\tlhw{lurDech}
\tlhd{n}{Tradition.}
\tlhf{TKD}

\tlhw{lurgh}
\tlhd{n}{Direction (spatial).}
\tlhf{TKD}

\tlhw{lurSa\'}
\tlhd{name}{Lursa.}
\tlhf{KGT, HolQeD v5n3p15}[S26]

\tlhw{luSpet}
\tlhd{n}{Black hole (astronomical phenomenon).}
\tlhf{KGT}

\tlhw{lut}
\tlhd{n}{Story.}
\tlhf{TKD}

\tlhw{lutlh}
\tlhd{v}{Be primitive.}
\tlhf{TKD}

\tlhw{lu\'}
\tlhd{excl}{Yes, okay, I will.}
\tlhf{TKD}
