\chapter{\tlhe{m}}

\tlhw{mab}
\tlhd{n}{Treaty.}
\tlhf{TKD}

\tlhw{mach}
\tlhd{v}{Be small.}
\tlhf{TKD}

\tlhw{maDyar}
\tlhd{n}{Hungary.}
\tlhf{Eurotalk}

\tlhw{magh}
\tlhd{v}{Betray.}
\tlhf{TKD, MSN, HolQeD v5n3p15}[S26]

\tlhw{maghwI\'}
\tlhd{n}{Traitor.}
\tlhf{TKD}

\tlhw{maH num,pro.}
\tlhd{num}{ 1. Ten (combining form)/(number-forming element).}
\tlhd{pro}{We, us.}
\tlhco{The E-K listing somewhat confusingly includes \tlhe{maH} as "Ten (combining form)" while the K-E lists it as "ten (number-forming element)". There is no such confusion over words for hundred, thousand, ten-thousand, etc.}{\emph{Ed}}
\tlhf{TKD}

\tlhw{maHpIn}
\tlhd{n}{Bowl (large).}
\tlhf{KGT}

\tlhw{maj}
\tlhd{excl}{Good (expressing satisfaction).}
\tlhf{TKD}

\tlhw{majQa\'}
\tlhd{excl}{Well done, very good.}
\tlhf{TKD}

\tlhw{malja\'}
\tlhd{n}{Business.}
\tlhf{TKD}

\tlhw{mang}
\tlhd{n}{Soldier.}
\tlhf{TKD}

\tlhw{mangghom}
\tlhd{n}{Army.}
\tlhf{TKD}

\tlhw{manggho\' naH}
\tlhd{n}{Mango.}
\tlhf{Eurotalk}

\tlhw{mangHom}
\tlhd{n}{Cadet.}
\tlhf{TKD}

\tlhw{maq}
\tlhd{v}{Proclaim.}
\tlhf{TKD}

\tlhw{maqoch}
\tlhd{n}{Buddy, pal, close male friend of a male.}
\tlhf{KGT}

\tlhw{maQmIgh}
\tlhd{n}{Dark omen, sign of evil coming.}
\tlhf{HolQeD v1n3p9}

\tlhw{mar}
\tlhd{v}{Use the big toe.}
\tlhf{HolQeD v10n2p10}

\tlhw{mara}
\tlhd{name}{Mara, TOS "Day of the Dove" Kang's wife and Sci. officer.}
\tlhf{TKD}

\tlhw{martaq}
\tlhd{name}{Martok.}
\tlhf{KGT}

\tlhw{marwI\'}
\tlhd{v}{Big toe, first toe.}
\tlhf{HolQeD v10n2p7}

\tlhw{maS}
\tlhd{n}{Moon.}
\tlhd{v}{Prefer.}
\tlhf{TKD}

\tlhw{maSIr}
\tlhd{n}{Egypt.}
\tlhf{Eurotalk}

\tlhw{maSwov}
\tlhd{n}{Moonlight.}
\tlhf{TKD}

\tlhw{matHa\'}
\tlhd{n}{Gunner.}
\tlhf{TKD}

\tlhw{matlh name,}
\tlhd{v}{name. 1. Maltz.}
\tlhd{v}{Be loyal.}
\tlhf{TKD}

\tlhw{matlhHa\'}
\tlhd{v}{Be disloyal.}
\tlhf{TKD}

\tlhw{mavje\'}
\tlhd{n}{Liver. (Regional)}
\tlhf{TKD}

\tlhw{mavjop}
\tlhd{n}{Paper clip.}
\tlhf{TKD}

\tlhw{maw}
\tlhd{v}{Offend.}
\tlhf{TKD}

\tlhw{maw\'}
\tlhd{v}{Be crazy.}
\tlhf{TKD}

\tlhw{may}
\tlhd{n}{The consonant "m".}
\tlhd{v}{Be fair.}
\tlhf{TKD}[qep\'a\' wa\'maH javDIch 2009]

\tlhw{may\'}
\tlhd{n}{Battle.}
\tlhf{TKD}

\tlhw{may\' bom}
\tlhd{n}{Battle song.}
\tlhf{KGT}

\tlhw{may\'Duj}
\tlhd{n}{Battle cruiser.}
\tlhf{TKD}

\tlhw{may\'luch}
\tlhd{n}{Battle gear, panolpy, complete set of armor and weapons.}
\tlhf{KGT}

\tlhw{may\'morgh}
\tlhd{n}{Battle array.}
\tlhf{TKD}

\tlhw{may\'ron}
\tlhd{n}{Accordian, concertina.}
\tlhf{KGT}

\tlhw{ma\'}
\tlhd{v}{Accommodate.}
\tlhf{TKD}

\tlhw{ma\'veq}
\tlhd{n}{A ceremonial knife used in the ritual of Mauk To'Vor. One Federation Standard spelling of this weapon's name is /M'veQ/.}
\tlhco{Klingon spelling confirmed via email by Okrand}{\emph{Ed}}
\tlhf{M.Ok.}

\tlhw{meb}
\tlhd{n}{Guest.}
\tlhf{TKD}

\tlhw{mebpa\'mey}
\tlhd{n}{Hotel.}
\tlhf{KGT}

\tlhw{mech}
\tlhd{v}{Trade.}
\tlhf{TKD}

\tlhw{megh}
\tlhd{n}{Lunch.}
\tlhf{TKD}

\tlhw{megh\'an}
\tlhd{n}{End (of stick, rope, etc.), other end from \tlhe{\'er\'In}.}
\tlhcf{'er'In}
\tlhco{This can refer to either end of the rope, stick or whatever, but once you arbitrarily choose an end to call \tlhe{megh\'an}, the OTHER end is called \tlhe{\'er\'In}. Before you make this initial reference, either end can be called either \tlhe{\'er\'In} or \tlhe{megh\'an}.}
\tlhf{HolQeD v12n2p6}

\tlhw{meH}
\tlhd{n}{Bridge (of a ship).}
\tlhf{TKD}

\tlhw{mej}
\tlhd{v}{Leave, depart.}
\tlhf{TKD}

\tlhw{melchoQ}
\tlhd{n}{Marrow, bone marrow.}
\tlhf{KGT}

\tlhw{mellota\'}
\tlhd{name}{Melota, from the Klingon Opera.}
\tlhf{TKW}

\tlhw{mem}
\tlhd{n}{Catalog.}
\tlhf{TKD}

\tlhw{mep}
\tlhd{n}{Plastic.}
\tlhf{TKD}

\tlhw{meq}
\tlhd{n}{Reason, motive.}[Reason, logical thinking.]
\tlhd{v}{Reason.}
\tlhf{TKD, TKW}[KGT]

\tlhw{meqba\'}
\tlhd{n}{Legal proceeding, type of legal proceeding; portion of the trial or appeal where evidence is heard.}
\tlhf{TKD}[Star Trek Encyclopedia]

\tlhw{meqlaH}
\tlhd{n}{/mek'leth/, type of weapon.}
\tlhf{KGT}

\tlhw{meq\'ro vaq}
\tlhd{n}{Mekro'vak (region).}
\tlhf{KGT}

\tlhw{meQ}
\tlhd{v}{Burn.}
\tlhf{TKD}

\tlhw{mer}
\tlhd{v}{Surprise.}
\tlhf{TKD}

\tlhw{meSchuS}
\tlhd{n}{Large wind instrument.}
\tlhf{KGT}

\tlhw{mev}
\tlhd{v}{Stop, cease.}
\tlhf{TKD}

\tlhw{mevyap}
\tlhd{excl}{Stop, enough already.}
\tlhco{It comes from two Klingon verbs, \tlhe{mev} ("stop, cease") and \tlhe{yap} ("be enough, be sufficient"). No doubt in the past, the locution was longer, perhaps \tlhe{yImev, yap!} ("stop! it\'s enough"). The dropping of the imperative prefix \tlhe{yI-][pe-} is what would be expected in Clipped Klingon. The two words have been used in juxtaposition for so long, they have come to be accepted as a single, though grammatically peculiar, word.}{KGT 113}
\tlhf{KGT}

\tlhw{meyrI\'}
\tlhd{n}{Square (geometric shape).}
\tlhf{qep\'a\' wa\'maH cha\'DIch 2005}

\tlhw{meyIS tIr}
\tlhd{n}{Sweet corn.}
\tlhf{Eurotalk}

\tlhw{me\'}
\tlhd{n}{Aunt, mother's sister.}
\tlhf{HolQeD v9n3p8}

\tlhw{me\'nal}
\tlhd{n}{Aunt, mother's brother's wife.}
\tlhf{HolQeD v9n3p8}

\tlhw{mIch}
\tlhd{n}{Sector, zone.}
\tlhf{TKD}

\tlhw{mID}
\tlhd{n}{Colony.}
\tlhf{TKD}

\tlhw{mIgh}
\tlhd{v}{Be evil.}
\tlhf{TKD}

\tlhw{mIl}
\tlhd{v}{Be formerly honored.}
\tlhcf{'utlh}
\tlhco{Loss of honor is implied; an adjective to describe one who has fallen from grace; not applicable to \tlhe{\'utlh}, which implies a voluntary, honorable retirement.}{HolQeD v12n3p8}
\tlhf{HolQeD}

\tlhw{mIllogh qonwI\'}
\tlhd{n}{Camera.}
\tlhf{Eurotalk}

\tlhw{mIllogh qonwI\' qoSta\'}
\tlhd{n}{Film.}
\tlhf{Eurotalk}

\tlhw{mIl\'oD}
\tlhd{n}{Type of animal, sabre bear.}
\tlhf{KGT}

\tlhw{mIl\'oD ngeb}
\tlhd{n}{Teddy bear.}
\tlhf{Eurotalk}

\tlhw{mIm}
\tlhd{v}{Delay.}
\tlhf{TKD}

\tlhw{mIn}
\tlhd{n}{Eye.}
\tlhf{TKD}

\tlhw{mIn QanwI\' nguv}
\tlhd{n}{Sunglasses.}
\tlhf{Eurotalk}

\tlhw{mIp}
\tlhd{n}{Wealth, richness.}
\tlhd{v}{Be rich.}
\tlhf{TKD}[KGT]

\tlhw{mIqta\'}
\tlhd{n}{Machinery.}
\tlhf{TKD}

\tlhw{mIQ}
\tlhd{n}{Forehead. (regional)}
\tlhd{v}{Fry, deep-fry.}
\tlhco{Entries differs in source. One of the E K entries is used here.}
\tlhpun{Marc nodded, smiling infuriatingly, to confirm \tlhe{mIQ} (deep fry) as in \tlhe{mIQ}-Donald\'s french fries." (Robyn Stewart, qep\'a\' 1997)}
\tlhf{TKW}[KGT]

\tlhw{mIr}
\tlhd{n}{Chain.}
\tlhf{TKD}

\tlhw{mIS}
\tlhd{n}{Confusion.}
\tlhd{v}{Be confused, mixed up.}
\tlhf{TKD}

\tlhw{mISmoH}
\tlhd{v}{Confuse.}
\tlhf{TKD}

\tlhw{mIStaq}
\tlhd{name}{Sculptor who made Blood Oath Circles.}
\tlhf{KCD}

\tlhw{mIv}
\tlhd{n}{Helmet, hat.}[Metal, flat-bottomed pot. (regional)]
\tlhf{TKD, KGT}[Eurotalk]

\tlhw{mIv bargh}
\tlhd{n}{Metal, flat-bottomed pot. (regional)}
\tlhf{KGT}

\tlhw{mIv\'a\'}
\tlhd{n}{Crown.}
\tlhf{KGT}

\tlhw{mIw}
\tlhd{n}{Procedure, process, step, stage (in a process).}
\tlhf{TKD}[KGT]

\tlhw{mIy}
\tlhd{v}{Brag.}
\tlhf{TKD}

\tlhw{mI\'}
\tlhd{n}{Number.}
\tlhd{v}{Dance, run in place, do callisthenics, exercise.}
\tlhco{Entries differs in source. They are combined here.}{\emph{Ed}}
\tlhf{TKD}[KGT]

\tlhw{mob}
\tlhd{v}{Be alone.}
\tlhf{TKD}

\tlhw{moch}
\tlhd{n}{Superior.}
\tlhf{TKD}

\tlhw{moD}
\tlhd{v}{Hurry.}
\tlhf{TKD}

\tlhw{mogh name,}
\tlhd{v}{name. 1. Mogh. ­-v.}[Be frustrated.]
\tlhf{TKD}[KGT]

\tlhw{moH}
\tlhd{v}{Be ugly.}[Influence, exert undue influence on. (slang)]
\tlhco{Entries differs in source. One of the E-K entries is used here.}{\emph{Ed}}
\tlhf{TKD}

\tlhw{moHaq}
\tlhd{n}{Prefix.}
\tlhf{TKD}

\tlhw{moj}
\tlhd{v}{Become.}
\tlhf{TKD}

\tlhw{mojaq}
\tlhd{n}{Suffix.}
\tlhco{In English-Klingon side of TKD, this is \tlhe{\#mojaQ\#}.}{\emph{Ed}}
\tlhf{TKD}

\tlhw{mol}
\tlhd{n}{Grave.}
\tlhd{v}{Bury.}
\tlhf{TKD}

\tlhw{mon}
\tlhd{n}{Capital (of a place).}[Smile, grin, sneer.]
\tlhf{TKD}[KGT]

\tlhw{mong}
\tlhd{n}{Neck.}
\tlhf{TKD}

\tlhw{mongDech}
\tlhd{n}{Collar.}
\tlhf{TKD}

\tlhw{mong Ha\'quj}
\tlhd{n}{A tie.}
\tlhf{Eurotalk}

\tlhw{mop}
\tlhd{n}{Robe.}
\tlhf{TKD}

\tlhw{moq}
\tlhd{v}{Beat (something with an implement).}
\tlhf{KGT}

\tlhw{moQ}
\tlhd{n}{Sphere.}[Spiked pommel of a /d'k tahg/ knife.]
\tlhf{TKD}[KGT]

\tlhw{moQbara\'}
\tlhd{n}{/Mok'bara/, martial arts form.}
\tlhf{KGT}

\tlhw{morgh}
\tlhd{v}{Protest.}
\tlhf{TKD}

\tlhw{moS}
\tlhd{v}{Compromise.}
\tlhf{TKD}

\tlhw{motlh adv, v.}
\tlhd{adv}{Usually, typically, as expected, expectedly.}
\tlhd{v}{Be usual, be normal, be standard.}
\tlhco{Entries differs in source. They are combined here.}{\emph{Ed}}
\tlhf{TKD, BoP}[S14]

\tlhw{motlhbe\'}
\tlhd{v}{Be unusual.}
\tlhf{TKD}

\tlhw{mov}
\tlhd{n}{Top of foot.}
\tlhf{KGT}

\tlhw{moy\'bI\'}
\tlhd{n}{Slingshot.}
\tlhf{TKD}

\tlhw{mo\'}
\tlhd{n}{Cage.}[Motive, motivation, grounds, reason, rationale. (slang)]
\tlhf{TKD}[KGT]

\tlhw{mub}
\tlhd{v}{Be legal.}
\tlhf{TKD}

\tlhw{much}
\tlhd{n}{Presentation.}
\tlhd{v}{Present.}[Perform (music).]
\tlhf{TKD}[KGT]

\tlhw{muchwI\'}
\tlhd{n}{Musician.}
\tlhf{KGT}

\tlhw{muD}
\tlhd{n}{Atmosphere.}[Weather (in general).]
\tlhf{TKD}[KGT]

\tlhw{muD Dotlh}
\tlhd{n}{Weather (at any given moment).}
\tlhf{KGT}

\tlhw{mugh}
\tlhd{v}{Translate.}
\tlhf{TKD}

\tlhw{mughato\'}
\tlhd{n}{Mugato.}
\tlhf{TOS A Private Little War}[Star Trek: The Worlds Of The Federation]

\tlhw{mughwI\'}
\tlhd{n}{Translator.}
\tlhf{TKD}

\tlhw{muH}
\tlhd{v}{Execute, put to death.}
\tlhf{TKD}

\tlhw{muj}
\tlhd{v}{Be wrong.}
\tlhf{TKD}

\tlhw{mul}
\tlhd{v}{Be stubborn.}
\tlhf{TKD}

\tlhw{mum}
\tlhd{v}{Taste, sense flavor.}
\tlhf{KGT}

\tlhw{mun}
\tlhd{v}{Intervene.}
\tlhf{TKD}

\tlhw{mung}
\tlhd{n}{Origin.}
\tlhf{TKD}

\tlhw{mup}
\tlhd{v}{Impact, strike.}
\tlhf{TKD}

\tlhw{mupwI\'}
\tlhd{n}{Hammer.}
\tlhf{KGT}

\tlhw{mupwI\'Hom}
\tlhd{n}{Mallet (for striking a musical instrument).}
\tlhf{KGT}

\tlhw{muq}
\tlhd{v}{Have a volume of.}
\tlhf{TKD}

\tlhw{muS}
\tlhd{v}{Hate, detest.}
\tlhf{TKD}

\tlhw{mut}
\tlhd{n}{Species. ­-v.}[Be selfish.]
\tlhf{TKD}

\tlhw{muv}
\tlhd{v}{Join.}
\tlhf{TKD}

\tlhw{muvmoH}
\tlhd{v}{Recruit.}
\tlhf{TKD}

\tlhw{muvtay}
\tlhd{n}{Initiation.}
\tlhf{TKD}

\tlhw{mu\'}
\tlhd{n}{Word.}
\tlhf{TKD}

\tlhw{mu\'ghom}
\tlhd{n}{Dictionary.}
\tlhf{TKD}

\tlhw{mu\'mey Doy}
\tlhd{n}{"Tired words" (used to refer to older Klingons language).}
\tlhf{KGT}

\tlhw{mu\'mey ghoQ}
\tlhd{n}{Slang, "fresh words" (mostly used by younger Klingons).}
\tlhf{KGT}

\tlhw{mu\'mey ru\'}
\tlhd{n}{Temporary words (ungrammatical expression used for impact).}
\tlhf{KGT}

\tlhw{mu\'qaD}
\tlhd{n}{Curse.}
\tlhf{KGT}

\tlhw{mu\'qaD veS}
\tlhd{n}{Curse warfare.}

\tlhw{Written on the printed card.}
\tlhf{PK}

\tlhw{mu\'tay\'}
\tlhd{n}{Vocabulary.}
\tlhf{TKD}

\tlhw{mu\'tlhegh}
\tlhd{n}{Sentence.}
\tlhf{TKD}