\chapter{\tlhe{n}}

\tlhw{nab}
\tlhd{n}{Plan.}[Procedure.]
\tlhd{v}{Plan.}
\tlhf{TKD}

\tlhw{nach}
\tlhd{n}{Head.}
\tlhf{TKD}

\tlhw{naD}
\tlhd{n}{Commendation.}
\tlhd{v}{Praise.}[Commend.][Approve.]
\tlhf{TKD}

\tlhw{naDev}
\tlhd{n}{Here}[Hereabouts.]
\tlhf{TKD}

\tlhw{naDHa\'}
\tlhd{v}{Discommend.}[Disapprove.]
\tlhf{TKD}

\tlhw{naDHa\'ghach}
\tlhd{n}{Discommendation.}
\tlhf{TKD}

\tlhw{nagh}
\tlhd{n}{Rock.}[Stone.]
\tlhf{TKD}

\tlhw{naH}
\tlhd{n}{Fruit.}[Vegetable.]
\tlhf{TKD}

\tlhw{naj}
\tlhd{v}{Dream.}
\tlhf{TKD}

\tlhw{nap}
\tlhd{v}{Be simple.}
\tlhf{TKD}

\tlhw{naQ}
\tlhd{n}{Cain.}[Staff.]
\tlhd{v}{Be full.}[Be whole.][Be entire.]
\tlhf{TKD}

\tlhw{nargh}
\tlhd{v}{Appear.}[Escape.]
\tlhco{According to mythology, when a dishonored Klingon dies, his or her spirit goes to a place called Gre'thor, there to remain. To speak of spirits escaping from Gre'thor is to speak of an impossibility. The phrase usually follows the statement of what it is that supposedly cannot happen: \tlhtr{jIjegh ghe'torvo' narghDI' qa'pu'}{I will surrender when spirits escape from Gre'thor}.}{KGT}{117}
\tlhco{In Klingon, opportunities are captured (\tlhe{jon}), not taken; a missed opportunity is said to have escaped (\tlhe{nargh}).}{TKW}{51}
\tlhco{The verb \tlhe{nargh} ... means "escape", but the same word, or a phonetically identical one, means "appear". Thus, perhaps the Klingons are saying that when a warrior dies, his spirit appears, whereas prior to death it was hidden or disguised by the body. Another interpretation is that the spirit was held prisoner by the body. Worf told Jeremy, whose mother had been killed, "In my tradition, we do not grieve the loss of the body. We celebrate the releasing of the spirt."}{TKW}{145}
\tlhco{I'd probably take an idiomatic approach incorporating the phrase \tlhe{nargh \'eb} "the opportunity escapes" ... This goes along with other expressions such as \tlhe{\'eb jon} "he/she captures the opportunity" or, more colloquially "he/she seizes the opportunity.}{st.k 1/98}
\tlhc
\tlhca{jInargh}{I escape.}[KLS]
\tlhe{narghta\'}{[They have] Escaped.}
\tlhe{taj DopmeyDaq nargh cha\' tajHom}{Two small, secondary blades eject from the knife's sides.}
\tlhe{narghbe\'chugh SuvwI\' qa\' taH may\'}{If the warrior's spirit has not escaped, the battle is still going on.}
\tlhe{HeghDI\' SuvwI\' nargh SuvwI\' qa\'}{When a warrior dies, his spirit escapes.}
\tlhe{\'avwI\' nejDI\' narghta\'bogh qama\' reH \'avwI\' Sambej}{When an escaped prisoner looks for a guard, he always finds one.}
\tlhe{ghe\'torvo\' narghDI\' qa\'pu\'}{When spirits escape from Gre'thor.}
\tlhe{jIpaSqu\'mo\' narghpu\' qaSuchmeH \'eb}{I was too late to visit you. (Because I'm very late, the opportunity to visit you has escaped.)}
\tlhe{qaSuch vIneH \'ach narghpu\' \'eb. jIpaSqu\'}{I was too late to visit you. (I want to visit you, but the opportunity has escaped. I am very late.)}
\tlhe{qaSuchlaHbe\'. jIpaSqu\' vaj narghpu\' \'eb}{I was too late to visit you. (I cannot visit you. I am very late, thus the opportunity has escaped.)}
\tlhcf{Haw'}[mej][qeD][tlheD]
\tlhf{TKD, TKW, KGT}[st.k]

\tlhw{naS}
\tlhd{v}{Be vicious.}
\tlhf{TKD}

\tlhw{nav}
\tlhd{n}{Paper.}
\tlhf{TKD}

\tlhw{nav qatwI\'}
\tlhd{n}{Envelope.}
\tlhf{Eurotalk}

\tlhw{nawlogh}
\tlhd{n}{Squadron.}
\tlhf{TKD}

\tlhw{naw\'}
\tlhd{v}{Access.}
\tlhf{TKD}

\tlhw{nay}
\tlhd{n}{The consonant "n".}
\tlhd{v}{Marry (wife does this).}

\tlhw{Cf.}[\tlhe{Saw}]

\tlhw{PUN:}[\tlhe{nay} "marry (wife does this)" and \tlhe{Saw] "marry (husband does this)" comes from "nay" and "sure" with a Southern drawl. (Chet Braun)}
\tlhf{TKD}[qep\'a\' wa\'maH javDIch 2009]

\tlhw{nay\'}
\tlhd{n}{Course.}[Dish (at a meal).]
\tlhf{TKD}

\tlhw{nech}
\tlhd{v}{Be lateral.}[Move laterally.]
\tlhf{TKD}

\tlhw{negh}
\tlhd{n}{Soldiers.}
\tlhf{TKD}

\tlhw{neH adv, v.}
\tlhd{adv}{Only, merely, just.}
\tlhd{v}{Want.}
\tlhf{TKD}

\tlhw{neHmaH}
\tlhd{n}{Neutral zone.}
\tlhf{TKD}

\tlhw{nej}
\tlhd{v}{Look for, seek, search for.}
\tlhf{TKD}

\tlhw{nejwI\'}
\tlhd{n}{Probe.}
\tlhf{TKD}

\tlhw{nem}
\tlhd{n}{Years from now.}
\tlhf{TKD}

\tlhw{nenghep}
\tlhd{n}{Age of Ascension.}
\tlhf{TKD}

\tlhw{nentay}
\tlhd{n}{Rite of Ascension.}
\tlhf{TKD}

\tlhw{nep}
\tlhd{v}{Lie, fib.}
\tlhf{TKD}

\tlhw{net}
\tlhd{pro}{That (previous topic).}
\tlhf{TKD}

\tlhw{netlh}
\tlhd{num}{Ten thousand.}
\tlhf{TKD}

\tlhw{ne\'}
\tlhd{n}{Yeoman.}
\tlhf{TKD}

\tlhw{ne\'Derlan}
\tlhd{n}{The Netherlands.}
\tlhf{Eurotalk}

\tlhw{nIb}
\tlhd{v}{Be identical.}
\tlhf{TKD}

\tlhw{nIch}
\tlhd{n}{Ammunition.}
\tlhf{TKD}

\tlhw{nID}
\tlhd{v}{Attempt.}[Try.]
\tlhf{TKD}

\tlhw{nIH}
\tlhd{n}{Right (side).}
\tlhd{v}{Steal.}
\tlhf{TKD}

\tlhw{nIHwI\'}
\tlhd{n}{Thief.}
\tlhf{TKD}

\tlhw{nIj}
\tlhd{v}{Leak.}
\tlhf{TKD}

\tlhw{nIm tlhagh}
\tlhd{n}{Butter.}
\tlhf{Eurotalk}

\tlhw{nIm wIb ngogh}
\tlhd{n}{Cheese.}
\tlhf{Eurotalk}

\tlhw{nImbuS wej}
\tlhd{n}{Nimbus III.}
\tlhf{TKD}

\tlhw{nIn}
\tlhd{n}{Fuel.}
\tlhf{TKD}

\tlhw{nIpon}
\tlhd{n}{Japan.}
\tlhf{Eurotalk}

\tlhw{nIQ}
\tlhd{n}{Breakfast.}
\tlhf{TKD}

\tlhw{nIS}
\tlhd{v}{Hinder.}[Interfere.][Interfere with.][Disrupt.]
\tlhf{TKD}

\tlhw{nIteb}
\tlhd{adv}{Alone.}[Acting alone.]
\tlhf{TKD}

\tlhw{nItlh}
\tlhd{n}{Finger.}
\tlhf{TKD}

\tlhw{nItlhpach}
\tlhd{n}{Fingernail.}
\tlhf{TKD}

\tlhw{nIv}
\tlhd{v}{Be superior.}
\tlhf{TKD}

\tlhw{nIvnav}
\tlhd{n}{Pajamas.}
\tlhf{TKD}

\tlhw{nIyma\'}
\tlhd{n}{Phantom.}[Apparition.]
\tlhco{Something that seems to appear, but isn't really there. It's not the same as \tlhtr{qa'}{ghost}, which is the spirit of a dead person.}{M.Ok.}
\tlhf{email from Marc Okrand, 23 November 2008}

\tlhw{nI\'}
\tlhd{v}{Be long.}[Lengthy (duration).]
\tlhf{TKD}

\tlhw{nob}
\tlhd{n}{Gift.}
\tlhd{v}{Give.}
\tlhf{TKD}

\tlhw{noch}
\tlhd{n}{Sensor.}
\tlhf{TKD}

\tlhw{noD}
\tlhd{v}{Retaliate.}
\tlhf{TKD}

\tlhw{nogh}
\tlhd{v}{Writhe.}
\tlhf{TKD}

\tlhw{noH}
\tlhd{n}{War.}
\tlhd{v}{Judge.][Estimate.}
\tlhf{TKD}

\tlhw{noj}
\tlhd{v}{Lend.}
\tlhf{TKD}

\tlhw{nol}
\tlhd{n}{Funeral.}
\tlhf{TKD}

\tlhw{nom}
\tlhd{adv}{Fast.}[Quickly.]
\tlhf{TKD}

\tlhw{non}
\tlhd{v}{Be rotten.}
\tlhf{TKD}

\tlhw{nong}
\tlhd{v}{Be passionate.}
\tlhf{TKD}

\tlhw{nop}
\tlhd{v}{Omit.}
\tlhf{TKD}

\tlhw{noregh}
\tlhd{n}{Norway.}
\tlhf{Eurotalk}

\tlhw{noSvagh}
\tlhd{n}{Deodorant.}
\tlhf{TKD}

\tlhw{not}
\tlhd{adv}{Never.}
\tlhf{TKD}

\tlhw{notlh}
\tlhd{v}{Be obsolete.}
\tlhf{TKD}

\tlhw{nov}
\tlhd{n}{Alien.}[Foreigner.]
\tlhd{v}{Be foreign.}[Be alien.]
\tlhf{TKD}

\tlhw{noy}
\tlhd{v}{Be famous.}[Well known.]
\tlhf{TKD}

\tlhw{no\'}
\tlhd{n}{Ancestors.}
\tlhf{TKD}

\tlhw{nub}
\tlhd{v}{Be suspect.}
\tlhf{TKD}

\tlhw{nubwI\'}
\tlhd{n}{Predecessor.}
\tlhf{TKD}

\tlhw{nuch}
\tlhd{n}{Coward.}
\tlhf{TKD}

\tlhw{nuD}
\tlhd{v}{Examine.}
\tlhf{TKD}

\tlhw{nugh}
\tlhd{n}{Society.}
\tlhf{TKD}

\tlhw{nuH}
\tlhd{n}{Weapon.}
\tlhf{TKD}

\tlhw{nuHHom}
\tlhd{n}{Small arms.}
\tlhf{TKD}

\tlhw{nuHmey}
\tlhd{n}{Arsenal.}
\tlhf{TKD}

\tlhw{nuj}
\tlhd{n}{Mouth.}
\tlhf{TKD}

\tlhw{num}
\tlhd{v}{Promote.}
\tlhf{TKD}

\tlhw{nung}
\tlhd{v}{Precede.}
\tlhf{TKD}

\tlhw{nup}
\tlhd{v}{Decrease.}
\tlhf{TKD}

\tlhw{nuqjatlh}
\tlhd{excl}{What did you say?}
\tlhf{TKD}

\tlhw{nural}
\tlhd{n}{Neural.}
\tlhf{TKD}

\tlhw{nuralngan}
\tlhd{n}{Neuralese.}
\tlhf{TKD}

\tlhw{nuq}
\tlhd{ques}{What?}
\tlhf{TKD}

\tlhw{nuqDaq}
\tlhd{ques}{Where?}
\tlhf{TKD}

\tlhw{nuqneH excel. What do you want? (greeting)}
\tlhf{TKD}

\tlhw{nuQ}
\tlhd{v}{Annoy}[Bother.]
\tlhf{TKD}

\tlhw{nur}
\tlhd{n}{Dignity.}
\tlhf{TKD}

\tlhw{nuS}
\tlhd{v}{Ridicule.}
\tlhf{TKD}

\tlhw{nuv}
\tlhd{n}{Person (humanoid).}
\tlhf{TKD}

\tlhw{nu\'SIylan}
\tlhd{n}{New Zealand.}
\tlhf{Eurotalk}
