\chapter{\tlhe{p}}

\tlhw{pab}
\tlhd{n}{Grammar.}
\tlhd{v}{Follow (rules).}
\tlhf{TKD}

\tlhw{pach}
\tlhd{n}{Claw (n)}
\tlhf{TKD}

\tlhw{pagh}
\tlhd{conj}{Or, either/or (joining sentences).}
\tlhd{n}{Nothing, none.}
\tlhf{TKD}

\tlhw{pagh}
\tlhd{num}{Zero.}
\tlhf{TKD}

\tlhw{paH}
\tlhd{n}{Gown, dress.}
\tlhf{TKD}[Eurotalk]

\tlhw{paH bID}
\tlhd{n}{Skirt.}
\tlhf{Eurotalk}

\tlhw{paj}
\tlhd{v}{Resign.}
\tlhf{TKD}

\tlhw{pan}
\tlhd{v}{Spark, emit sparks.}
\tlhco{As the duel between Kahless and Molor continues, the opera repeats a phrase several times: \tlhtr{pan qeylIS betleH 'ej jach molor 'etlh}{Kahless's batlh'etlh sparks and Molor's sword roars.} I might not be remembering the lyrics exactly, so please don't take them as perfect canon.}{ghunchu'wI' 'utlh}
\tlhf{qep\'a\' wa\'maH chorghDIch 2011}

\tlhw{paq}
\tlhd{n}{Book.}
\tlhf{TKD}

\tlhw{paQDI\'norgh}
\tlhd{n}{Teachings.}
\tlhf{TKD}

\tlhw{par}
\tlhd{v}{Dislike.}
\tlhf{TKD}

\tlhw{parHa\'}
\tlhd{v}{Like.}
\tlhf{TKD}

\tlhw{paSlogh}
\tlhd{n}{Socks.}
\tlhf{TKD}

\tlhw{pat}
\tlhd{n}{System.}
\tlhf{TKD}

\tlhw{patat \'oQparmey}
\tlhd{pl n}{Potatoes.}
\tlhf{Eurotalk}

\tlhw{patlh}
\tlhd{n}{Rank (military, governmental).}
\tlhf{TKD}

\tlhw{pav}
\tlhd{v}{Be urgent.}
\tlhf{TKD}

\tlhw{paw}
\tlhd{v}{Arrive.}
\tlhf{TKD}

\tlhw{paw\'}
\tlhd{v}{Collide.}
\tlhf{TKD}

\tlhw{pay}
\tlhd{n}{The consonant "p".}
\tlhf{qep\'a\' wa\'maH javDIch 2009}

\tlhw{pay}
\tlhd{v}{Regret.}
\tlhf{TKD}

\tlhw{pay\'}
\tlhd{adv}{Suddenly.}
\tlhf{TKD}

\tlhw{pa\'}
\tlhd{n}{Room.}[There, over there, thereabouts.]
\tlhf{TKD}

\tlhw{peD}
\tlhd{v}{Snow.}
\tlhf{TKD}

\tlhw{pegh}
\tlhd{n}{Secret.}
\tlhd{v}{Keep something secret.}
\tlhf{TKD}

\tlhw{peHghep}
\tlhd{n}{Age of Inclusion.}
\tlhf{TKD}

\tlhw{pej}
\tlhd{v}{Demolish.}
\tlhf{TKD}

\tlhw{pem}
\tlhd{n}{Daytime.}
\tlhf{TKD}

\tlhw{pemjep}
\tlhd{n}{Midday.}
\tlhf{TKD}

\tlhw{peng}
\tlhd{n}{Torpedo.}
\tlhf{TKD}

\tlhw{pep}
\tlhd{v}{Raise.}
\tlhf{TKD}

\tlhw{peQ}
\tlhd{n}{Magnetism.}
\tlhf{TKD}

\tlhw{peQ chem}
\tlhd{n}{Magnetic field.}
\tlhf{TKD}

\tlhw{per}
\tlhd{n}{Label.}
\tlhd{v}{Label.}
\tlhf{TKD}

\tlhw{per naHmey}
\tlhd{pl n}{Pears.}
\tlhf{Eurotalk}

\tlhw{pey}
\tlhd{n}{Acid.}
\tlhf{TKD}

\tlhw{pe\'}
\tlhd{v}{Cut.}
\tlhf{TKD}

\tlhw{pe\'vIl}
\tlhd{adv}{Forcefully, by force.}
\tlhf{TKD}

\tlhw{pIch}
\tlhd{n}{Fault, blame.}
\tlhd{v}{Blame.}
\tlhf{TKD}

\tlhw{pIgh}
\tlhd{n}{Ruins.}
\tlhf{TKD}

\tlhw{pIH}
\tlhd{n}{Expect.}
\tlhd{v}{Be suspicious.}
\tlhf{TKD}

\tlhw{pIlam naH}
\tlhd{n}{Plum.}
\tlhf{Eurotalk}

\tlhw{pIj}
\tlhd{adv}{Often.}
\tlhf{TKD}

\tlhw{pIm}
\tlhd{v}{Be different.}
\tlhf{TKD}

\tlhw{pIn}
\tlhd{n}{Boss.}
\tlhf{TKD}

\tlhw{pIn\'a\'}
\tlhd{n}{Master.}
\tlhf{TKD}

\tlhw{pIp}
\tlhd{n}{Spine.}
\tlhf{TKD}

\tlhw{pIpyuS}
\tlhd{n}{Pipius.}
\tlhf{TKD}

\tlhw{pIqaD}
\tlhd{n}{Klingon writing system.}
\tlhf{TKD}

\tlhw{pIQHa\'}
\tlhd{v}{Be indirect, be roundabout, be devious.}
\tlhco{So \tlhe{pIQ} would be "be direct."}{M.Ok.}
\tlhf{email to Agnieszka, 2 january 2012}

\tlhw{pItSa\' chab}
\tlhd{n}{Pizza.}
\tlhf{Eurotalk}

\tlhw{pItlh excel. Done.}
\tlhf{TKD}

\tlhw{pIv}
\tlhd{v}{Be healthy.}
\tlhc
\tlhca{tlhIngan yoH pIv verengan yoH rop}{The Klingon is braver than the Ferengi. (slang)}{KGT}[179]
\tlhf{TKD}[KTG.]

\tlhw{pIvghor}
\tlhd{n}{Warp drive.}
\tlhf{TKD}

\tlhw{pIvlob}
\tlhd{n}{Warp factor.}
\tlhf{TKD}

\tlhw{pIw}
\tlhd{n}{Odor.}
\tlhf{TKD}

\tlhw{pI\'}
\tlhd{v}{Be fat.}
\tlhf{TKD}

\tlhw{po}
\tlhd{n}{Morning.}
\tlhf{TKD}

\tlhw{pob}
\tlhd{n}{Hair (on body).}
\tlhf{TKD}

\tlhw{poch}
\tlhd{v}{Plant.}
\tlhf{TKD}

\tlhw{poD}
\tlhd{v}{Be clipped.}
\tlhf{TKD}

\tlhw{pogh}
\tlhd{n}{Glove.}
\tlhf{TKD}

\tlhw{poH}
\tlhd{n}{Period of time.}
\tlhd{v}{Time.}
\tlhf{TKD}

\tlhw{poj}
\tlhd{n}{Analysis.}
\tlhd{v}{Analyze.}
\tlhf{TKD}

\tlhw{pol}
\tlhd{v}{Keep, save.}
\tlhf{TKD}

\tlhw{pom}
\tlhd{n}{Dysentery.}
\tlhf{TKD}

\tlhw{pon}
\tlhd{v}{Persuade.}[Convince.]
\tlhf{TKD}

\tlhw{pong}
\tlhd{n}{Name.}
\tlhd{v}{Name.}[Call.]
\tlhf{TKD}

\tlhw{pop}
\tlhd{n}{Reward.}
\tlhf{TKD}

\tlhw{poq}
\tlhd{n}{Indigestion.}
\tlhf{TKD}

\tlhw{poQ}
\tlhd{v}{Demand.}[Require.]
\tlhf{TKD}

\tlhw{porgh}
\tlhd{n}{Body.}
\tlhf{TKD}

\tlhw{portughal}
\tlhd{n}{Portugal.}
\tlhf{Eurotalk}

\tlhw{poS}
\tlhd{n}{Left (side).}
\tlhd{v}{Be open.}[Opened.]
\tlhf{TKD}

\tlhw{poSmoH}
\tlhd{v}{Open (v)}
\tlhf{TKD}

\tlhw{potlh}
\tlhd{n}{Consequential thing.}[Something important.]
\tlhd{v}{Be important.}
\tlhf{TKD}

\tlhw{pov}
\tlhd{n}{Afternoon.}
\tlhd{v}{Be excellent.}
\tlhf{TKD}

\tlhw{po\'}
\tlhd{v}{Be expert.}[Skilled.]
\tlhf{TKD}

\tlhw{pub}
\tlhd{v}{Boil.}
\tlhf{TKD}

\tlhw{puch}
\tlhd{n}{Toilet.}
\tlhf{TKD}

\tlhw{puchpa\'}
\tlhd{n}{Washroom.}
\tlhf{TKD}

\tlhw{pugh}
\tlhd{n}{Dregs.}
\tlhf{TKD}

\tlhw{puH}
\tlhd{n}{Land.}
\tlhf{TKD}

\tlhw{puj .v Be weak.}
\tlhf{TKD}

\tlhw{pujmoH}
\tlhd{v}{Weaken.}
\tlhf{TKD}

\tlhw{pujwI\'}
\tlhd{n}{Weakling.}
\tlhf{TKD}

\tlhw{pum}
\tlhd{n}{Accusation.}
\tlhd{v}{Accuse.}[Fall.]
\tlhf{TKD}

\tlhw{pung}
\tlhd{n}{Mercy.}
\tlhf{TKD}

\tlhw{pup}
\tlhd{v}{Be perfect.}[Exact.][Kick.]
\tlhf{TKD}

\tlhw{puq}
\tlhd{n}{Child.}[Offspring.]
\tlhf{TKD}

\tlhw{puqbe\'}
\tlhd{n}{Daughter.}
\tlhf{TKD}

\tlhw{puqloD}
\tlhd{n}{Son.}
\tlhf{TKD}

\tlhw{puq chonnaQ}
\tlhd{n}{Fork.}
\tlhf{Eurotalk}

\tlhw{puQ}
\tlhd{v}{Be fed up.}
\tlhf{TKD}

\tlhw{puS}
\tlhd{v}{Be few.}[Be several.][Be a handful.][Sight (with gunsight).]
\tlhf{TKD}

\tlhw{puv}
\tlhd{v}{Fly.}
\tlhf{TKD}

\tlhw{puy}
\tlhd{v}{Wreck.}
\tlhf{TKD}

\tlhw{puyjaq}
\tlhd{n}{Nova.}
\tlhf{TKD}

\tlhw{pu\'}
\tlhd{n}{Phaser.}
\tlhf{TKD}

\tlhw{pu\'beH}
\tlhd{n}{Phaser rifle.}
\tlhf{TKD}

\tlhw{pu\'beq}
\tlhd{n}{Phaser crew.}
\tlhf{TKD}

\tlhw{pu\'DaH}
\tlhd{n}{Phaser banks.}
\tlhf{TKD}

\tlhw{pu\'HIch}
\tlhd{n}{Phaser pistol.}
\tlhf{TKD}

\tlhw{pu\'jIn}
\tlhd{n}{Map, plan [big piece of paper showing streets etc.]}[Construction plan, drawing, blueprint.]
\tlhf{qep'a' 2007}[email from Marc Okrand on 20 november 2009]
