\chapter{\tlhe{q}}

\tlhw{qab}
\tlhd{n}{Face.}
\tlhd{v}{Be bad.}
\tlhf{TKD}

\tlhw{qach}
\tlhd{n}{Building.}[Structure.]
\tlhf{TKD}

\tlhw{qaD}
\tlhd{n}{Challenge.}
\tlhd{v}{Challenge.}
\tlhf{TKD}

\tlhw{qagh n, v,}
\tlhd{n}{Serpent worm (as food).}[The glottal stop "'".]
\tlhd{v}{Interrupt.}
\tlhf{TKD}[qep\'a\' wa\'maH javDIch 2009]

\tlhw{qaH}
\tlhd{n}{Sir.}
\tlhf{TKD}

\tlhw{qal}
\tlhd{v}{Be corrupt.}
\tlhf{TKD}

\tlhw{qalmoH}
\tlhd{v}{Corrupt.}
\tlhf{TKD}

\tlhw{qam}
\tlhd{n}{Foot.}
\tlhf{TKD}

\tlhw{qama\'}
\tlhd{n}{Prisoner.}
\tlhf{TKD}

\tlhw{qan}
\tlhd{v}{Be old (not young).}
\tlhf{TKD}

\tlhw{qap}
\tlhd{v}{Insist.}
\tlhf{TKD}

\tlhw{qaq}
\tlhd{v}{Be preferable.}
\tlhf{TKD}

\tlhw{qar}
\tlhd{v}{Be accurate.}
\tlhf{TKD}

\tlhw{qarDaSQa\'}
\tlhd{n}{Cardassian Union.}
\tlhf{Klingon Monopoly}

\tlhw{qaS}
\tlhd{v}{Occur, happen.}
\tlhf{TKD}

\tlhw{qat}
\tlhd{v}{Wrap.}
\tlhf{TKD}

\tlhw{qatlhDa\'}
\tlhd{n}{Water pipe.}
\tlhco{The kind that carries water from place to place, not what you use for smoking.}{M.Ok.}
\tlhf{email to Lieven, 9 january 2012}

\tlhw{qatlh}
\tlhd{ques}{Why?}
\tlhf{TKD}

\tlhw{qav\'ap}
\tlhd{n}{[Untranslated.]}
\tlhc
\tlhca{valuevoHDajbo' qav'ap.}{Value mortgage.}[Klingon Monopoly]
\tlhf{Klingon Monopoly}

\tlhw{qaw}
\tlhd{v}{Remember.}
\tlhf{TKD}

\tlhw{qawHaq}
\tlhd{n}{Data banks (singular).}
\tlhf{TKD}

\tlhw{qawHaq}
\tlhd{n}{Memory banks.}
\tlhf{TKD}

\tlhw{qawmoH}
\tlhd{v}{Remind.}
\tlhf{TKD}

\tlhw{qay}
\tlhd{n}{The consonant "q".}
\tlhf{qep\'a\' wa\'maH javDIch 2009}

\tlhw{qay\'}
\tlhd{v}{Be a problem, be a hassle.}
\tlhf{TKD}

\tlhw{qa\'}
\tlhd{v}{Replace.}
\tlhco{This verb has only been used in the noun \tlhw{qa\'meH vIttlhegh} so far, derived from \tlhw{quvqa\'meH}, here it's a verb.}{?}
\tlhc
\tlhca{Quj wa\'DIch Monopoly Huch nav qa\' tlhIngan QaS.}{Klingon forces replace original Monopoly bills.}
\tlhf{Klingon Monopoly}

\tlhw{qa\'naDa\'}
\tlhd{n}{Canada.}
\tlhf{Eurotalk}

\tlhw{qa\'vam}
\tlhd{n}{Genesis.}
\tlhf{TKD}

\tlhw{qech}
\tlhd{n}{Idea.}
\tlhf{TKD}

\tlhw{qeD}
\tlhd{v}{Vacate.}
\tlhf{TKD}

\tlhw{qeH}
\tlhd{v}{Resent.}
\tlhf{TKD}

\tlhw{qej}
\tlhd{v}{Be grouchy, mean.}
\tlhf{TKD}

\tlhw{qel}
\tlhd{v}{Consider, take into account.}
\tlhf{TKD}

\tlhw{qelI\'qam}
\tlhd{n}{Kellicam.}
\tlhf{TKD}

\tlhw{qem}
\tlhd{v}{Bring.}
\tlhf{TKD}

\tlhw{qempa\'}
\tlhd{n}{Ancestor.}
\tlhf{TKD}

\tlhw{qeng}
\tlhd{v}{Carry, convey.}
\tlhf{TKD}

\tlhw{qep}
\tlhd{n}{Meeting.}
\tlhf{TKD}

\tlhw{qeq}
\tlhd{n}{Drill (military).}
\tlhf{TKD}

\tlhw{qeq}
\tlhd{v}{Practice, train, prepare.}
\tlhf{TKD}

\tlhw{qeS}
\tlhd{n}{Advice.}
\tlhd{v}{Advise.}
\tlhf{TKD}

\tlhw{qet}
\tlhd{v}{Run, jog.}
\tlhf{TKD}

\tlhw{qetlh}
\tlhd{v}{Be dull, uninteresting.}
\tlhf{TKD}

\tlhw{qev}
\tlhd{v}{Crowd.}
\tlhf{TKD}

\tlhw{qevaS}
\tlhd{n}{Kevas.}
\tlhf{TKD}

\tlhw{qevpob}
\tlhd{n}{Cheek.}
\tlhf{TKD}

\tlhw{qe\'rot \'oQqar}
\tlhd{n}{Carrot.}
\tlhf{Eurotalk}

\tlhw{qIb}
\tlhd{n}{Galaxy.}
\tlhf{TKD}

\tlhw{qIbHeS}
\tlhd{n}{Galactic rim.}
\tlhf{TKD}

\tlhw{qIch}
\tlhd{v}{Condemn.}
\tlhf{TKD}

\tlhw{qIgh}
\tlhd{n}{Shortcut.}
\tlhf{TKD}

\tlhw{qIH}
\tlhd{v}{Meet (for the first time).}
\tlhf{TKD}

\tlhw{qIj}
\tlhd{v}{Be black.}
\tlhf{TKD}

\tlhw{qIj \'e wov}
\tlhd{v}{Be gray.}
\tlhf{Eurotalk}

\tlhw{qIl}
\tlhd{v}{Cancel.}
\tlhf{TKD}

\tlhw{qIm}
\tlhd{v}{Pay attention, concentrate.}
\tlhf{TKD}

\tlhw{qImHa\'}
\tlhd{v}{Disregard.}
\tlhf{TKD}

\tlhw{qIp}
\tlhd{v}{Hit (with hand, fist, implement).}
\tlhf{TKD}

\tlhw{qIQ}
\tlhd{v}{Mutiny.}
\tlhf{TKD}

\tlhw{qIt}
\tlhd{v}{Be possible.}
\tlhf{TKD}

\tlhw{qIv}
\tlhd{n}{Knee.}
\tlhf{TKD}

\tlhw{qI\'}
\tlhd{v}{Sign (a treaty).}
\tlhf{TKD}

\tlhw{qoch}
\tlhd{n}{Partner.}
\tlhf{TKD}

\tlhw{qogh}
\tlhd{n}{Belt.}
\tlhf{TKD}

\tlhw{qoH}
\tlhd{n}{Fool.}
\tlhf{TKD}

\tlhw{qoj}
\tlhd{n}{Cliff.}
\tlhf{TKD}

\tlhw{qoj}
\tlhd{conj}{Or, and/or (joining sentences.)}
\tlhf{TKD}

\tlhw{qolqoS}
\tlhd{n}{Core, the (important) core of a group.}
\tlhf{qep'a' 2007}[http://www.kli.org/tlhIngan-Hol/2007/August/msg00021.html]

\tlhw{qon}
\tlhd{v}{Record.}
\tlhf{TKD}

\tlhw{qop}
\tlhd{v}{Arrest.}
\tlhf{TKD}

\tlhw{qoq}
\tlhd{n}{Robot.}
\tlhf{TKD}

\tlhw{qor}
\tlhd{v}{Scavenge.}
\tlhf{TKD}

\tlhw{qorDu\'}
\tlhd{n}{Family.}

\tlhw{References: naHqun\'s charts of Klingon family relationships.}
\tlhf{TKD}

\tlhw{qoSta\'}
\tlhd{n}{Strip, tape (not necessarily sticky)}
\tlhco{untranslated in TalkNow-CD, suggestion for a translation by SuStel.}{SuStel}
\tlhf{SuStel, 2 November 2011}

\tlhw{qotlh}
\tlhd{v}{Tickle.}
\tlhf{TKD}

\tlhw{qoS}
\tlhd{n}{Birthday.}
\tlhf{TKD}

\tlhw{qoy\'}
\tlhd{v}{Plead, beg.}
\tlhf{TKD}

\tlhw{qo\'}
\tlhd{n}{World, realm.}
\tlhf{TKD}

\tlhw{qo\'la\' \'awje\'}
\tlhd{n}{Cola.}
\tlhf{Eurotalk}

\tlhw{qub}
\tlhd{v}{Be rare.}
\tlhf{TKD}

\tlhw{quchv}
\tlhd{v}{Kidnap.}
\tlhf{TKD}

\tlhw{qugh}
\tlhd{v}{Cruise.}
\tlhf{TKD}

\tlhw{qughDuj}
\tlhd{n}{cruiser.}
\tlhf{TKD}

\tlhw{quHvaj}
\tlhd{n}{Dandruff.}
\tlhf{TKD}

\tlhw{qul}
\tlhd{n}{Fire.}
\tlhf{TKD}

\tlhw{qul naQmey}
\tlhd{pl n}{Matches.}
\tlhf{Eurotalk}

\tlhw{qum}
\tlhd{n}{Government.}
\tlhd{v}{Govern.}
\tlhco{The entirety of a government's military apparatus is its \tlhe{QI\'}, usually translated simply as "military"."}{KGT}{51}
\tlhco{To determine one's place in Klingon society or to understand who has authority over whom, one must be aware of the \tlhe{qutluch patlh} of the military, the government, the house with which one is associated, and so on.}{KGT}{111}
\tlhco{Literally meaning 'cape', this slang term is an example of metonymy, whereby one word (or phrase) stands for another with which it is associated. In this case, the leader of the Klingon High Council typically wears a ceremonial cape. The slang sense of \tlhe{ngup} does not refer only to this position on the High Council, however, but to anyone in power or even to "the power structure." Someone unsatisfied with the status quo, a rebel, might be said to want to \tlhe{ngup qaD} (\'challenge those in charge\'; literally, \'challenge the cape\'). Nonslang equivalents of \tlhe{ngup} are \tlhe{woQ} ("authority, political power") and \tlhe{qum} ("government").}{KGT}{157}
\tlhco{The Order of the Bat'leth (betleH 'obe'), the highest honor that the government bestows on a Klingon.}{KGT}{183}
\tlhc
\tlhca{pa\' \'oHtaH vaS\'a\'\'e\'. tlhIngan qum waw\' \'oH.}{This is where the Klingon Great Hall is located, the center of the Klingon government.}[TKD]
\tlhcf{loH}[HI'tuy][che'][ghatlh][vu'][loH][lu]
\tlhf{TKD}

\tlhw{qumwI\'}
\tlhd{n}{Governor.}
\tlhcf{DevwI'}[vu'wI'][HI'][pIn][ta'/voDleH][Qang][loHwI']
\tlhf{TKD}

\tlhw{qun}
\tlhd{n}{History.}
\tlhf{TKD}

\tlhw{qun}
\tlhd{v}{Scold.}
\tlhf{TKD}

\tlhw{qup}
\tlhd{n}{Elder.}
\tlhf{TKD}

\tlhw{quprIp}
\tlhd{n}{Council of Elders.}
\tlhf{TKD}

\tlhw{quq}
\tlhd{v}{Happen simultaneously.}
\tlhf{TKD}

\tlhw{qur}
\tlhd{v}{Be greedy.}
\tlhf{TKD}

\tlhw{quS}
\tlhd{n}{Chair.}
\tlhf{TKD}

\tlhw{qut}
\tlhd{n}{Crystal (geologic formation).}
\tlhf{TKD}

\tlhw{qutluch}
\tlhd{n}{Type of hand weapon.}
\tlhf{TKD}

\tlhw{quv}
\tlhd{n}{Honor.}
\tlhd{v}{Be honored.}
\tlhf{TKDa.}

\tlhw{quvmoH}
\tlhd{v}{Honor.}
\tlhf{TKD}

\tlhw{qu\'}
\tlhd{v}{Be fierce.}
\tlhf{TKD}
