\chapter{\tlhe{r}}

\tlhw{rach}
\tlhd{v}{invigorate, fortify, strengthen}
\tlhf{MSN news group}

\tlhw{rachwI\'}
\tlhd{n}{nurse}
\tlhf{MSN news group}

\tlhw{raD}
\tlhd{v}{Force, compel.}
\tlhf{TKD}

\tlhw{ragh}
\tlhd{v}{Decay, rust, corrode.}
\tlhf{TKD}[TKW][KGT]

\tlhw{rajma\'}
\tlhd{n}{Kidney.}
\tlhf{qep'a' 2011}

\tlhw{ral}
\tlhd{v}{Be violent.}
\tlhf{TKD}

\tlhw{ram}
\tlhd{n}{Night.}
\tlhd{v}{Be trivial, trifling, unimportant.}[Be insignificant.]
\tlhf{TKD}[TKW][KGT]

\tlhw{ramjep}
\tlhd{n}{Midnight.}
\tlhf{TKD}

\tlhw{rap}
\tlhd{v}{Be the same.}
\tlhf{TKD}

\tlhw{raQ}
\tlhd{n}{Camp (military term).}[Manipulate by hand, handle.]
\tlhf{TKD}[KGT]

\tlhw{raQpo\'}
\tlhd{n}{Passenger.}
\tlhf{TKD}

\tlhw{rar}
\tlhd{v}{Connect.}
\tlhf{TKD}

\tlhw{raS}
\tlhd{n}{Table.}
\tlhf{KGT}

\tlhw{raSber nahmey}
\tlhd{pl n}{Raspberries.}
\tlhf{Eurotalk}

\tlhw{raSya\'}
\tlhd{n}{Russia.}
\tlhf{Eurotalk}

\tlhw{ratlh}
\tlhd{n}{Remain.}
\tlhf{TKD}

\tlhw{rav}
\tlhd{n}{Floor.}
\tlhf{TKD}

\tlhw{rav\'eq}
\tlhd{n}{Ceiling.}
\tlhf{NEWS}

\tlhw{raw\'}
\tlhd{n}{An aquatic bird with colorful plumage.}
\tlhf(HolQeD)

\tlhw{ray}
\tlhd{n}{The consonant "r".}
\tlhf{qep\'a\' wa\'maH javDIch 2009}

\tlhw{ray\'}
\tlhd{n}{Targets.}
\tlhf{TKD}

\tlhw{ray\' tIr}
\tlhd{n}{Rice.}
\tlhf{Eurotalk}

\tlhw{ra\'}
\tlhd{v}{Order, command.}
\tlhf{TKD}

\tlhw{ra\'ghomquv}
\tlhd{n}{High Command.}
\tlhf{TKD}[KGT]

\tlhw{ra\'taj}
\tlhd{n}{Type of beverage, coffee with liquor.}
\tlhd{n}{KGT}

\tlhw{ra\'wI\'}
\tlhd{n}{Commander.}
\tlhf{TKD}

\tlhw{rech}
\tlhd{v}{Exhale.}
\tlhf{HolQeD}

\tlhw{reD}
\tlhd{n}{Exterior wall.}
\tlhf{NEWS}


\tlhw{reghuluS}
\tlhd{n}{Regulus.}
\tlhc
\tlhca{HIq qIj reghuluS \'Iw HIq ghap jab.}[They serve Black Ale or Regulan bloodwine.][KGT]
\tlhf{TKD}[CK][KGT]

\tlhw{reghuluSngan}
\tlhd{n}{Regulan.}
\tlhf{TKD}

\tlhw{reghuluS \'Iwghargh}
\tlhd{n}{Reghulus bloodworm.}
\tlhco{A soft, 4-inch long worm. According to Dr Phlox, Regulan bloodworms are excellent for filtering toxins. Phlox has used these harmless worms for years to clean a patient's lymphatic system, but they are difficult to replace.}[KGT]
\tlhc
\tlhca{tun; reghuluS \'Iwghargh rur}{Soft as a Regulan bloodworm.}
\tlhf{KGT}

\tlhw{reH}
\tlhd{adv}{Always.}
\tlhd{v}{Play.}
\tlhf{TKD}

\tlhw{rejmorgh}
\tlhd{n}{Worrywart.}
\tlhf{TKD K-E}

\tlhw{rep}
\tlhd{n}{Hour.}
\tlhf{TKD}

\tlhw{retlh}
\tlhd{n}{Area beside, area next to.}
\tlhf{TKD}

\tlhw{rewbe\'}
\tlhd{n}{Citizen.}
\tlhf{TKD}

\tlhw{rIgh}
\tlhd{v}{Be lame.}
\tlhf{TKD}

\tlhw{rIH}
\tlhd{v}{Energize.}
\tlhf{TKD}

\tlhw{rIHwI\'}
\tlhd{n}{Energizer.}
\tlhf{TKD}

\tlhw{rIn}
\tlhd{v}{Be accomplished, finished.}
\tlhf{TKD}

\tlhw{rIp}
\tlhd{n}{Council, assembly.}
\tlhf{TKD K-E.}

\tlhw{rIQ}
\tlhd{v}{Be injured.}
\tlhf{TKD}

\tlhw{rIQmoH}
\tlhd{v}{Injure.}
\tlhf{TKD}

\tlhw{rIvSo\'}
\tlhd{n}{Embassy.}
\tlhf{TKD}

\tlhw{rIymuS}
\tlhd{n}{Remus.}
\tlhf{TKD}

\tlhw{rI\'}
\tlhd{v}{Hail.}
\tlhf{TKD}

\tlhw{rI\'Se\'}
\tlhd{n}{Hailing frequency.}
\tlhf{TKD}

\tlhw{ro}
\tlhd{n}{Trunk (of body).}
\tlhf{TKD}

\tlhw{roghvaH}
\tlhd{n}{Population.}
\tlhf{TKD}

\tlhw{roj}
\tlhd{n}{Peace.}
\tlhd{v}{Make peace.}
\tlhf{TKD}

\tlhw{rojHom}
\tlhd{n}{Truce.}
\tlhf{TKD}

\tlhw{rojmab}
\tlhd{n}{Peace treaty.}
\tlhf{TKD}

\tlhw{rol}
\tlhd{n}{Beard.}
\tlhf{TKD}

\tlhw{rol taj}
\tlhd{n}{Razor.}
\tlhf{Eurotalk}

\tlhw{rom}
\tlhd{n}{Accord.}
\tlhf{TKD}

\tlhw{romuluS}
\tlhd{n}{Romulus.}
\tlhf{TKD}

\tlhw{romuluSngan}
\tlhd{n}{Romulan.}
\tlhf{TKD}

\tlhw{rop}
\tlhd{n}{Disease.}
\tlhd{v}{Be sick, ill.}
\tlhf{TKD}

\tlhw{ropyaH}
\tlhd{n}{Infirmary.}
\tlhf{TKD}

\tlhw{roQ}
\tlhd{v}{Put down.}
\tlhf{TKD}

\tlhw{ror}
\tlhd{v}{Be fat.}
\tlhf{TKD}

\tlhw{rotlh}
\tlhd{v}{Be tough.}
\tlhf{TKD}

\tlhw{ro\'}
\tlhd{n}{Fist.}
\tlhf{TKD}

\tlhw{ro\'qegh\'Iwchab}
\tlhd{n}{Rokeg blood pie.}
\tlhf{TKD}

\tlhw{ruch}
\tlhd{v}{Proceed, go ahead, do it.}
\tlhf{TKD}

\tlhw{rugh}
\tlhd{n}{Antimatter.}
\tlhf{TKD}

\tlhw{runpI\'}
\tlhd{n}{Teapot.}
\tlhf{Eurotalk}

\tlhw{ruQ}
\tlhd{v}{Control manually, by hand.}
\tlhf{TKD}

\tlhw{rup}
\tlhd{v}{Fine, tax.}
\tlhf{TKD}

\tlhw{rur}
\tlhd{v}{Resemble.}
\tlhf{TKD}

\tlhw{rut}
\tlhd{adv}{Sometimes.}
\tlhf{TKD}

\tlhw{ruv}
\tlhd{n}{Justice.}
\tlhf{TKD}

\tlhw{rura\' pente\'}
\tlhd{n}{Rura Penthe.}
\tlhf{TKD}

\tlhw{ruStay}
\tlhd{n}{Bonding ritual.}
\tlhf{TKD}

\tlhw{ru\'}
\tlhd{v}{Be temporary.}
\tlhf{TKD}