\chapter{\tlhe{S}}

\tlhw{Sab}
\tlhd{v}{Decline, deteriorate.}
\tlhf{TKD}

\tlhw{Sach}
\tlhd{v}{Expand.}
\tlhf{TKD}

\tlhw{SaD}
\tlhd{num}{Thousand.}
\tlhf{TKD}

\tlhw{Sagh}
\tlhd{v}{Be serious.}
\tlhf{TKD}

\tlhw{SaH}
\tlhd{v}{Be present (not absent).}
\tlhf{TKD}

\tlhw{SaH}
\tlhd{v}{Care (about), be concerned (about).}
\tlhf{TKD}

\tlhw{Saj}
\tlhd{n}{Pet.}
\tlhf{TKD}

\tlhw{Sal}
\tlhd{v}{Ascend.}
\tlhf{TKD}

\tlhw{Sam}
\tlhd{v}{Locate, seek and find.}
\tlhf{TKD}

\tlhw{San}
\tlhd{n}{Fate.}
\tlhf{TKD}

\tlhw{SanID}
\tlhd{num}{Thousand.}
\tlhf{TKD}

\tlhw{Sang}
\tlhd{v}{Obliterate.}
\tlhf{TKD}

\tlhw{Sap}
\tlhd{v}{Volunteer.}
\tlhf{TKD}

\tlhw{Saq}
\tlhd{v}{Land.}
\tlhf{TKD}

\tlhw{Saqghom}
\tlhd{n}{Landing party.}
\tlhf{TKD}

\tlhw{SaQ}
\tlhd{v}{Cry.}
\tlhf{TKD}

\tlhw{Sar}
\tlhd{v}{Be varied, various.}
\tlhf{TKD}

\tlhw{Sar}
\tlhd{n}{Variety.}
\tlhf{TKD}

\tlhw{Satlh}
\tlhd{n}{Agriculture.}
\tlhf{TKD}

\tlhw{SaS}
\tlhd{v}{Be horizontal.}
\tlhf{TKD}

\tlhw{Saw}
\tlhd{v}{Marry (husband does this).}
\tlhf{TKD}

\tlhw{Say}
\tlhd{n}{The consonant "S".}
\tlhf{qep\'a\' wa\'maH javDIch 2009}

\tlhw{Say\'}
\tlhd{v}{Be clean.}
\tlhf{TKD}

\tlhw{Say\'moHwI\' tlhagh}
\tlhd{n}{Soap.}
\tlhf{Eurotalk}

\tlhw{Sa\'}
\tlhd{n}{General (rank).}
\tlhf{TKD}

\tlhw{Segh}
\tlhd{n}{Race (type, sort, class).}
\tlhf{TKD}

\tlhw{SeH}
\tlhd{v}{Control.}
\tlhf{TKD}

\tlhw{SeHlaw}
\tlhd{n}{Control panel.}
\tlhf{TKD}

\tlhw{Seng}
\tlhd{v}{Cause trouble.}
\tlhf{TKD}

\tlhw{Seng}
\tlhd{n}{Trouble.}
\tlhf{TKD}

\tlhw{Sep}
\tlhd{v}{Breed.}
\tlhf{TKD}

\tlhw{Sep}
\tlhd{n}{Region.}
\tlhf{TKD}

\tlhw{Seq}
\tlhd{n}{Fault (seismic).}
\tlhf{TKD}

\tlhw{Ser}
\tlhd{n}{Progress.}
\tlhf{TKD}

\tlhw{SermanyuQ}
\tlhd{n}{Sherman's Planet.}
\tlhf{TKD}

\tlhw{SeS}
\tlhd{n}{Steam.}
\tlhf{TKD}

\tlhw{Sev}
\tlhd{n}{Bandage.}[Publication of any kind (print or otherwise).]
\tlhd{v}{Contain (an enemy).}
\tlhf{TKD}[paq’batlh]

\tlhw{Sey}
\tlhd{v}{Be excited.}
\tlhf{TKD}

\tlhw{SeymoH}
\tlhd{v}{Excite.}
\tlhf{TKD}

\tlhw{Se\'}
\tlhd{n}{Frequency (radio).}
\tlhf{TKD}

\tlhw{Se\'tu\'}
\tlhd{n}{Saucepan.}
\tlhf{Eurotalk}

\tlhw{SIbDoH}
\tlhd{n}{Satellite.}
\tlhf{TKD}

\tlhw{SIbI\'}
\tlhd{adv}{Immediately.}
\tlhf{TKD}

\tlhw{SIch}
\tlhd{v}{Reach.}
\tlhf{TKD}

\tlhw{SID}
\tlhd{n}{Patient.}
\tlhf{TKD}

\tlhw{SIgh}
\tlhd{v}{Influence.}
\tlhf{TKD}

\tlhw{SIH}
\tlhd{v}{Bend.}
\tlhf{TKD}

\tlhw{SIj}
\tlhd{v}{Slit.}
\tlhf{TKD}

\tlhw{SIm}
\tlhd{v}{Calculate.}
\tlhf{TKD}

\tlhw{SIp}
\tlhd{n}{Gas.}
\tlhf{TKD}

\tlhw{SIQ}
\tlhd{v}{Endure, bear.}
\tlhf{TKD}

\tlhw{SIS}
\tlhd{v}{Rain.}
\tlhf{TKD}

\tlhw{SIS yoD}
\tlhd{n}{Umbrella.}
\tlhf{Eurotalk}

\tlhw{SIv}
\tlhd{v}{Wonder.}
\tlhf{TKD}

\tlhw{Soch}
\tlhd{num}{Seven.}
\tlhf{TKD}

\tlhw{SochDIch}
\tlhd{num}{Seventh.}
\tlhf{TKD}

\tlhw{SoD}
\tlhd{n}{Flood.}
\tlhf{TKD}

\tlhw{SoD}
\tlhd{v}{Flood.}
\tlhf{TKD}

\tlhw{Sogh}
\tlhd{n}{Lieutenant.}
\tlhf{TKD}

\tlhw{SoH}
\tlhd{pro}{You.}
\tlhf{TKD}

\tlhw{Soj}
\tlhd{n}{Food [carries the idiomatic sense of "matter, concern, affair".]}
\tlhc
\tlhca{tlhIngan Soj \'oH -- not bIyaj}{It is a Klingon matter -- you\'ll never understand.}
\tlhf{HolQeD}[v12n3p9.]

\tlhw{Soj naQmey}
\tlhd{pl n}{Chopsticks.}
\tlhf{Eurotalk}

\tlhw{Sol}
\tlhd{v}{Quarrel.}
\tlhf{TKD}

\tlhw{Som}
\tlhd{n}{Hull.}
\tlhf{TKD}

\tlhw{Somraw}
\tlhd{n}{Muscle.}
\tlhf{TKD}

\tlhw{Son}
\tlhd{v}{Relieve.}
\tlhf{TKD}

\tlhw{SonchIy}
\tlhd{n}{Death ritual (for a leader).}
\tlhf{TKD}

\tlhw{Sop}
\tlhd{v}{Eat.}
\tlhf{TKD}

\tlhw{SoQ}
\tlhd{v}{Be closed, shut.}
\tlhf{TKD}

\tlhw{SoQ}
\tlhd{n}{Speech, lecture, address.}
\tlhf{TKD}

\tlhw{SoQmoH}
\tlhd{v}{Close, shut.}
\tlhf{TKD}

\tlhw{Sor}
\tlhd{n}{Tree.}
\tlhf{TKD}

\tlhw{Sorgh}
\tlhd{v}{Sabotage.}
\tlhf{TKD}

\tlhw{SoS}
\tlhd{n}{Mother.}
\tlhf{TKD}
\tlhco{Within the family, a child usually addresses his or her mother as SoS (Mother) and father as vav (Father), though it is not uncommon for younger children to use the words SoSoy (Mommy) and \tlhe{vavoy} (Daddy). These are the regular words for "mother" and "father" followed by the suffix \tlhe{-oy}, which indicates endearment. Most older children drop the \tlhe{-oy} around the time of their Age of Ascension, though some continue to use it even after that, especially when addressing the parent of the opposite sex [...] Though almost always heard as terms of direct address (as in \tlhe{Sosoy jIghung} ["Mommy, I\'m hungry"]), kinship terms with the suffix \tlhe{-oy} are occasionally used as subjects or objects of sentences, particularly in the speech of younger children. For example, a proud child may say, \tlhe{SuvwI\' ghaH vavoy\'e\'} ("My daddy is a warrior").}{KGT}{198-99}
\tlhc
\tlhca{Hab SoSlI\' Quch}{Your mother has a smooth forehead!!}[PK]

\tlhca{SoSwI\' vavwI\' je quvmoHjaj paqvam}{Okrand\'s dedication}[KGT]
\tlhf{KGT}

\tlhw{SoSnI\'}
\tlhd{n}{Grandmother.}
\tlhf{TKD}

\tlhw{Sot}
\tlhd{v}{Be distressed, be in distress.}
\tlhf{TKD}

\tlhw{Sotlaw\'}
\tlhd{n}{Distress call.}
\tlhf{TKD}

\tlhw{Sov}
\tlhd{v}{Know.}
\tlhf{TKD}

\tlhw{Sov}
\tlhd{n}{Knowledge.}
\tlhf{TKD}

\tlhw{Soy\'}
\tlhd{v}{Be clumsy.}
\tlhf{TKD}

\tlhw{So\'}
\tlhd{v}{Hide, cloak.}
\tlhf{TKD}

\tlhw{So\'wI\'}
\tlhd{n}{Cloaking device.}
\tlhf{TKD}

\tlhw{Sub}
\tlhd{v}{Be solid.}
\tlhf{TKD}

\tlhw{Such}
\tlhd{v}{Visit.}
\tlhf{TKD}

\tlhw{SuD}
\tlhd{v}{Be green, blue, yellow.}[Gamble, take a chance, take a risk.]
\tlhf{TKD}

\tlhw{SuD \'ej Hurgh}
\tlhd{v}{Be dark blue.}
\tlhf{Eurotalk}

\tlhw{SuD \'ej wov}
\tlhd{v}{Be light blue.}
\tlhf{Eurotalk}

\tlhw{Sugh}
\tlhd{v}{Install (in office).}
\tlhf{TKD}

\tlhw{SuH}
\tlhd{excl}{Ready, standing by.}
\tlhf{TKD}

\tlhw{Suj}
\tlhd{v}{Disturb.}
\tlhf{TKD}

\tlhw{Sun}
\tlhd{n}{Discipline.}
\tlhf{TKD}

\tlhw{Sung}
\tlhd{n}{Native.}
\tlhf{TKD}

\tlhw{Sup}
\tlhd{v}{Jump.}
\tlhf{TKD}

\tlhw{Sup}
\tlhd{n}{Resource.}
\tlhf{TKD}

\tlhw{Suq}
\tlhd{v}{Acquire, obtain, get.}
\tlhf{TKD}

\tlhw{SuQ}
\tlhd{v}{Be toxic.}
\tlhf{TKD}

\tlhw{Surchem}
\tlhd{n}{Force field.}
\tlhf{TKD}

\tlhw{Surgh}
\tlhd{v}{Skin.}
\tlhf{TKD}

\tlhw{SuS}
\tlhd{n}{Wind, breeze.}
\tlhf{TKD}

\tlhw{Sut}
\tlhd{n}{Clothing.}
\tlhf{TKD}

\tlhw{Sut HabmoHwI\'}
\tlhd{n}{Iron (clothing).}
\tlhf{Eurotalk}

\tlhw{Sutra\'ber naHmey}
\tlhd{pl n}{Strawberries.}
\tlhf{Eurotalk}

\tlhw{Sutlh}
\tlhd{v}{Negotiate.}
\tlhf{TKD}

\tlhw{Suv}
\tlhd{v}{Fight.}
\tlhf{TKD}

\tlhw{Suverya\'}
\tlhd{n}{Sweden.}
\tlhf{Eurotalk}

\tlhw{SuvwI\'}
\tlhd{n}{Warrior.}
\tlhf{TKD}

\tlhw{Suwomly}
\tlhd{n}{Finland.}
\tlhf{Eurotalk}

\tlhw{Suy}
\tlhd{n}{Merchant.}
\tlhf{TKD}

\tlhw{SuyDuj}
\tlhd{n}{Merchant ship.}
\tlhf{TKD}

\tlhw{Su\'}
\tlhd{excl}{Ready, standing by.}
\tlhf{TKD}

\tlhw{Su\'ghar qutmey}
\tlhd{pl n}{Sugar.}
\tlhf{Eurotalk}