\chapter{\tlhe{t}}

\tlhw{ta}
\tlhd{n}{Record.}
\tlhf{TKD}

\tlhw{tach}
\tlhd{n}{Bar, saloon, cocktail lounge.}
\tlhf{TKD}

\tlhw{taD}
\tlhd{v}{Be frozen.}
\tlhf{TKD}

\tlhw{taDmoH}
\tlhd{v}{Freeze.}
\tlhf{TKD}

\tlhw{tagh}
\tlhd{n}{Lung.}
\tlhf{TKD}

\tlhw{tagh}
\tlhd{v}{Begin a process, initiate.}
\tlhf{TKD}

\tlhw{taH}
\tlhd{v}{To be at a negative angle. This is referring to a ship, in space, that has rotated its nose downwards.}[To continue, go on, endure.]
\tlhf{TKD}[TKDa.]

\tlhw{taj}
\tlhd{n}{Knife, dagger.}
\tlhf{TKD}

\tlhw{tal}
\tlhd{n}{Cannon.}
\tlhf{BoP Poster.}

\tlhw{tam}
\tlhd{v}{Exchange.}[Substitute.][Be quiet.]
\tlhco{By the same token, literally translating a sentence such as "The restaurant changed hands" into Klingon, presumably as \tlhe{ghopDu\' choHpu\' Qe\'} ("The restaurant has altered hands") or perhaps \tlhe{ghopDu\' tampu\' Qe\'} ("The restaurant has exchanged hands"), does not tell the Klingon that the restaurant is now under new ownership, which is what the Federation Standard phrase really means.}{KGT}{106-7}
\tlhco{An altered form of this expression substitutes \tlhe{bo\'DaghHom} ("little scoop") for \tlhe{bo\'Dagh\'a\'} ("big scoop").}{KGT}{125}
\tlhco{Mispronunciation may turn dangerous, on the other hand, if one Klingon sound is substituted for another.}{KGT}{188}
\tlhc
\tlhca{boq lucherDI\' tlhIngan wo\' romuluS Hov wo\' je So\'wI\' cham Suqpu\' tlhIngan wo\'. chaq tampu\'. chaq romuluSnganpu\'vaD pIvghor cham lunobpu\' tlhInganpu\'.}{Cloaking technology was gained through an alliance with the Romulan Star Empire, possibly in exchange for Warp Drive technology.}[S33]
\tlhca{ghopDu\' tampu\' Qe\'}{The restaurant has exchanged hands.}[S33]
\tlhcf{lIw}[qa'meH][choH][nob]
\tlhf{TKD}

\tlhw{tammoH}
\tlhd{v}{Silence.}
\tlhf{TKD}

\tlhw{taQ}
\tlhd{v}{Be weird.}
\tlhf{TKD}

\tlhw{taQbang}
\tlhd{n}{exhaust.}
\tlhf{TKD}

\tlhw{tar}
\tlhd{n}{Poison.}
\tlhf{TKD}

\tlhw{targh}
\tlhd{n}{Targ.}
\tlhf{TKD}

\tlhw{taS}
\tlhd{n}{Solution (liquid).}
\tlhf{TKD}

\tlhw{tat}
\tlhd{n}{Ion.}
\tlhf{TKD}

\tlhw{tay}
\tlhd{n}{Ceremony, rite, ritual.}[The consonant "t".]
\tlhd{v}{Be civilized.}
\tlhf{TKD}[qep\'a\' wa\'maH javDIch 2009]

\tlhw{taymoH}
\tlhd{v}{Civilize.}
\tlhf{TKD}

\tlhw{tayqeq}
\tlhd{n}{Civilization.}
\tlhf{TKD}

\tlhw{tay\'}
\tlhd{v}{Be together.}
\tlhf{TKD}

\tlhw{ta\'}
\tlhd{n}{Emperor.}[Accomplishment.]
\tlhd{v}{Accomplish.}
\tlhf{TKD}

\tlhw{teb}
\tlhd{v}{Fill.}
\tlhf{TKD}

\tlhw{teblaw\'}
\tlhd{n}{Jurisdiction.}
\tlhf{TKD}

\tlhw{teghbat}
\tlhd{n}{teg'bat. A popular effervescent drink, black in color, is called \tlhe{\'awje\'}. This is frequently, though not accurately, translated as "rootbeer," probably because of its superficial resemblance to the Federation soft drink, but it is made from, among other things, the marrow ({melchoQ}) taken from the bones of a \tlhe{teghbat} (teg\'bat).}
\tlhf{KGT}

\tlhw{teH}
\tlhd{v}{Be true.}
\tlhf{TKD}

\tlhw{tej}
\tlhd{n}{Scientist.}
\tlhf{TKD}

\tlhw{tel}
\tlhd{n}{Wing.}
\tlhf{TKD}

\tlhw{telun Hovtay\'}
\tlhd{n}{Tellun Star System.}
\tlhf{TKD}

\tlhw{tem}
\tlhd{v}{Deny.}
\tlhf{TKD}

\tlhw{ten}
\tlhd{v}{Embark.}
\tlhf{TKD}

\tlhw{tengchaH}
\tlhd{n}{Space station.}
\tlhf{TKD}

\tlhw{tep}
\tlhd{n}{Cargo.}
\tlhf{TKD}

\tlhw{tepqengwI\'}
\tlhd{n}{Cargo carrier.}
\tlhf{TKD}

\tlhw{teq}
\tlhd{v}{Remove, take off.}
\tlhf{TKD}

\tlhw{tera\'}
\tlhd{n}{Earth.}
\tlhf{TKD}

\tlhw{tera\' na\'ran}
\tlhd{n}{Orange.}
\tlhf{Eurotalk}

\tlhw{tera\' na\'ran wIb}
\tlhd{n}{Lemon.}
\tlhf{Eurotalk}

\tlhw{tera\' na\'ran\'a\'}
\tlhd{n}{Grapefruit.}
\tlhf{Eurotalk}

\tlhw{tera\' peb\'ot}
\tlhd{n}{Cucumber.}
\tlhf{Eurotalk}

\tlhw{tera\' yav \'atlhqammey}
\tlhd{pl n}{Mushrooms.}
\tlhf{Eurotalk}

\tlhw{tera\'ngan}
\tlhd{n}{Terran, Earther.}
\tlhf{TKD}

\tlhw{tet}
\tlhd{v}{Melt.}
\tlhf{TKD}

\tlhw{tev}
\tlhd{n}{Prize.}
\tlhf{TKD}

\tlhw{tey\'}
\tlhd{v}{Confide.}
\tlhf{TKD}

\tlhw{tI}
\tlhd{n}{Vegetation.}
\tlhf{TKD}

\tlhw{tIch}
\tlhd{v}{Insult.}
\tlhf{TKD}

\tlhw{tIgh}
\tlhd{n}{Custom.}
\tlhf{TKD}

\tlhw{tIH}
\tlhd{n}{Ray.}
\tlhf{TKD}

\tlhw{tIj}
\tlhd{v}{Board, go aboard.}
\tlhf{TKD}

\tlhw{tIn}
\tlhd{v}{Be big.}
\tlhf{TKD}

\tlhw{tIq}
\tlhd{n}{Heart.}
\tlhd{v}{Be long, lengthy (of an object).}
\tlhf{TKD}

\tlhw{tIQ}
\tlhd{v}{Be ancient.}
\tlhf{TKD}

\tlhw{tIr}
\tlhd{n}{Grain.}
\tlhf{TKD}

\tlhw{tIr ngogh}
\tlhd{n}{Bread.}
\tlhf{Eurotalk}

\tlhw{tIr ngogh QaD}
\tlhd{n}{Toast.}
\tlhf{Eurotalk}

\tlhw{tIS}
\tlhd{v}{Be light (weight).}
\tlhf{TKD}

\tlhw{tIv}
\tlhd{v}{Enjoy.}
\tlhf{TKD}

\tlhw{tI\'}
\tlhd{v}{Fix, repair.}
\tlhf{TKD}

\tlhw{tob}
\tlhd{v}{Prove.}
\tlhf{TKD}

\tlhw{tob}
\tlhd{v}{Test conclusively.}[Prove.]
\tlhf{TKD}

\tlhw{toch}
\tlhd{n}{Palm (of hand).}
\tlhf{TKD}

\tlhw{toD}
\tlhd{v}{Save, rescue.}
\tlhf{TKD}

\tlhw{toD}
\tlhd{n}{Rescue.}
\tlhf{TKD}

\tlhw{toDuj}
\tlhd{n}{Courage, bravery.}
\tlhf{TKD}

\tlhw{togh}
\tlhd{v}{Count.}
\tlhco{The suffix \tlhe{-logh}, when attached to numbers, is used to count the number of instances of something: \tlhe{wa\'logh} ("once"), \tlhe{cha\'logh} ("twice"), \tlhe{vaghlogh} ("five times")".}{KGT}{178}
\tlhc
\tlhe{cha\' vI\' vagh choQmey lutoghlu\'}{Number of Decks: 2.5}
\tlhcf{SIm}[chel]
\tlhf{TKD}[KGT][BoP]

\tlhw{toH}
\tlhd{excl}{So, well.}
\tlhf{TKD}

\tlhw{toj}
\tlhd{v}{Deceive, trick.}
\tlhf{TKD}

\tlhw{tongDuj}
\tlhd{n}{Freighter.}
\tlhf{TKD}

\tlhw{tomat naHmey}
\tlhd{pl n}{Tomatoes.}
\tlhf{Eurotalk}

\tlhw{toplIn}
\tlhd{n}{Topaline.}
\tlhf{TKD}

\tlhw{toq}
\tlhd{v}{Be inhabited.}
\tlhf{TKD}

\tlhw{toQDuj}
\tlhd{n}{Bird of Prey (vessel).}
\tlhf{TKD}

\tlhw{tor}
\tlhd{v}{Kneel.}
\tlhf{TKD}

\tlhw{toS}
\tlhd{v}{Climb.}
\tlhf{TKD}

\tlhw{totlh}
\tlhd{n}{Commodore.}
\tlhf{TKD}

\tlhw{toy\'}
\tlhd{v}{Serve (a master).}
\tlhf{TKD}

\tlhw{toy\'wI\'}
\tlhd{n}{Servant.}
\tlhf{TKD}

\tlhw{toy\'wI\'\'a\'}
\tlhd{n}{Slave.}
\tlhf{TKD}

\tlhw{to\'}
\tlhd{n}{Tactics.}
\tlhf{TKD}

\tlhw{tuch}
\tlhd{v}{Forbid.}
\tlhf{TKD}

\tlhw{tugh}
\tlhd{adv}{Soon.}
\tlhf{TKD}

\tlhw{tuH}
\tlhd{v}{Be ashamed.}
\tlhf{TKD}

\tlhw{tuH}
\tlhd{n}{Maneuver (military term).}
\tlhf{TKD}

\tlhw{tuHmoH}
\tlhd{v}{Shame.}
\tlhf{TKD}

\tlhw{tuj}
\tlhd{v}{Be hot.}
\tlhf{TKD}

\tlhw{tuj}
\tlhd{n}{Heat.}
\tlhf{TKD}

\tlhw{tul}
\tlhd{v}{Hope.}
\tlhf{TKD}

\tlhw{tum}
\tlhd{n}{Agency.}
\tlhf{TKD}

\tlhw{tun}
\tlhd{v}{Be soft.}
\tlhf{TKD}

\tlhw{tung}
\tlhd{v}{Discourage.}
\tlhf{TKD}

\tlhw{tungHa\'}
\tlhd{v}{Encourage.}
\tlhf{TKD}

\tlhw{tup}
\tlhd{n}{Minute (of time).}
\tlhf{TKD}

\tlhw{tuqjIjQa\'}
\tlhd{n}{United Kingdom.}
\tlhf{Eurotalk}

\tlhw{tuQ}
\tlhd{v}{Wear (clothes).}
\tlhf{TKD}

\tlhw{tuQDoq}
\tlhd{n}{Mind sifter (Klingon psychic probe).}
\tlhf{TKD}

\tlhw{tuQHa\'moH}
\tlhd{v}{Undress.}
\tlhf{TKD}

\tlhw{tuQmoH}
\tlhd{v}{Put on (clothes).}
\tlhf{TKD}

\tlhw{turIqya\'}
\tlhd{n}{Turkey.}
\tlhf{Eurotalk}

\tlhw{tut}
\tlhd{n}{Column.}
\tlhf{TKD}

\tlhw{tuv}
\tlhd{v}{Be patient.}
\tlhf{TKD}

\tlhw{tu\'}
\tlhd{v}{Discover, find, observe, notice.}
\tlhf{TKD}