\chapter{\tlhe{v}}

\tlhw{vagh}
\tlhd{num}{Five.}
\tlhf{TKD}

\tlhw{vaghDIch}
\tlhd{num}{Fifth.}
\tlhf{TKD}

\tlhw{vaH}
\tlhd{n}{Holster.}
\tlhf{TKD}

\tlhw{vaj adv,n.adv. 1. So, then, thus, in that case.}
\tlhd{n}{Warrior.}
\tlhf{TKD}[TKDa.]

\tlhw{val}
\tlhd{v}{Be clever, smart, intelligent.}
\tlhf{TKD}

\tlhw{van}
\tlhd{v}{Salute.}
\tlhf{TKD}

\tlhw{van}
\tlhd{n}{Salute, tribute.}
\tlhf{TKD}

\tlhw{van\'a\'}
\tlhd{n}{Award.}
\tlhf{TKD}

\tlhw{vang}
\tlhd{v}{Act, take action.}
\tlhf{TKD}

\tlhw{vaq}
\tlhd{v}{Mock.}
\tlhf{TKD}

\tlhw{vaQ}
\tlhd{v}{Be aggressive.}
\tlhf{TKD}

\tlhw{vaS}
\tlhd{n}{Hall, assembly hall.}
\tlhf{TKD}

\tlhw{vaS\'a\'}
\tlhd{n}{Great Hall.}
\tlhf{TKD}

\tlhw{vatlh}
\tlhd{num}{Hundred.}
\tlhf{TKD}

\tlhw{vatlhvI\'}
\tlhd{n}{Percent.}
\tlhf{TKD}

\tlhw{vav}
\tlhd{n}{Father.}
\tlhf{TKD}

\tlhw{vavnI\'}
\tlhd{n}{Grandfather.}
\tlhf{TKD}

\tlhw{vay}
\tlhd{n}{The consonant "v".}
\tlhf{qep\'a\' wa\'maH javDIch 2009}

\tlhw{vay}
\tlhd{v}{Mid-level ferocity battle.}

\tlhw{Cf. lul, Hargh, Qor, tlhaS.}
\tlhf{KGT}

\tlhw{vay\'}
\tlhd{n}{Somebody, something, anybody, anything.}
\tlhf{TKD}

\tlhw{veH}
\tlhd{n}{Boundary.}
\tlhf{TKD}

\tlhw{veH tIn}
\tlhd{n}{Great Barrier.}
\tlhf{TKD}

\tlhw{vem}
\tlhd{v}{Wake up, cease sleeping.}
\tlhf{TKD}

\tlhw{vemmoH}
\tlhd{v}{Wake (someone) up.}
\tlhf{TKD}

\tlhw{veng}
\tlhd{n}{City.}
\tlhf{TKD}

\tlhw{vengHom}
\tlhd{n}{Village.}
\tlhf{TKD}

\tlhw{veqlargh}
\tlhd{n}{Devil, demon, Fek'lhr.}
\tlhf{TKD}

\tlhw{veQ}
\tlhd{n}{Garbage.}
\tlhf{TKD}

\tlhw{veQDuj}
\tlhd{n}{Garbage scow.}
\tlhf{TKD}

\tlhw{verengan}
\tlhd{n}{Ferengi.}
\tlhf{TKD}

\tlhw{vergh}
\tlhd{n}{Dock.}
\tlhf{TKD}

\tlhw{vergh}
\tlhd{v}{Dock.}
\tlhf{TKD}

\tlhw{veS}
\tlhd{n}{War.}
\tlhf{TKD}

\tlhw{veSDuj}
\tlhd{n}{Warship.}
\tlhf{TKD}

\tlhw{vetlh}
\tlhd{n}{Cockroach.}
\tlhf{TKD}

\tlhw{vID}
\tlhd{v}{Be belligerent.}
\tlhf{TKD}

\tlhw{vIH}
\tlhd{v}{Move, be in motion.}
\tlhf{TKD}

\tlhw{vIj}
\tlhd{n}{Thruster.}
\tlhf{TKD}

\tlhw{vIlle\'}
\tlhd{n}{Minion.}
\tlhf{qep\'a\' wa\'maH javDIch 2009}

\tlhw{vIng}
\tlhd{v}{Whine.}
\tlhf{TKD}

\tlhw{vIraS}
\tlhd{n}{France.}
\tlhf{Eurotalk}

\tlhw{vIt}
\tlhd{v}{Tell the truth.}
\tlhf{TKD}

\tlhw{vIt}
\tlhd{n}{Truth.}
\tlhf{TKD}

\tlhw{vI\'}
\tlhd{n}{Sharpshooting, marksmanship.}
\tlhf{TKD}

\tlhw{vI\'}
\tlhd{v}{Accumulate.}
\tlhf{TKD}

\tlhw{voDleH}
\tlhd{n}{Emperor.}
\tlhf{TKD}

\tlhw{vogh}
\tlhd{n}{Somewhere.}
\tlhf{TKD}

\tlhw{voHDajbo\'}
\tlhd{n}{Ransom.}
\tlhf{TKD}

\tlhw{volchaH}
\tlhd{n}{Shoulder.}
\tlhf{TKD}

\tlhw{vong}
\tlhd{v}{Hypnotize.}
\tlhf{TKD}

\tlhw{voq}
\tlhd{v}{Trust, have faith in.}
\tlhf{TKD}

\tlhw{voqHa\'}
\tlhd{v}{Distrust.}
\tlhf{TKD}

\tlhw{voQ}
\tlhd{v}{Choke.}
\tlhf{TKD}

\tlhw{vor}
\tlhd{v}{Cure.}
\tlhf{TKD}

\tlhw{vo\'}
\tlhd{v}{Propel.}
\tlhf{TKD}

\tlhw{vub}
\tlhd{n}{Hostage.}
\tlhc
\tlhe{vubpu\' jonta\' HeSwI\'}{A criminal has succeeded in taking hostages.}
\tlhe{wa\' vub ghaH tlhIngan gharwI\'\'e\'.}{One hostage is a Klingon diplomat.}
\tlhe{tlhIngan ghaH wa\' vub\'e\'.}{One of the hostages is a Klingon.}
\tlhe{vubpu\' jon nuchpu\'. jonbe\' tlhInganpu\'.}{Cowards take hostages. Klingons do not.}
\tlhcf{qama'}[voHDajbo'][quch]
\tlhf{TKD}[TKW][ST5]

\tlhw{vuD}
\tlhd{n}{Opinion.}
\tlhf{TKD}

\tlhw{vul}
\tlhd{v}{Be unconscious.}
\tlhf{TKD}

\tlhw{vulqan}
\tlhd{n}{Vulcan (planet).}
\tlhf{TKD}

\tlhw{vulqangan}
\tlhd{n}{Vulcan (person).}
\tlhf{TKD}

\tlhw{vum}
\tlhd{n}{Work, toil.}
\tlhf{TKD}

\tlhw{vup}
\tlhd{v}{Pity.}
\tlhf{TKD}

\tlhw{vuQ}
\tlhd{v}{Fascinate.}
\tlhf{TKD}

\tlhw{vuS}
\tlhd{v}{Limit.}
\tlhf{TKD}

\tlhw{vut}
\tlhd{v}{Cook.}
\tlhf{TKD}

\tlhw{vutpa\'}
\tlhd{n}{Galley.}
\tlhf{TKD}

\tlhw{vuv}
\tlhd{v}{Respect.}
\tlhf{TKD}

\tlhw{vu\'}
\tlhd{v}{Manage.}
\tlhf{TKD}

\tlhw{vu\'wI\'}
\tlhd{n}{Manager.}
\tlhf{TKD}