\chapter{\tlhe{w}}

\tlhw{wab HevwI\'}
\tlhd{n}{Radio.}
\tlhf{Eurotalk}

\tlhw{wab qosta\' \'aplo\'}
\tlhd{n}{Cassette.}
\tlhf{Eurotalk}

\tlhw{wam}
\tlhd{v}{Hunt.}
\tlhf{TKD}

\tlhw{wanI\'}
\tlhd{n}{Phenomenon, event, occurrence.}
\tlhf{TKD}

\tlhw{waq}
\tlhd{n}{Shoe.}
\tlhf{TKD}

\tlhw{waqmey}
\tlhd{pl n}{Shoes.}
\tlhf{Eurotalk}

\tlhw{waQ}
\tlhd{v}{Obstruct.}
\tlhf{TKD}

\tlhw{watlh}
\tlhd{v}{Be pure.}
\tlhf{TKD}

\tlhw{wav}
\tlhd{v}{Divide.}
\tlhf{TKD}

\tlhw{waw\'}
\tlhd{n}{Base (military term).}
\tlhf{TKD}

\tlhw{way}
\tlhd{n}{The consonant "w".}
\tlhf{qep\'a\' wa\'maH javDIch 2009}

\tlhw{wa\'}
\tlhd{num}{One (num)}
\tlhf{TKD}

\tlhw{wa\'DIch}
\tlhd{num}{First.}
\tlhf{TKD}

\tlhw{wa\'Hu\'}
\tlhd{n}{Yesterday.}
\tlhf{TKD}

\tlhw{wa\'leS}
\tlhd{n}{Tomorrow.}
\tlhf{TKD}

\tlhw{wa\'logh}
\tlhd{adv}{Once.}
\tlhf{TKD}

\tlhw{wa\'maH}
\tlhd{num}{Ten.}
\tlhf{TKD}

\tlhw{wa\'maHDIch}
\tlhd{num}{Tenth.}
\tlhf{TKD}

\tlhw{web}
\tlhd{v}{Be disgraced.}
\tlhf{TKD}

\tlhw{wegh}
\tlhd{v}{Confine.}
\tlhf{TKD}

\tlhw{weH}
\tlhd{v}{Raid.}
\tlhf{TKD}

\tlhw{wej}
\tlhd{adv}{Not yet.}
\tlhf{TKD}

\tlhw{wej}
\tlhd{num}{Three.}
\tlhf{TKD}

\tlhw{wejDIch}
\tlhd{num}{Third.}
\tlhf{TKD}

\tlhw{wejpuH}
\tlhd{excl}{Charming (used only ironically).}
\tlhf{TKD}

\tlhw{wel}
\tlhd{v}{Owe (someone).}
\tlhco{When someone has earned a reward but the prize is not yet available to be given, the person in charge of presenting it can say \tlhe{qawel} "I owe you." It is probably also appropriate when you have convinced someone to give you some service or product with the understanding that you will pay for it later.}{ghunchu\'wI\' \'utlh}
\tlhf{qep\'a\' wa\'maH chorghDIch 2011}

\tlhw{wem}
\tlhd{v}{Violate.}
\tlhf{TKD}

\tlhw{wem}
\tlhd{n}{Violation.}
\tlhf{TKD}

\tlhw{wep}
\tlhd{n}{Jacket, coat.}
\tlhf{TKD}

\tlhw{weQ}
\tlhd{n}{Candle.}
\tlhf{TKD}

\tlhw{wew}
\tlhd{v}{Glow.}
\tlhf{TKD}

\tlhw{wev}
\tlhd{v}{Owe (someone).}
\tlhco{When someone has earned a reward but the prize is not yet available to be given, the person in charge of presenting it can say \tlhe{qawel} "I owe you." It is probably also appropriate when you have convinced someone to give you some service or product with the understanding that you will pay for it later.}{ghunchu\'wI\' \'utlh}
\tlhf{qep\'a\' wa\'maH chorghDIch 2011}

\tlhw{wIb}
\tlhd{v}{Be sour.}
\tlhf{TKD}

\tlhw{wIch}
\tlhd{n}{Myth.}
\tlhf{TKD}

\tlhw{wIgh}
\tlhd{n}{Genius.}
\tlhf{TKD}

\tlhw{wIH}
\tlhd{v}{Be ruthless.}
\tlhf{TKD}

\tlhw{wIj}
\tlhd{v}{Farm.}
\tlhf{TKD}

\tlhw{wIv}
\tlhd{n}{Choice.}
\tlhf{TKD}

\tlhw{wIv}
\tlhd{v}{Choose, select.}
\tlhf{TKD}

\tlhw{wIy}
\tlhd{n}{Tactical display.}
\tlhf{TKD}

\tlhw{woD}
\tlhd{v}{Throw away.}
\tlhf{TKD}

\tlhw{woH}
\tlhd{v}{Pick up.}
\tlhf{TKD}

\tlhw{woj}
\tlhd{n}{Radiation.}
\tlhf{TKD}

\tlhw{woj choHwI\'}
\tlhd{n}{Reactor.}
\tlhf{TKD}

\tlhw{woQ}
\tlhd{n}{Authority, political power.}
\tlhf{TKD}

\tlhw{woS}
\tlhd{n}{Chin.}
\tlhf{TKD}

\tlhw{wot}
\tlhd{n}{Verb.}
\tlhf{TKD}

\tlhw{wov}
\tlhd{v}{Be light, bright.}
\tlhf{TKD}

\tlhw{wovmoHwI\' moQ}
\tlhd{n}{Lightbulb.}
\tlhf{Eurotalk}

\tlhw{wo\'}
\tlhd{n}{Empire.}
\tlhf{TKD}

\tlhw{wuq}
\tlhd{v}{Decide.}
\tlhf{TKD}

\tlhw{wuq}
\tlhd{v}{Decide upon.}
\tlhf{TKD}

\tlhw{wuQ}
\tlhd{v}{Have a headache.}
\tlhf{TKD}

\tlhw{wuS}
\tlhd{n}{Lip.}
\tlhf{TKD}

\tlhw{wuS rItlh naQ}
\tlhd{n}{Lipstick.}
\tlhf{Eurotalk}

\tlhw{wutlh}
\tlhd{n}{Underground.}
\tlhf{TKD}

\tlhw{wuv}
\tlhd{v}{Depend on, rely on.}
\tlhf{TKD}