\chapter{\tlhe{tlh}}

\tlhw{tlhab}
\tlhd{v}{Be free, independent.}
\tlhf{TKD}

\tlhw{tlhab}
\tlhd{n}{Freedom, independence.}
\tlhf{TKD}

\tlhw{tlhagh patat \'oQqar naQHommey}
\tlhd{pl n}{Fries.}
\tlhf{Eurotalk}

\tlhw{tlham}
\tlhd{n}{Gravity.}
\tlhf{TKD}

\tlhw{tlhap}
\tlhd{v}{Take.}
\tlhf{TKD}

\tlhw{tlhaq}
\tlhd{n}{Chronometer, clock.}
\tlhf{TKD}[Eurotalk]

\tlhw{tlhaQ}
\tlhd{v}{Be funny.}
\tlhf{TKD}

\tlhw{tlhaS}
\tlhd{v}{Relatively minor fight.}[Battle.]
\tlhcf{lul}[Hargh][Qor][vay]
\tlhf{KGT}

\tlhw{tlhay}
\tlhd{n}{The consonant "tlh".}
\tlhf{qep\'a\' wa\'maH javDIch 2009}

\tlhw{tlha\'}
\tlhd{v}{Chase, follow.}
\tlhf{TKD}

\tlhw{tlheD}
\tlhd{v}{Depart.}
\tlhf{TKD}

\tlhw{tlhegh}
\tlhd{n}{Line, rope.}
\tlhf{TKD}

\tlhw{tlhej}
\tlhd{v}{Accompany.}
\tlhf{TKD}

\tlhw{tlhetlh}
\tlhd{v}{progress.}
\tlhf{TKD}

\tlhw{tlhe\'}
\tlhd{v}{Turn.}
\tlhf{TKD}

\tlhw{tlhIb}
\tlhd{v}{Be incompetent.}
\tlhf{TKD}

\tlhw{tlhIch}
\tlhd{n}{Smoke.}
\tlhf{TKD}

\tlhw{tlhIH}
\tlhd{pro}{You (plural).}
\tlhf{TKD}

\tlhw{tlhIj}
\tlhd{v}{Apologize.}
\tlhf{TKD}

\tlhw{tlhIl}
\tlhd{n}{Mineral.}
\tlhd{v}{Mine.}
\tlhf{TKD}

\tlhw{tlhIlHal}
\tlhd{n}{Mine.}
\tlhf{TKD}

\tlhw{tlhIlwI\'}
\tlhd{n}{Miner.}
\tlhf{TKD}

\tlhw{tlhIngan}
\tlhd{n}{Klingon.}
\tlhc
\tlhca{tlhIngan Hol Dajatlh\'a\'}{Do you speak Klingon?}
\tlhca{tlhIngan Hol vIjatlhlaHbe\'}{I cannot speak Klingon.}
\tlhf{TKD}

\tlhw{tlhIngan Hubbeq}
\tlhd{n}{Klingon Defense Force.}
\tlhf{TKD}

\tlhw{tlhIngan wo\'}
\tlhd{n}{Klingon Empire.}
\tlhf{TKD}

\tlhw{tlhIv}
\tlhd{v}{Be insubordinate.}
\tlhf{TKD}

\tlhw{tlhob}
\tlhd{v}{Ask.}
\tlhf{TKD}

\tlhw{tlhob}
\tlhd{v}{Request, ask, plead.}
\tlhf{TKD}

\tlhw{tlhoch}
\tlhd{v}{Contradict.}
\tlhf{TKD}

\tlhw{tlhogh}
\tlhd{n}{Marriage.}
\tlhf{TKD}

\tlhw{tlhoj}
\tlhd{v}{Realize.}
\tlhf{TKD}

\tlhw{tlhol}
\tlhd{v}{Be raw, unprocessed.}
\tlhf{TKD}

\tlhw{tlhon}
\tlhd{n}{Nostril.}
\tlhf{TKD}

\tlhw{tlhong}
\tlhd{v}{Barter, bargain.}
\tlhf{TKD}

\tlhw{tlhoQ}
\tlhd{n}{Conglomeration.}
\tlhf{TKD}

\tlhw{tlhov}
\tlhd{v}{Wheeze.}
\tlhf{TKD}

\tlhw{tlho\'}
\tlhd{n}{Appreciation.}[Gratitude.]
\tlhd{v}{Thank.}
\tlhco{At qep'a' loSDIch Okrand was puzzled by a comment someone made and said something like, "There's a word for 'Thank you'?" Someone told him \tlhe{qatlho\'}, and he said something along the lines of, "I never intended it to mean "thanks" and elaborated a little on the circumstances it would be used in.}{Qov}
\tlhco{When Marc Okrand signed my TKD I think I said \tlhe{qatlho\'neS}, which was probably a bit effusive, but hey, I was excited. He replied in English something close to, \'Well, there\'s NO Klingon response to that\'.}{Qov}
\tlhco{When Picard told Amb. Kell "{\#K'adlo\#} Thank you," Kell gave him an odd look, then nodded (or grunted) and left the room.}{TNG "The Mind's Eye"}
\tlhc
\tlhca{chotlho\'}{You thank me.}{STC:KLS}
\tlhca{qatlho\'}{Thank you!}{STE}
\tlhf{TKDa.}

\tlhw{tlhuch}
\tlhd{v}{Exhaust.}
\tlhf{TKD}

\tlhw{tlhuH}
\tlhd{n}{Breath.}
\tlhd{v}{Breathe.}
\tlhf{TKD}

\tlhw{tlhup}
\tlhd{v}{Whisper.}
\tlhf{TKD}

\tlhw{tlhutlh}
\tlhd{v}{Drink.}
\tlhf{TKD}

\tlhw{tlhu\'}
\tlhd{v}{Be tempted.}
\tlhf{TKD}

\tlhw{tlhu\'moH}
\tlhd{v}{Tempt.}
\tlhf{TKD}