\chapter{\tlhe{y}}

\tlhw{ya}
\tlhd{n}{Tactical officer.}
\tlhf{TKD}

\tlhw{yab}
\tlhd{n}{Mind, brain.}
\tlhf{TKD}

\tlhw{yach}
\tlhd{v}{Pet, stroke.}
\tlhf{TKD}

\tlhw{yaD}
\tlhd{n}{Toe.}
\tlhf{TKD}

\tlhw{yagh}
\tlhd{n}{Organism.}
\tlhf{TKD}

\tlhw{yaH}
\tlhd{n}{Duty station, station.}
\tlhf{TKD}

\tlhw{yaH}
\tlhd{v}{Be taken away.}
\tlhf{TKD}

\tlhw{yaj}
\tlhd{v}{Understand.}
\tlhf{TKD}

\tlhw{yajHa\'}
\tlhd{v}{Misinterpret.}
\tlhf{TKD}

\tlhw{yejquv}
\tlhd{n}{High Council.}
\tlhf{TKD}

\tlhw{yap}
\tlhd{v}{Be enough, sufficient.}
\tlhf{TKD}

\tlhw{yaS}
\tlhd{n}{Officer.}
\tlhf{TKD}

\tlhw{yav}
\tlhd{n}{Ground.}
\tlhf{TKD}

\tlhw{yay}
\tlhd{n}{The consonant "y".}
\tlhf{qep\'a\' wa\'maH javDIch 2009}

\tlhw{yay}
\tlhd{n}{Victory, triumph.}
\tlhf{TKD}

\tlhw{yay\'}
\tlhd{v}{Be shocked, dumbfounded.}
\tlhf{TKD}

\tlhw{yej}
\tlhd{n}{Assembly, council.}
\tlhf{TKD}

\tlhw{yem}
\tlhd{v}{Sin.}
\tlhf{TKD}

\tlhw{yep}
\tlhd{v}{Be careful.}
\tlhf{TKD}

\tlhw{yepHa\'}
\tlhd{v}{Be careless.}
\tlhf{TKD}

\tlhw{yev}
\tlhd{v}{Pause.}
\tlhf{TKD}

\tlhw{yIb}
\tlhd{n}{Vent.}
\tlhf{TKD}

\tlhw{yIH}
\tlhd{n}{Tribble.}
\tlhf{TKD}

\tlhw{yIn}
\tlhd{n}{Life.}
\tlhf{TKD}

\tlhw{yIn}
\tlhd{v}{Live.}
\tlhf{TKD}

\tlhw{yInroH}
\tlhd{n}{Life signs.}
\tlhf{TKD}

\tlhw{yIntagh}
\tlhd{n}{Life-support system.}
\tlhf{TKD}

\tlhw{yIQ}
\tlhd{v}{Be wet.}
\tlhf{TKD}

\tlhw{yIr}
\tlhd{n}{Gather.}
\tlhcf{boS}
\tlhco{This was used in a command for a warrior to gather his armor and weapons before a battle. Before seeing this word, I might have used \tlhe{boS} "collect" to refer to packing for a trip. Now, \tlhe{yIr} seems a better choice.}{ghunchu\'wI\' \'utlh}
\tlhf{qep\'a\' wa\'maH chorghDIch 2011}

\tlhw{yIv}
\tlhd{v}{Chew.}[Annoy, bother, irk, irritate. (slang)]
\tlhco{This verb literally means "chew". In its slang sense, it is found in such constructions as \tlhe{choyIv} ("You bother me"--literally, "You chew me") or \tlhe{muyIv romuluS Ha\'DIbaH} ("The Romulan dog irks me"; literally, "The Romulan animal chews me", where "Romulan animal" is a derogatory way to refer to a Romulan). The subject of \tlhe{yIv} can be only a person or creature, not an inanimate object or a situation. Thus, it is inappropriate to say something like \tlhe{muyIv \'Iw HIq bIr} ("cold bloodwine chews me"). Standard ways to express the same idea are \tlhe{nuQ} ("annoy, bother") and \tlhe{berghmoH} ("irritate").}{KGT}{167}
\tlhc
\tlhca{DeSvetlh yIv}{Chew that arm!}{PK}
\tlhe{to\'waQ yIyIv!}{Chew ligament! (idiom: "Take your time thinking about it!")}
\tlhe{to\'waQ vIyIvtaH}{I'm chewing ligament (i.e. "I'm in the process of taking my time considering the matter").}
\tlhe{to\'waQ yIyIvQo\'; DaH yIwuq!}{Don't chew ligament; decide now!}
\tlhe{HoD QaQbe\' ghaH; reH to\'waQ yIv}{He/she is not a good captain.}
\tlhcf{chop}[ghup][noS][roS][Sop][waH]
\tlhf{TKD}[KGT]

\tlhw{yIvbeH}
\tlhd{n}{Shirt.}
\tlhf{Eurotalk}

\tlhw{yIvbeH SeQHa\'}
\tlhd{n}{T-shirt.}
\tlhf{Eurotalk}

\tlhw{yISra\'el}
\tlhd{n}{Isreal.}
\tlhf{Eurotalk}

\tlhw{yIt}
\tlhd{v}{Walk.}
\tlhf{TKD}

\tlhw{yIvbeH}
\tlhd{n}{Tunic.}
\tlhf{TKD}

\tlhw{yob}
\tlhd{v}{Harvest.}
\tlhf{TKD}

\tlhw{yoD}
\tlhd{n}{Shield.}
\tlhf{TKD}

\tlhw{yoD}
\tlhd{v}{Shield.}
\tlhf{TKD}

\tlhw{yoH}
\tlhd{v}{Be brave.}
\tlhf{TKD}

\tlhw{yoj}
\tlhd{n}{Judgment.}
\tlhf{TKD}

\tlhw{yol}
\tlhd{n}{Conflict.}
\tlhf{TKD}

\tlhw{yon}
\tlhd{v}{Be satisfied.}
\tlhf{TKD}

\tlhw{yonmoH}
\tlhd{v}{Satisfy.}
\tlhf{TKD}

\tlhw{yong}
\tlhd{v}{Get in.}
\tlhf{TKD}

\tlhw{yopwaH}
\tlhd{n}{Pants.}
\tlhf{TKD}

\tlhw{yopwaH bId}
\tlhd{n}{Shorts.}
\tlhf{Eurotalk}

\tlhw{yoq}
\tlhd{n}{Humanoid.}
\tlhf{TKD}

\tlhw{yoS}
\tlhd{n}{District, area.}
\tlhf{TKD}

\tlhw{yot}
\tlhd{v}{Invade.}
\tlhf{TKD}

\tlhw{yot}
\tlhd{n}{Invasion.}
\tlhf{TKD}

\tlhw{yotlh}
\tlhd{n}{Field (of land).}
\tlhf{TKD}

\tlhw{yov}
\tlhd{v}{Charge (military term).}
\tlhf{TKD}

\tlhw{yoy}
\tlhd{v}{Be upside down.}
\tlhf{TKD}

\tlhw{yo\'}
\tlhd{n}{Fleet (of ships).}
\tlhf{TKD}

\tlhw{yuch}
\tlhd{n}{Chocolate.}
\tlhf{TKD}

\tlhw{yuD}
\tlhd{v}{Be dishonest.}
\tlhf{TKD}

\tlhw{yuDHa\'}
\tlhd{v}{Be honest.}
\tlhf{TKD}

\tlhw{yupma\'}
\tlhd{n}{Festival.}
\tlhf{TKD}

\tlhw{yuQ}
\tlhd{n}{Planet.}
\tlhf{TKD}

\tlhw{yuQHom}
\tlhd{n}{Planetoid.}
\tlhf{TKD}

\tlhw{yuQjIjQa\'}
\tlhd{n}{United Federation of Planets.}
\tlhf{TKD}

\tlhw{yuv}
\tlhd{v}{Push.}
\tlhf{TKD}

\tlhw{yu\'}
\tlhd{v}{Question, interrogate.}
\tlhf{TKD}

\tlhw{yu\'egh}
\tlhd{n}{Wave.}