\chapter{\tlhe{\'}}

\tlhw{\'a}
\tlhd{conj}{But, nevertheless, even so, however.}
\tlhf{TKD}

\tlhw{\'ach}
\tlhd{conj}{But, nevertheless, even so, however.}
\tlhf{TKD}

\tlhw{\'aD}
\tlhd{n}{Vein.}
\tlhf{TKD}

\tlhw{\'aH}
\tlhd{n}{Paraphernalia.}
\tlhf{TKD}

\tlhw{\'aj}
\tlhd{n}{Admiral.}
\tlhf{TKD}

\tlhw{\'anyan \'oQqar}
\tlhd{n}{Onion.}
\tlhf{Eurotalk}

\tlhw{\'ang}
\tlhd{v}{Show, reveal.}
\tlhf{TKD}

\tlhw{\'amerI\'qa\' SepjIjQa\'}
\tlhd{n}{United States of America.}
\tlhf{Eurotalk}

\tlhw{\'ar}
\tlhd{ques}{How many? how much?}
\tlhf{TKD}

\tlhw{\'argh}
\tlhd{v}{Worsen.}
\tlhf{TKD}

\tlhw{\'aSralya\'}
\tlhd{n}{Australia.}
\tlhf{Eurotalk}

\tlhw{\'at}
\tlhd{n}{The vowel "a".}
\tlhf{qep\'a\' wa\'maH javDIch 2009}

\tlhw{\'av}
\tlhd{v}{Guard.}
\tlhf{TKD}

\tlhw{\'avwI\'}
\tlhd{n}{Guard.}
\tlhco{\tlhe{\'avwI\'pu\'} may be the Klingon version of police and border patrol - they interrogate the Terran tourist at the spaceport and they patrol the streets in PK. A slang term for \tlhe{\'avwI} (and perhaps other bothersome officials) appears to be \tlhe{ghew} (lit. "bug, insect"). In ST6 Kesla tells Uhura, whom he thinks is a smuggler, \tlhe{ghewmey SuqQo\'!} "Don\'t catch any bugs!" In the novelization this is explained as smuggler\'s code for "Good luck in avoiding border officials!" We also have this joke in PK: \tlhe{Qo\'noSDaq paw cha\' DIvI\' beq. tlhIngan \'avwI\' lughom. lutlhob <<naDevvo\' vaS\'a\'Daq majaHlaH\'a\'?>> jang \'avwI\' <<lIchopbe\'chugh ghewmey!>>} Two Federation crewmen arrive on Kronos. They meet a Klingon guard. They ask him, "Can we get to the Great Hall from here?" The guard answers, "If the bugs do not bite you!"}{PK}
\tlhco{If this is a reference to getting picked up by the authorities, the phrase \tlhe{ghIlab ghewmey tIbuSQo\'} "Pay no heed to glob flies" (i.e. "Don\'t pay attention to insignificant things" [TKW]) may mean something different in the Klingon underworld."}{AKD}
\tlhc
\tlhe{mughoS \'avwI\' \'ar}{How many guards are' coming?}
\tlhca{\'avwI\' nejDI\' narghta\'bogh qama\' reH \'avwI\' Sambej}{When an escaped prisoner looks for a guard, he always finds one.}[TKW]
\tlhca{pawpu\' \'avwI\' paghDIch}{The zeroth guard has arrived. [...implying that no guard has arrived even though one (and probably more than one) is expected.]}[KGT 177]
\tlhca{\'avwI\' vaghDIch}{Fifth guard.}[KGT]
\tlhca{\'avwI\' Hay\' yaS}{The officer duels the guard.}[KGT]
\tlhca{jav luHpu\' \'avwI\'}{The guard has caused a prisoner to confess. [The guard has yanked a six].}[KGT]
\tlhca{Qo\'noSDaq paw cha\' DIvI\' beq. tlhIngan \'avwI\' lughom. lutlhob <<naDevvo\' vaS\'a\'Daq majaHlaH\'a\'?>> jang \'avwI\' <<lIchopbe\'chugh ghewmey!>>}{Two Federation crewmen arrive on Kronos. They meet a Klingon guard. They ask him, "Can we get to the Great Hall from here?" The guard answers, "If the bugs do not bite you!"}[PK]
\tlhf{TKD}[TKW][KGT][CK][PK][ST6]

\tlhw{\'aw\'}
\tlhd{v}{Sting.}
\tlhf{TKD}

\tlhw{\'ay\'}
\tlhd{n}{Section.}
\tlhf{TKD}

\tlhw{\'eb}
\tlhd{n}{Opportunity.}
\tlhf{TKD}

\tlhw{\'ech}
\tlhd{n}{Brigadier.}
\tlhf{TKD}

\tlhw{\'eH}
\tlhd{excl}{Ready.}
\tlhf{TKD}

\tlhw{\'ej}
\tlhd{conj}{And (joining sentences).}
\tlhf{TKD}

\tlhw{\'ejDo\'}
\tlhd{n}{Starship, starship class.}
\tlhf{TKD}

\tlhw{\'ejyo\'}
\tlhd{n}{Starfleet.}
\tlhf{TKD}

\tlhw{\'ejyo\'waw\'}
\tlhd{n}{Star base.}
\tlhf{TKD}

\tlhw{\'el}
\tlhd{v}{Enter, go in.}
\tlhf{TKD}

\tlhw{\'elaDya\'}
\tlhd{n}{Greece.}
\tlhf{Eurotalk}

\tlhw{\'elaS}
\tlhd{n}{Elas.}
\tlhf{TKD}

\tlhw{\'elI\'jaH}
\tlhd{n}{Unexpected visitor.}
\tlhco{Mark Okrand wrote the word in Angieszka Solska's notebook during \tlhe{qep\'a\'} 2006, but she expected it to be published \tlhe{HolQeD}, and did not want to spoil the surprise, therefore the word was not made public until a year later, in a discussion about new words from the next \tlhe{qep\'a\'} on the tlhingan-hol mailing list.}{Zrajm C Akfohg}
\tlhf{qep\'a\' wa\'maH wejDIch 2006}

\tlhw{\'eng}
\tlhd{n}{Cloud.}
\tlhf{TKD}

\tlhw{\'epIl naHmey}
\tlhd{pl n}{Apples.}
\tlhf{Eurotalk}

\tlhw{\'er}
\tlhd{n}{Type of animal.}
\tlhf{TKD}

\tlhw{\'er\'In}
\tlhd{n}{End (of stick, rope, etc.), other end from \tlhe{megh\'an}.}
\tlhcf{megh'an}
\tlhco{This can refer to either end of the rope, stick or whatever, but once you arbitrarily choose an end to call \tlhe{megh\'an}, the OTHER end is called \tlhe{\'er\'In}. Before you make this initial reference, either end can be called either \tlhe{\'er\'In} or \tlhe{megh\'an}.}{HolQeD}
\tlhf{HolQeD v12n2p6}

\tlhw{\'eSpanya\'}
\tlhd{n}{Spain.}
\tlhf{Eurotalk}

\tlhw{\'et}
\tlhd{n}{Fore.}[The vowel "e".]
\tlhf{TKD}[qep\'a\' wa\'maH javDIch 2009]

\tlhw{\'etlh}
\tlhd{n}{Sword.}
\tlhf{TKD}

\tlhw{\'evnagh}
\tlhd{n}{Subspace.}
\tlhf{TKD}

\tlhw{\'e\'}
\tlhd{pro}{That (previous topic).}
\tlhf{TKD}

\tlhw{\'IH}
\tlhd{v}{Be beautiful, handsome.}
\tlhf{TKD}

\tlhw{\'Ij}
\tlhd{v}{Listen.}
\tlhf{TKD}

\tlhw{\'Il}
\tlhd{v}{Be sincere.}
\tlhf{TKD}

\tlhw{\'Ip}
\tlhd{n}{Oath.}
\tlhf{TKD}

\tlhw{\'Ip}
\tlhd{v}{Vow, swear.}
\tlhf{TKD}

\tlhw{\'IQ}
\tlhd{v}{Be sad.}
\tlhf{TKD}

\tlhw{\'ISjaH}
\tlhd{n}{Calendar.}
\tlhf{TKD}

\tlhw{\'It}
\tlhd{n}{The vowel "I".}
\tlhd{v}{Be depressed.}
\tlhf{TKD}[qep\'a\' wa\'maH javDIch 2009]

\tlhw{\'Italya\'}
\tlhd{n}{Italy.}
\tlhf{Eurotalk}

\tlhw{\'Itlh}
\tlhd{v}{Be advanced, highly developed.}
\tlhf{TKD}

\tlhw{\'Iv}
\tlhd{n}{Altitude.}
\tlhf{TKD}

\tlhw{\'Iv}
\tlhd{ques}{Who?}
\tlhf{TKD}

\tlhw{\'Iw}
\tlhd{n}{Blood.}
\tlhf{TKD}

\tlhw{\'o}
\tlhd{excl}{O.}
\tlhco{{'o} appears before a name used as direct address. As we saw it used, it acted a little like a vocative prefix: \tlhe{\'o qeylIS, qeylIS, qeylIS...} It seems to be an honorific exclamation, or maybe it\'s just used for direct address in general and doesn\'t actually have the implication of deference or reverence that I inferred. If it hadn\'t been consistently present when the opera "speaks to" Kahless, and consistently absent when it "speaks to" Molor, I might have ignored it as an unimportant background sound. English has almost exactly the same sound with pretty much the same meaning: "O Mouse, do you know the way out of this pool? I am very tired of swimming about here, O Mouse!" (from Alice\'s Adventures in Wonderland)}{ghunchu\'wI\' \'utlh}
\tlhf{qep\'a\' wa\'maH chorghDIch 2011}

\tlhw{\'och}
\tlhd{n}{Tunnel.}
\tlhf{TKD}

\tlhw{\'oD}
\tlhd{v}{Arbitrate, mediate.}
\tlhf{TKD}

\tlhw{\'ogh}
\tlhd{v}{Invent, devise.}
\tlhf{TKD}

\tlhw{\'oH}
\tlhd{pro}{It.}
\tlhf{TKD}

\tlhw{\'oj}
\tlhd{v}{Be thirsty.}
\tlhf{TKD}

\tlhw{\'ol}
\tlhd{v}{Verify.}
\tlhf{TKD}

\tlhw{\'ong}
\tlhd{v}{Be cunning, sly.}
\tlhf{TKD}

\tlhw{\'orghen}
\tlhd{n}{Organia.}
\tlhf{TKD}

\tlhw{\'orghen rojmab}
\tlhd{n}{Organian Peace Treaty.}
\tlhf{TKD}

\tlhw{\'orghengan}
\tlhd{n}{Organian.}
\tlhf{TKD}

\tlhw{\'oS}
\tlhd{v}{Represent.}
\tlhf{TKD}

\tlhw{\'oSwI\'}
\tlhd{n}{Emissary.}
\tlhf{TKD}

\tlhw{\'ot}
\tlhd{n}{The vowel "o".}
\tlhf{qep\'a\' wa\'maH javDIch 2009}

\tlhw{\'otlh}
\tlhd{n}{Photon.}
\tlhf{TKD}

\tlhw{\'ov}
\tlhd{v}{Compete.}
\tlhf{TKD}

\tlhw{\'oy\'}
\tlhd{n}{Ache, pain, sore.}
\tlhd{v}{Ache, hurt, be sore.}
\tlhf{TKD}

\tlhw{\'oy\'naQ}
\tlhd{n}{Painstick.}
\tlhf{TKD}

\tlhw{\'o\'}
\tlhd{n}{Aft.}
\tlhf{TKD}

\tlhw{\'ugh}
\tlhd{v}{Be heavy.}
\tlhf{TKD}

\tlhw{\'uH}
\tlhd{v}{Have a hangover, be hung over.}
\tlhf{TKD}

\tlhw{\'ul Sech}
\tlhd{n}{Flashlight.}
\tlhf{Eurotalk}

\tlhw{\'ul \'aplo\'mey}
\tlhd{pl n}{Batteries.}
\tlhf{Eurotalk}

\tlhw{\'um}
\tlhd{v}{Be qualified.}
\tlhf{TKD}

\tlhw{\'uQ}
\tlhd{n}{Dinner.}
\tlhf{TKD}

\tlhw{\'urmang}
\tlhd{n}{Treason.}
\tlhf{TKD}

\tlhw{\'uS}
\tlhd{n}{Leg.}
\tlhf{TKD}

\tlhw{\'uSgheb}
\tlhd{n}{A noisy bird known to make a ruckus at dawn, likened to a rooster [much fiercer than a rooster.]}
\tlhf{HolQeD v10n4p5.}

\tlhw{\'ut}
\tlhd{n}{Necessary.}[The vowel "u".]
\tlhd{v}{Be essential.}
\tlhf{TKD}[qep\'a\' wa\'maH javDIch 2009]

\tlhw{\'utlh}
\tlhd{n}{Officer.}
\tlhf{TKD}

\tlhw{\'uy}
\tlhd{v}{Press down.}
\tlhf{TKD}

\tlhw{\'uy\'}
\tlhd{num}{Million.}
\tlhf{TKD}

\tlhw{\'u\'}
\tlhd{n}{Universe.}
\tlhf{TKD}